---
title: "00 — Master country reference"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Builds the `countries` reference table that every downstream source joins
through. One row per country with all the ID systems we need:

- `iso3c` — ISO 3166-1 alpha-3 (matches WDI, UNHCR, V-Dem `country_text_id`)
- `iso3n` — ISO numeric (for ACLED/GDELT when those arrive)
- `vdem_id` — V-Dem's internal country id
- `gw_code` — Gleditsch-Ward code (for Pilster-Böhmelt, de Bruin)
- `name` — canonical English name
- `region`, `subregion` — UN region groupings (for dashboard reference class)

Source: V-Dem's country roster (broadest coverage) plus the `countrycode`
package for region/GW crosswalks.

## Install

```{r}
#| label: install
#| eval: false
install.packages("countrycode")
```

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(countrycode)
```

## Build the reference table

V-Dem already carries `country_text_id` (ISO3) for every country it covers.
We start from its distinct country list, then use `countrycode` to add
ISO numeric, GW code, and UN regions.

```{r}
#| label: build
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

vdem_countries <- dbGetQuery(con, "
  SELECT DISTINCT country_id, country_text_id, country_name
  FROM vdem
  WHERE country_text_id IS NOT NULL AND year >= 1990
  ORDER BY country_name
")

countries <- data.frame(
  iso3c       = vdem_countries$country_text_id,
  name        = vdem_countries$country_name,
  vdem_id     = vdem_countries$country_id,
  iso3n       = countrycode(vdem_countries$country_text_id, "iso3c", "iso3n",
                            warn = FALSE),
  gw_code     = countrycode(vdem_countries$country_text_id, "iso3c", "gwn",
                            warn = FALSE),
  region      = countrycode(vdem_countries$country_text_id, "iso3c", "region",
                            warn = FALSE),
  subregion   = countrycode(vdem_countries$country_text_id, "iso3c", "un.regionsub.name",
                            warn = FALSE),
  stringsAsFactors = FALSE
)

cat("Total countries:", nrow(countries), "\n")
cat("With ISO numeric:", sum(!is.na(countries$iso3n)), "\n")
cat("With GW code:    ", sum(!is.na(countries$gw_code)), "\n")
cat("With region:     ", sum(!is.na(countries$region)), "\n")
head(countries, 5)
```

## Patch known gaps

`countrycode` doesn't have entries for some V-Dem countries (historical states
like Czechoslovakia / Yugoslavia, contested cases like Kosovo / Palestine /
Taiwan). For sovereign-but-contested cases we want in the panel, patch them.

```{r}
#| label: patches
patches <- list(
  XKX = list(name = "Kosovo",            region = "Europe",  subregion = "Southern Europe"),
  PSE = list(name = "Palestine",         region = "Asia",    subregion = "Western Asia"),
  TWN = list(name = "Taiwan",            region = "Asia",    subregion = "Eastern Asia")
)

for (iso in names(patches)) {
  idx <- which(countries$iso3c == iso)
  if (length(idx) > 0) {
    for (field in names(patches[[iso]])) {
      if (is.na(countries[[field]][idx])) countries[[field]][idx] <- patches[[iso]][[field]]
    }
  }
}

cat("After patches — with region:", sum(!is.na(countries$region)), "of", nrow(countries), "\n")
```

## Write to DuckDB

```{r}
#| label: write
dbWriteTable(con, "countries", countries, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS countries FROM countries")
```

## Region distribution

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT region, COUNT(*) AS n
  FROM countries
  WHERE region IS NOT NULL
  GROUP BY region
  ORDER BY n DESC
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
