---
title: "00 — Master country reference"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** Every other notebook in this project joins data on
"the country." But different datasets identify countries differently
(some use 3-letter ISO codes like `USA`, some use numeric codes like
840, some use names like "United States of America" or "United States").
Before we can join anything, we need ONE table that maps all these
identifier systems to each other. That's what this notebook builds.

**Data-analytics concepts shown here.**

- *Master / reference tables*: one row per entity, all identifiers,
  built first so downstream joins are clean.
- *Identifier reconciliation*: real-world data uses competing ID
  systems for the same things (ISO-3, COW, GW, V-Dem internal IDs).
  Crosswalking them is unglamorous but essential.
- *Honest handling of edge cases*: contested-sovereignty entities
  (Kosovo, Palestine, Taiwan) aren't in standard ID systems. We patch
  them in explicitly with a documented `patches` list rather than
  silently dropping them.

**Domain terms used here.**

- **ISO 3166-1 alpha-3** — three-letter country code (USA, CUB, FRA).
- **COW / Correlates of War code** — numeric code used by many
  political-science datasets.
- **Gleditsch-Ward (GW) code** — alternative numeric system used by
  Pilster-Böhmelt and others.
- **V-Dem country_id** — V-Dem's internal numbering.
- **`countrycode`** — the R package that does the crosswalking; the
  workhorse tool for this kind of task.

**What success looks like.** The output is one DuckDB table named
`countries` with ~200 rows. Every other loader joins on `iso3c`,
which means this table determines the universe of countries the model
can predict for. If your country of interest is missing from this
table, fix it here.
:::

Builds the `countries` reference table that every downstream source joins
through. One row per country with all the ID systems we need:

- `iso3c` — ISO 3166-1 alpha-3 (matches WDI, UNHCR, V-Dem `country_text_id`)
- `iso3n` — ISO numeric (for ACLED/GDELT when those arrive)
- `vdem_id` — V-Dem's internal country id
- `gw_code` — Gleditsch-Ward code (for Pilster-Böhmelt, de Bruin)
- `name` — canonical English name
- `region`, `subregion` — UN region groupings (for dashboard reference class)

Source: V-Dem's country roster (broadest coverage) plus the `countrycode`
package for region/GW crosswalks.

## Install

```{r}
#| label: install
#| eval: false
install.packages("countrycode")
```

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(countrycode)
```

## Build the reference table

V-Dem already carries `country_text_id` (ISO3) for every country it covers.
We start from its distinct country list, then use `countrycode` to add
ISO numeric, GW code, and UN regions.

```{r}
#| label: build
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

vdem_countries <- dbGetQuery(con, "
  SELECT DISTINCT country_id, country_text_id, country_name
  FROM vdem
  WHERE country_text_id IS NOT NULL AND year >= 1990
  ORDER BY country_name
")

countries <- data.frame(
  iso3c       = vdem_countries$country_text_id,
  name        = vdem_countries$country_name,
  vdem_id     = vdem_countries$country_id,
  iso3n       = countrycode(vdem_countries$country_text_id, "iso3c", "iso3n",
                            warn = FALSE),
  gw_code     = countrycode(vdem_countries$country_text_id, "iso3c", "gwn",
                            warn = FALSE),
  region      = countrycode(vdem_countries$country_text_id, "iso3c", "region",
                            warn = FALSE),
  subregion   = countrycode(vdem_countries$country_text_id, "iso3c", "un.regionsub.name",
                            warn = FALSE),
  stringsAsFactors = FALSE
)

cat("Total countries:", nrow(countries), "\n")
cat("With ISO numeric:", sum(!is.na(countries$iso3n)), "\n")
cat("With GW code:    ", sum(!is.na(countries$gw_code)), "\n")
cat("With region:     ", sum(!is.na(countries$region)), "\n")
head(countries, 5)
```

## Patch known gaps

`countrycode` doesn't have entries for some V-Dem countries (historical states
like Czechoslovakia / Yugoslavia, contested cases like Kosovo / Palestine /
Taiwan). For sovereign-but-contested cases we want in the panel, patch them.

```{r}
#| label: patches
patches <- list(
  XKX = list(name = "Kosovo",            region = "Europe",  subregion = "Southern Europe"),
  PSE = list(name = "Palestine",         region = "Asia",    subregion = "Western Asia"),
  TWN = list(name = "Taiwan",            region = "Asia",    subregion = "Eastern Asia")
)

for (iso in names(patches)) {
  idx <- which(countries$iso3c == iso)
  if (length(idx) > 0) {
    for (field in names(patches[[iso]])) {
      if (is.na(countries[[field]][idx])) countries[[field]][idx] <- patches[[iso]][[field]]
    }
  }
}

cat("After patches — with region:", sum(!is.na(countries$region)), "of", nrow(countries), "\n")
```

## Write to DuckDB

```{r}
#| label: write
dbWriteTable(con, "countries", countries, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS countries FROM countries")
```

## Region distribution

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT region, COUNT(*) AS n
  FROM countries
  WHERE region IS NOT NULL
  GROUP BY region
  ORDER BY n DESC
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
