---
title: "01 — Load V-Dem"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Loads V-Dem v16 (~28K country-years × 4,618 variables) into the `vdem` table
of `data/panel.duckdb`. Lands wide; downstream notebooks select variables.

## Install

V-Dem ships as the GitHub R package `vdemdata`. `remotes` is a lighter
alternative to `devtools` that handles pure-R sources without needing Rtools.

```{r}
#| label: install
#| eval: false
install.packages(c("DBI", "duckdb", "ggplot2", "remotes"))
remotes::install_github("vdeminstitute/vdemdata")
```

If the GitHub install fails, download
[V-Dem CY Full+Others CSV from v-dem.net](https://v-dem.net/data/the-v-dem-dataset/),
save it at `data/raw/V-Dem-CY-Full+Others-v16.csv`, and swap the `load` chunk
for a DuckDB `read_csv_auto()` against that path.

## Libraries

```{r}
#| label: setup
library(vdemdata)
library(DBI)
library(duckdb)
library(ggplot2)
source("_helpers.R")
```

## Write to DuckDB

```{r}
#| label: load
dir.create("../data", showWarnings = FALSE, recursive = TRUE)
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

dbWriteTable(con, "vdem", vdem, overwrite = TRUE)
srs_table(
  dbGetQuery(con, "SELECT COUNT(*) AS rows FROM vdem"),
  big_mark = TRUE
)
```

## Coverage

```{r}
#| label: inspect
srs_table(dbGetQuery(con, "
  SELECT
    COUNT(DISTINCT country_name) AS countries,
    MIN(year) AS earliest_year,
    MAX(year) AS latest_year
  FROM vdem
"))
```

## Cuba sanity check

Four representative indices we'll likely use downstream:

- `v2x_regime` — Regimes of the World (0 = closed autocracy, 3 = liberal democracy)
- `v2regsuploc` — geographic locus of regime support (ordinal 0–4)
- `v2x_freexp_altinf` — freedom of expression + alternative information (0–1)
- `v2x_civlib` — civil liberties (0–1)

```{r}
#| label: cuba-recent
cuba_recent <- dbGetQuery(con, "
  SELECT year,
         v2x_regime         AS regime_type,
         v2regsuploc        AS regime_support_locus,
         v2x_freexp_altinf  AS freedom_expression,
         v2x_civlib         AS civil_liberties
  FROM vdem
  WHERE country_name = 'Cuba' AND year >= 2015
  ORDER BY year
")
srs_table(
  cuba_recent,
  digits = 3,
  caption = "Cuba — selected V-Dem indices, 2015 onward"
)
```

```{r}
#| label: cuba-plot
#| fig-cap: "Cuba's civil-liberties index, 2015 onward (V-Dem v16)"
#| fig-width: 7
#| fig-height: 4
ggplot(cuba_recent, aes(x = year, y = civil_liberties)) +
  geom_line(color = srs_palette$accent, linewidth = 0.9) +
  geom_point(color = srs_palette$accent, size = 2.5) +
  scale_x_continuous(breaks = cuba_recent$year) +
  labs(x = NULL, y = "v2x_civlib (0–1)",
       caption = "Source: V-Dem v16 via vdemdata R package") +
  srs_theme()
```

## Cleanup

DuckDB locks the file while a connection is open; disconnect cleanly so
subsequent notebooks can write.

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
