---
title: "02 — Load World Bank WDI"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** Economic conditions are one of the strongest
predictors of political unrest (inflation shocks, growth collapses,
high unemployment). The World Bank publishes ~1,500 indicators in its
World Development Indicators (WDI) database with global coverage 1960
to present. We pull six: inflation, GDP growth, GDP per capita, trade
share, unemployment, and population.

**Data-analytics concepts shown here.**

- *REST API ingestion via an R wrapper*. The `WDI` package wraps the
  World Bank API so we can pull data with one function call rather
  than constructing HTTP requests manually.
- *Living data vs. frozen data*. WDI revises retroactively — the
  number we got today for 2023 may differ next quarter. The render
  date is therefore part of the data; pipeline reproducibility requires
  preserving the snapshot.
- *Naming things consistently*. We rename the cryptic WDI indicator
  codes (e.g., `FP.CPI.TOTL.ZG`) to human-readable column names
  (`cpi_inflation`) at load time. Downstream code never sees the
  cryptic codes.

**Domain terms used here.**

- **WDI** — World Development Indicators (World Bank). The canonical
  source of country-level macro stats.
- **CPI** — Consumer Price Index, the standard measure of inflation.
- **GDP per capita (constant USD)** — wealth level, adjusted for
  inflation. Constant-USD rather than current-USD so it's comparable
  across years.
- **Trade as % of GDP** — exports + imports divided by GDP. A measure
  of economic openness.

**What success looks like.** The `wdi` table has ~270 reporting
entities × ~35 years × 6 indicators. Coverage is patchy for some
indicators (Cuba doesn't report CPI to the World Bank, for example),
which is OK — we handle the NAs honestly in `08_build_panel.qmd`.
:::

Pulls five macroeconomic indicators (inflation, GDP growth, GDP/capita, trade
exposure, unemployment) from the World Bank's REST API and lands them as the
`wdi` table in `data/panel.duckdb`. One row per country × year. WDI hits the
network on every render; values for the most recent 2–3 years may shift as
the World Bank back-revises.

## Install

```{r}
#| label: install
#| eval: false
install.packages("WDI")
```

## Libraries

```{r}
#| label: setup
library(WDI)
library(DBI)
library(duckdb)
library(ggplot2)
source("_helpers.R")
```

## Indicators

Names on the left become the column names in DuckDB.

```{r}
#| label: indicators
indicators <- c(
  cpi_inflation = "FP.CPI.TOTL.ZG",    # CPI inflation, annual %
  gdp_growth    = "NY.GDP.MKTP.KD.ZG", # GDP growth, annual %
  gdp_per_cap   = "NY.GDP.PCAP.KD",    # GDP per capita, constant 2015 US$
  trade_pct_gdp = "NE.TRD.GNFS.ZS",    # Trade (% of GDP)
  unemployment  = "SL.UEM.TOTL.ZS",    # Unemployment, % of labor force
  population    = "SP.POP.TOTL"        # Population, total (used by 03a for per-capita protest threshold)
)
```

## Fetch and write

One row per country-year with five named columns. ~270 reporting entities
(countries + a few regional aggregates) × ~35 years.

```{r}
#| label: fetch-write
wdi <- WDI(country = "all", indicator = indicators,
           start = 1990, end = format(Sys.Date(), "%Y"))
nrow(wdi); head(wdi, 3)

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "wdi", wdi, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows FROM wdi")
```

## Coverage

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT
    COUNT(DISTINCT country) AS countries,
    MIN(year) AS earliest_year,
    MAX(year) AS latest_year,
    SUM(CASE WHEN cpi_inflation IS NOT NULL THEN 1 ELSE 0 END) AS cpi_obs,
    SUM(CASE WHEN gdp_growth    IS NOT NULL THEN 1 ELSE 0 END) AS gdpg_obs
  FROM wdi
")
```

## Cuba sanity check

Cuba doesn't report CPI to the World Bank — `cpi_inflation` will be `NA` for
every Cuba row. The other four have full coverage. The 2020 GDP collapse
(−10.95%) is the COVID-19 economic shock.

```{r}
#| label: cuba-recent
cuba_recent <- dbGetQuery(con, "
  SELECT year, cpi_inflation, gdp_growth, gdp_per_cap, trade_pct_gdp, unemployment
  FROM wdi
  WHERE iso3c = 'CUB' AND year >= 2015
  ORDER BY year
")
cuba_recent
```

```{r}
#| label: cuba-plot
#| fig-cap: "Cuba's GDP growth, 2015 onward (World Bank WDI)"
#| fig-width: 7
#| fig-height: 4
ggplot(cuba_recent, aes(x = year, y = gdp_growth)) +
  geom_hline(yintercept = 0, color = "#A89880", linewidth = 0.4) +
  geom_line(color = srs_palette$accent, linewidth = 0.9) +
  geom_point(aes(color = gdp_growth < -5), size = 2.5) +
  scale_color_manual(values = c(`FALSE` = srs_palette$accent, `TRUE` = srs_palette$crisis), guide = "none") +
  scale_x_continuous(breaks = cuba_recent$year) +
  labs(x = NULL, y = "GDP growth (annual %)",
       caption = "Source: World Bank WDI via WDI R package") +
  srs_theme()
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
