---
title: "02b — Load IMF WEO supplements + Hanke estimates"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** The World Bank doesn't have inflation data for the
countries that matter most for political-risk forecasting. Cuba has
no WDI inflation at all. Venezuela's official numbers are politically
managed (the government denies hyperinflation). North Korea, Eritrea,
and parts of Syria have nothing. This notebook fills those gaps with
**three carefully prioritized backfill sources**.

**Data-analytics concepts shown here.**

- *Source priority chains* — when a value is missing from your primary
  source, fall back to source #2, then source #3. SQL's `COALESCE`
  function expresses this pattern cleanly.
- *Hand-curated overrides for tricky cases* — sometimes you have to
  manually maintain a small CSV of "the right values" because no
  programmatic source has them. The `data/curated/` directory is the
  pattern for this: small, hand-edited, version-controlled, justified.
- *Data quality flags* — when a value comes from an imputation or
  curation step rather than the primary source, flag it so downstream
  code can treat it differently (e.g., listwise-drop imputed rows from
  training).
- *Bias-aware modeling* — politically-managed economic data isn't
  random noise; it's biased in known directions (governments suppress
  inflation). We use independent academic estimates (Hanke) and IMF
  Staff numbers from Article IV consultations to cross-check.

**Domain terms used here.**

- **IMF WEO** — International Monetary Fund World Economic Outlook.
  Twice-yearly forecasts with country-team estimates.
- **IMF Article IV** — annual IMF country consultations; produce
  independent assessments of national accounts.
- **Hanke Troubled Currencies Project** — Steve Hanke's
  hyperinflation-tracking estimates (Cato/Johns Hopkins). Independent
  measurements for politically-managed economies.
- **PCPIPCH** — IMF's series code for "Consumer Prices, percent
  change" (year-over-year inflation).

**What success looks like.** Three DuckDB tables land here:
`imf_inflation` (hand-curated values for ~7 crisis countries),
`imf_weo` (bulk IMF WEO data for ~200 countries), and `hanke_inflation`
(Hanke's independent estimates). The CPI priority chain in
`08_build_panel.qmd` then becomes: `COALESCE(wdi, imf_curated, imf_weo)`.
:::

The World Bank WDI has inflation gaps for the countries that matter most for
political-risk forecasting — Cuba (all years NA), North Korea, parts of Syria
and Eritrea — and politically-managed series for others (Venezuela, Argentina,
Zimbabwe, Lebanon). This notebook lands three supplement tables:

- **`imf_inflation`** — hand-curated IMF Staff estimates (Article IV
  consultations + WEO projections). High-trust crisis-country numbers,
  reviewed line-by-line. Covers Cuba (where WEO has nothing) and the
  politically-managed series for Venezuela/Argentina/Zimbabwe/Lebanon.
- **`imf_weo`** — bulk IMF World Economic Outlook database (PCPIPCH series,
  "consumer prices, percent change"). Parsed from the wide-form CSV the
  IMF Data portal exports. Provides continuous 1980–present coverage for
  ~200 countries — the broad backfill behind the hand-curated values.
- **`hanke_inflation`** — Steve Hanke (Johns Hopkins) Troubled Currencies
  Project estimates, derived from free-market exchange-rate decay. These
  are systematically 2-10x higher than IMF/national figures because they
  capture unofficial/black-market reality. Used by CNN and others when
  reporting "hyperinflation."

The panel build (08) uses a **WDI → imf_inflation (curated) → imf_weo
(bulk) → region-median** COALESCE chain. Hand-curated wins over the bulk
file so reviewed crisis values aren't overwritten by WEO defaults. The
Hanke table is loaded separately and exposed to the dashboard as a toggle.

`data/curated/*.csv` is version-controlled (hand-maintained).
`data/raw/imf_weo_*.csv` is gitignored (large, re-downloadable) — pull
from https://data.imf.org/en/datasets/IMF.RES:WEO when refreshing.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Load IMF inflation supplement

```{r}
#| label: imf
imf <- read.csv("../data/curated/imf_inflation_supplement.csv",
                comment.char = "#", stringsAsFactors = FALSE)
imf <- imf[!is.na(imf$cpi_inflation), c("iso3c", "year", "cpi_inflation", "source")]
imf$year <- as.integer(imf$year)
cat("IMF supplement rows:", nrow(imf), "\n")
cat("Countries covered:", length(unique(imf$iso3c)), "\n")
print(table(imf$iso3c))
```

## Load curated unemployment supplement

Hand-curated unemployment + underemployment estimates for state-managed
economies (Cuba, Belarus, Venezuela, PRK, Eritrea, Turkmenistan, Cambodia,
Laos) where the official ILO/WDI rate doesn't reflect economic reality
because of state-employment-guarantee policies. Sourced from IMF Article
IV reviews + economic-NGO consensus estimates.

```{r}
#| label: unemployment-supplement
un <- read.csv("../data/curated/unemployment_supplement.csv",
               comment.char = "#", stringsAsFactors = FALSE)
un <- un[!is.na(un$unemployment_curated),
         c("iso3c", "year", "unemployment_curated", "source")]
un$year <- as.integer(un$year)
cat("Unemployment supplement rows:", nrow(un), "\n")
cat("Countries covered:", length(unique(un$iso3c)), "\n")
print(table(un$iso3c))
```

## Load IMF WEO bulk file

The WEO CSV is wide: ~65 metadata columns followed by year columns 1980–2031.
Each row is one (country, indicator) pair. We extract three indicators:

- **PCPIPCH** — Consumer prices, % change (inflation). Backfills WDI inflation.
- **NGDP_RPCH** — Real GDP, % change (growth). Backfills WDI gdp_growth for 2025+.
- **NGDPDPC** — Nominal GDP per capita, USD. Backfills WDI gdp_per_cap for 2025+.

The latter two unblock 2025 forecasting: WDI typically releases full-year
2025 GDP data ~12–18 months behind real time, but IMF country teams produce
year-T estimates by April of year T+1 (Article IV consultations + the WEO
Spring update), which the IMF then projects forward through year T+6.

```{r}
#| label: weo-load
weo_path <- "../data/raw/imf_weo_2026_04.csv"

if (!file.exists(weo_path)) {
  stop("Missing ", weo_path,
       ". Download the bulk CSV from https://data.imf.org/en/datasets/IMF.RES:WEO ",
       "and save to that path. The file is gitignored (~26MB).")
}

weo_raw <- read.csv(weo_path, check.names = FALSE, stringsAsFactors = FALSE,
                    fileEncoding = "UTF-8-BOM", na.strings = c("", "NA"))

# Extract ISO3 (first dotted segment) and indicator code (second segment).
weo_raw$iso3c     <- sub("\\..*", "", weo_raw$SERIES_CODE)
weo_raw$indicator <- sub("^[A-Z]{3}\\.([A-Z_]+)\\..*$", "\\1", weo_raw$SERIES_CODE)

# Drop IMF aggregates (codes ending in Q — CAQ/CBQ/etc).
weo_raw <- weo_raw[!grepl("Q$", weo_raw$iso3c), ]

year_cols <- grep("^(19|20)[0-9]{2}$", names(weo_raw), value = TRUE)

# Pull-and-pivot helper for one indicator.
weo_extract <- function(indicator_code, value_col) {
  rows <- weo_raw[weo_raw$indicator == indicator_code, ]
  cat(indicator_code, " series rows:", nrow(rows), "\n")
  if (nrow(rows) < 100) {
    stop(sprintf("WEO %s filter matched only %d rows — expected ~200 (one per country). SERIES_CODE format may have changed.",
                 indicator_code, nrow(rows)))
  }
  long <- do.call(rbind, lapply(year_cols, function(y) {
    df <- data.frame(
      iso3c = rows$iso3c,
      year  = as.integer(y),
      stringsAsFactors = FALSE
    )
    df[[value_col]] <- suppressWarnings(as.numeric(rows[[y]]))
    df
  }))
  long <- long[!is.na(long[[value_col]]), ]
  long$source <- "IMF WEO April 2026 (data.imf.org bulk export)"
  long
}

weo_long          <- weo_extract("PCPIPCH",   "cpi_inflation")
weo_gdp_growth    <- weo_extract("NGDP_RPCH", "gdp_growth")
weo_gdp_per_cap   <- weo_extract("NGDPDPC",   "gdp_per_cap")

cat("\nWEO PCPIPCH:  rows =", nrow(weo_long),
    "  countries =", length(unique(weo_long$iso3c)),
    "  year span =", min(weo_long$year), "-", max(weo_long$year), "\n")
cat("WEO NGDP_RPCH: rows =", nrow(weo_gdp_growth),
    "  countries =", length(unique(weo_gdp_growth$iso3c)),
    "  year span =", min(weo_gdp_growth$year), "-", max(weo_gdp_growth$year), "\n")
cat("WEO NGDPDPC:   rows =", nrow(weo_gdp_per_cap),
    "  countries =", length(unique(weo_gdp_per_cap$iso3c)),
    "  year span =", min(weo_gdp_per_cap$year), "-", max(weo_gdp_per_cap$year), "\n")
```

## Load Hanke inflation estimates

```{r}
#| label: hanke
hanke <- read.csv("../data/curated/hanke_inflation_estimates.csv",
                  comment.char = "#", stringsAsFactors = FALSE)
hanke <- hanke[!is.na(hanke$cpi_inflation_hanke),
               c("iso3c", "year", "cpi_inflation_hanke", "source")]
hanke$year <- as.integer(hanke$year)
cat("Hanke estimates rows:", nrow(hanke), "\n")
cat("Countries covered:", length(unique(hanke$iso3c)), "\n")
print(table(hanke$iso3c))
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "imf_inflation",        imf,              overwrite = TRUE)
dbWriteTable(con, "imf_weo",              weo_long,         overwrite = TRUE)
dbWriteTable(con, "imf_weo_gdp_growth",   weo_gdp_growth,   overwrite = TRUE)
dbWriteTable(con, "imf_weo_gdp_pc",       weo_gdp_per_cap,  overwrite = TRUE)
dbWriteTable(con, "hanke_inflation",      hanke,            overwrite = TRUE)
dbWriteTable(con, "unemployment_curated", un,               overwrite = TRUE)
dbGetQuery(con, "
  SELECT 'imf_curated'         AS source, COUNT(*) AS rows FROM imf_inflation
  UNION ALL SELECT 'imf_weo',             COUNT(*) FROM imf_weo
  UNION ALL SELECT 'imf_weo_gdp_growth',  COUNT(*) FROM imf_weo_gdp_growth
  UNION ALL SELECT 'imf_weo_gdp_pc',      COUNT(*) FROM imf_weo_gdp_pc
  UNION ALL SELECT 'hanke',               COUNT(*) FROM hanke_inflation
  UNION ALL SELECT 'unemployment_curated', COUNT(*) FROM unemployment_curated
")
```

## Cuba spot check (the original example)

```{r}
#| label: cuba-compare
dbGetQuery(con, "
  SELECT i.year,
         i.cpi_inflation AS imf,
         h.cpi_inflation_hanke AS hanke
  FROM imf_inflation i
  LEFT JOIN hanke_inflation h ON h.iso3c = i.iso3c AND h.year = i.year
  WHERE i.iso3c = 'CUB'
  ORDER BY i.year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
