---
title: "02b — Load IMF + Hanke inflation supplements"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

The World Bank WDI has inflation gaps for the countries that matter most for
political-risk forecasting — Cuba (all years NA), North Korea, parts of Syria
and Eritrea — and politically-managed series for others (Venezuela, Argentina,
Zimbabwe, Lebanon). This notebook lands two manually-curated supplement
tables:

- **`imf_inflation`** — IMF Staff estimates (Article IV consultations + WEO
  projections). These are the conservative end of credible estimates and
  what most journalists cite for "official IMF figures."
- **`hanke_inflation`** — Steve Hanke (Johns Hopkins) Troubled Currencies
  Project estimates, derived from free-market exchange-rate decay. These
  are systematically 2-10x higher than IMF/national figures because they
  capture unofficial/black-market reality. Used by CNN and others when
  reporting "hyperinflation."

The panel build (08) does **WDI → IMF backfill** by default. The Hanke
table is loaded separately and exposed to the dashboard as a toggle so
users can see what predictions look like under the more aggressive
estimates.

Data is in `data/curated/*.csv` — both CSVs are version-controlled and updated
manually as new IMF/Hanke reports come out (next update: when IMF WEO April
2025 releases).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Load IMF inflation supplement

```{r}
#| label: imf
imf <- read.csv("../data/curated/imf_inflation_supplement.csv",
                comment.char = "#", stringsAsFactors = FALSE)
imf <- imf[!is.na(imf$cpi_inflation), c("iso3c", "year", "cpi_inflation", "source")]
imf$year <- as.integer(imf$year)
cat("IMF supplement rows:", nrow(imf), "\n")
cat("Countries covered:", length(unique(imf$iso3c)), "\n")
print(table(imf$iso3c))
```

## Load Hanke inflation estimates

```{r}
#| label: hanke
hanke <- read.csv("../data/curated/hanke_inflation_estimates.csv",
                  comment.char = "#", stringsAsFactors = FALSE)
hanke <- hanke[!is.na(hanke$cpi_inflation_hanke),
               c("iso3c", "year", "cpi_inflation_hanke", "source")]
hanke$year <- as.integer(hanke$year)
cat("Hanke estimates rows:", nrow(hanke), "\n")
cat("Countries covered:", length(unique(hanke$iso3c)), "\n")
print(table(hanke$iso3c))
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "imf_inflation",   imf,   overwrite = TRUE)
dbWriteTable(con, "hanke_inflation", hanke, overwrite = TRUE)
dbGetQuery(con, "
  SELECT 'imf' AS source, COUNT(*) AS rows FROM imf_inflation
  UNION ALL
  SELECT 'hanke', COUNT(*) FROM hanke_inflation
")
```

## Cuba spot check (the original example)

```{r}
#| label: cuba-compare
dbGetQuery(con, "
  SELECT i.year,
         i.cpi_inflation AS imf,
         h.cpi_inflation_hanke AS hanke
  FROM imf_inflation i
  LEFT JOIN hanke_inflation h ON h.iso3c = i.iso3c AND h.year = i.year
  WHERE i.iso3c = 'CUB'
  ORDER BY i.year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
