---
title: "02b — Load IMF + Hanke inflation supplements"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

The World Bank WDI has inflation gaps for the countries that matter most for
political-risk forecasting — Cuba (all years NA), North Korea, parts of Syria
and Eritrea — and politically-managed series for others (Venezuela, Argentina,
Zimbabwe, Lebanon). This notebook lands three supplement tables:

- **`imf_inflation`** — hand-curated IMF Staff estimates (Article IV
  consultations + WEO projections). High-trust crisis-country numbers,
  reviewed line-by-line. Covers Cuba (where WEO has nothing) and the
  politically-managed series for Venezuela/Argentina/Zimbabwe/Lebanon.
- **`imf_weo`** — bulk IMF World Economic Outlook database (PCPIPCH series,
  "consumer prices, percent change"). Parsed from the wide-form CSV the
  IMF Data portal exports. Provides continuous 1980–present coverage for
  ~200 countries — the broad backfill behind the hand-curated values.
- **`hanke_inflation`** — Steve Hanke (Johns Hopkins) Troubled Currencies
  Project estimates, derived from free-market exchange-rate decay. These
  are systematically 2-10x higher than IMF/national figures because they
  capture unofficial/black-market reality. Used by CNN and others when
  reporting "hyperinflation."

The panel build (08) uses a **WDI → imf_inflation (curated) → imf_weo
(bulk) → region-median** COALESCE chain. Hand-curated wins over the bulk
file so reviewed crisis values aren't overwritten by WEO defaults. The
Hanke table is loaded separately and exposed to the dashboard as a toggle.

`data/curated/*.csv` is version-controlled (hand-maintained).
`data/raw/imf_weo_*.csv` is gitignored (large, re-downloadable) — pull
from https://data.imf.org/en/datasets/IMF.RES:WEO when refreshing.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Load IMF inflation supplement

```{r}
#| label: imf
imf <- read.csv("../data/curated/imf_inflation_supplement.csv",
                comment.char = "#", stringsAsFactors = FALSE)
imf <- imf[!is.na(imf$cpi_inflation), c("iso3c", "year", "cpi_inflation", "source")]
imf$year <- as.integer(imf$year)
cat("IMF supplement rows:", nrow(imf), "\n")
cat("Countries covered:", length(unique(imf$iso3c)), "\n")
print(table(imf$iso3c))
```

## Load IMF WEO bulk file

The WEO CSV is wide: ~65 metadata columns followed by year columns 1980–2031.
Each row is one (country, indicator) pair. We filter to PCPIPCH (CPI percent
change), pivot to long, drop the IMF country-aggregates (codes ending in
`Q`, e.g. CAQ = Caribbean) since the panel only knows ISO3 country codes.

```{r}
#| label: weo
weo_path <- "../data/raw/imf_weo_2026_04.csv"

if (!file.exists(weo_path)) {
  stop("Missing ", weo_path,
       ". Download the bulk CSV from https://data.imf.org/en/datasets/IMF.RES:WEO ",
       "and save to that path. The file is gitignored (~26MB).")
}

weo_wide <- read.csv(weo_path, check.names = FALSE, stringsAsFactors = FALSE,
                     fileEncoding = "UTF-8-BOM", na.strings = c("", "NA"))

# Filter to the CPI percent-change series.
weo_wide <- weo_wide[grepl("\\.PCPIPCH\\.A$", weo_wide$SERIES_CODE), ]
cat("PCPIPCH series in WEO file:", nrow(weo_wide), "\n")

# Extract ISO3 from SERIES_CODE (first dotted segment).
weo_wide$iso3c <- sub("\\..*", "", weo_wide$SERIES_CODE)

# Drop IMF country-aggregate codes (4th char Q in 3-letter codes won't happen;
# the aggregates use codes like CAQ/CBQ/CMQ that end in Q). Real ISO3 codes
# never end in Q, so this is safe.
weo_wide <- weo_wide[!grepl("Q$", weo_wide$iso3c), ]

# Identify year columns (1980-2031 as strings).
year_cols <- grep("^(19|20)[0-9]{2}$", names(weo_wide), value = TRUE)

# Reshape to long form.
weo_long <- do.call(rbind, lapply(year_cols, function(y) {
  data.frame(
    iso3c          = weo_wide$iso3c,
    year           = as.integer(y),
    cpi_inflation  = suppressWarnings(as.numeric(weo_wide[[y]])),
    stringsAsFactors = FALSE
  )
}))
weo_long <- weo_long[!is.na(weo_long$cpi_inflation), ]
weo_long$source <- "IMF WEO April 2026 (data.imf.org bulk export)"

cat("WEO long rows:", nrow(weo_long), "\n")
cat("Distinct countries:", length(unique(weo_long$iso3c)), "\n")
cat("Year span:", min(weo_long$year), "-", max(weo_long$year), "\n")
```

## Load Hanke inflation estimates

```{r}
#| label: hanke
hanke <- read.csv("../data/curated/hanke_inflation_estimates.csv",
                  comment.char = "#", stringsAsFactors = FALSE)
hanke <- hanke[!is.na(hanke$cpi_inflation_hanke),
               c("iso3c", "year", "cpi_inflation_hanke", "source")]
hanke$year <- as.integer(hanke$year)
cat("Hanke estimates rows:", nrow(hanke), "\n")
cat("Countries covered:", length(unique(hanke$iso3c)), "\n")
print(table(hanke$iso3c))
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "imf_inflation",   imf,      overwrite = TRUE)
dbWriteTable(con, "imf_weo",         weo_long, overwrite = TRUE)
dbWriteTable(con, "hanke_inflation", hanke,    overwrite = TRUE)
dbGetQuery(con, "
  SELECT 'imf_curated' AS source, COUNT(*) AS rows FROM imf_inflation
  UNION ALL
  SELECT 'imf_weo',     COUNT(*) FROM imf_weo
  UNION ALL
  SELECT 'hanke',       COUNT(*) FROM hanke_inflation
")
```

## Cuba spot check (the original example)

```{r}
#| label: cuba-compare
dbGetQuery(con, "
  SELECT i.year,
         i.cpi_inflation AS imf,
         h.cpi_inflation_hanke AS hanke
  FROM imf_inflation i
  LEFT JOIN hanke_inflation h ON h.iso3c = i.iso3c AND h.year = i.year
  WHERE i.iso3c = 'CUB'
  ORDER BY i.year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
