---
title: "03a — Load Mass Mobilization Project (MMP)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** Mass Mobilization Project (MMP) is one of the
canonical sources for "did a large protest happen here this year?"
across the world, 1990-2020. Each row in the raw data is one protest
event; we aggregate to country-year counts so it fits the panel
grain. MMP is one of the OR-fused channels driving
`outcome_mobilization` in `09_label_events.qmd`.

**Data-analytics concepts shown here.**

- *Event-level data → panel aggregation*. The raw data has one row
  per event with timestamps; the panel needs one row per (country,
  year). We `aggregate(... ~ iso3c + year, FUN = sum)` to compress.
- *Per-capita thresholds, not absolute thresholds*. A 1,000-person
  rally is country-defining in Iceland (0.26% of population) but
  rounding-error in India (0.00007%). The right threshold for "large
  protest" has to be **relative to population**, not absolute. We use
  0.1% of population, calibrated against Iran's 2022 WLF baseline.
- *ID system conversion*. MMP uses COW (Correlates of War) numeric
  codes, not ISO-3. `cow_to_iso3()` in `_helpers.R` does the
  conversion using `countrycode`.
- *Schema-drift assertion*. The code asserts the expected MMP size
  bucket names (`"1000-1999"`, `"2000-4999"`, etc.) match what's in
  the downloaded file. If MMP changes their coding scheme, the
  pipeline fails loudly rather than silently producing wrong results.

**Domain terms used here.**

- **MMP** — Mass Mobilization Project, Clark & Regan, Harvard
  Dataverse. Hand-coded protest events 1990-2020.
- **COW code** — Correlates of War country-numbering system.
- **WLF** — Woman, Life, Freedom. Iran's 2022 protest movement, used
  as the calibration baseline for "country-defining mobilization."

**What success looks like.** The `mmp_annual` table has one row per
(country, year) with three counts: total protests, large protests
(per-capita threshold), and violent protests. The number that matters
downstream is `n_large_protests` — used by `09_label_events.qmd` as a
mobilization channel.
:::

Hand-coded large-protest events 1990–2020 from Clark & Regan's Mass
Mobilization Project (Harvard Dataverse). Each row = one protest event with
size, violence, and date fields. We aggregate to **country-year counts**:

- `n_protests` — total protests this year
- `n_large_protests` — protests where `participants ≥ 0.1% of population`.
  Per-capita threshold, not absolute. v0.5.1 fix: previous version flagged
  any protest ≥1,000 participants ("Tuesday outside a courthouse" in big
  countries) or ≥10,000 (rounding error in India, but a mass event in
  Iceland). 0.1% of population catches Yellow Vests France (0.4%), Iran 2022
  Hijab protests (0.1–1%), Iceland 2008 (2.2%), HK 2019 (27%), USA BLM 2020
  (5–8%); excludes routine US/UK/EU rallies that don't destabilize regimes.
  We use MMP's bucket lower bound as a conservative participant estimate.
- `n_violent_protests` — protests where `protesterviolence = 1`

Outcome use (in `09_label_events.qmd`): a country-year is flagged as unrest
if it has any per-capita-large protest next year.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(countrycode)
source("_helpers.R")
```

## Download (Harvard Dataverse)

```{r}
#| label: fetch
dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
mmp_local <- "../data/raw/mmp_all.tab"
if (!file.exists(mmp_local) || file.info(mmp_local)$size < 1000) {
  download.file(
    "https://dataverse.harvard.edu/api/access/datafile/4291456",
    mmp_local, mode = "wb", quiet = TRUE
  )
}
```

## Parse + aggregate

```{r}
#| label: parse
mmp <- read.delim(mmp_local, sep = "\t", stringsAsFactors = FALSE, quote = "\"")
cat("Raw events:", nrow(mmp), " | Year range:", range(mmp$year, na.rm = TRUE), "\n")
table(mmp$participants_category, useNA = "ifany")
```

### Per-capita "large protest" threshold (calibrated to Iran 2022 WLF baseline)

A 1,000-person rally is a meaningful national event in Iceland (0.26% of
population) but rounding-error in India (0.00007%). We need a per-capita
threshold, not an absolute one.

**Threshold: 0.1% of population.** Calibrated to Iran's 2022 Woman-Life-Freedom
protest baseline (single-event peaks ~0.1–1% of national population). At
this threshold, the following events get flagged: Yellow Vests France 2018
(0.4%), Hong Kong 2019 (27%), Iceland 2008 (2.2%), US BLM 2020 (5–8%), Iran
2009 Green Movement peak (0.68%), Lebanon 2019 (14.7%). Routine 1k-vigil
events in large countries are correctly excluded.

We pull WDI population per (iso3c, year), use MMP's bucket lower bound as
a conservative participant estimate, and flag events where
`bucket_lower_bound / population ≥ 0.001`.

```{r}
#| label: aggregate
size_buckets <- c("1000-1999", "2000-4999", "5000-10000", ">10000")
bucket_lower <- c("1000-1999" = 1000, "2000-4999" = 2000,
                  "5000-10000" = 5000, ">10000" = 10000)

present <- intersect(size_buckets, unique(mmp$participants_category))
if (length(present) < length(size_buckets)) {
  stop(sprintf(
    "MMP participants_category schema drift: expected %s, found %s in data. Update size_buckets.",
    paste(size_buckets, collapse = ", "),
    paste(present, collapse = ", ")
  ))
}

mmp$is_violent <- mmp$protesterviolence == 1
mmp$iso3c <- cow_to_iso3(mmp$ccode)
mmp <- mmp[!is.na(mmp$iso3c), ]

# Per-event participant lower bound; events with no bucket info are 0
# (won't ever clear the threshold).
mmp$participants_lower <- bucket_lower[mmp$participants_category]
mmp$participants_lower[is.na(mmp$participants_lower)] <- 0

# Pull WDI population and join.
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
wdi_pop <- dbGetQuery(con, "SELECT iso3c, year, population FROM wdi WHERE population IS NOT NULL")
dbDisconnect(con, shutdown = TRUE)

mmp <- merge(mmp, wdi_pop, by = c("iso3c", "year"), all.x = TRUE)
mmp$participants_pct <- mmp$participants_lower / mmp$population

# 0.1% of population = WLF baseline. Skip events where we don't have
# population data (covers a tiny number of historical edge cases).
PROTEST_PCT_THRESHOLD <- 0.001
mmp$is_large <- !is.na(mmp$participants_pct) & mmp$participants_pct >= PROTEST_PCT_THRESHOLD

cat("Events meeting per-capita threshold:", sum(mmp$is_large), "of", nrow(mmp), "\n")
cat("Events without population data (excluded):", sum(is.na(mmp$participants_pct)), "\n")

mmp_annual <- aggregate(
  cbind(n_protests = rep(1, nrow(mmp)),
        n_large_protests   = mmp$is_large,
        n_violent_protests = mmp$is_violent) ~ iso3c + year,
  data = mmp, FUN = sum
)
mmp_annual$year <- as.integer(mmp_annual$year)
cat("Country-year aggregates:", nrow(mmp_annual), "\n")
head(mmp_annual, 5)
```

## Write

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "mmp_annual", mmp_annual, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year FROM mmp_annual")
```

## Top "large protest" years

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT iso3c, year, n_protests, n_large_protests, n_violent_protests
  FROM mmp_annual
  ORDER BY n_large_protests DESC, n_protests DESC
  LIMIT 10
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
