---
title: "03b — Load UCDP GED armed-conflict events"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

UCDP Georeferenced Event Dataset (GED) v25.1 — hand-coded armed-conflict
events from Uppsala 1989–2024. Each row = one fatal event. We aggregate to
**country-year**:

- `n_events` — count of events
- `fatalities_best` — sum of "best" fatality estimate
- `fatalities_high` — sum of "high" fatality estimate (upper bound)

The CSV is 240 MB raw. We read only the columns we need (`country_id`, `year`,
`best`, `high`, `low`) to keep memory manageable.

Outcome use (in `09_label_events.qmd`): a country-year is flagged as unrest
if `fatalities_best >= 100` next year (UCDP's traditional "minor armed
conflict" threshold is 25; we use 100 to bias toward larger episodes).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(countrycode)
source("_helpers.R")
```

## Download

```{r}
#| label: fetch
dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
ucdp_zip <- "../data/raw/ucdp_ged251.zip"
ucdp_dir <- "../data/raw/ucdp_ged251"
ucdp_csv <- file.path(ucdp_dir, "GEDEvent_v25_1.csv")

if (!file.exists(ucdp_csv)) {
  download.file("https://ucdp.uu.se/downloads/ged/ged251-csv.zip",
                ucdp_zip, mode = "wb", quiet = TRUE)
  unzip(ucdp_zip, exdir = ucdp_dir)
}
```

## Parse + aggregate

Only the 5 columns we need — saves ~200 MB of memory.

```{r}
#| label: parse
ucdp <- read.csv(ucdp_csv,
                 stringsAsFactors = FALSE,
                 colClasses = c("country_id" = "integer", "year" = "integer",
                                "best" = "integer", "high" = "integer", "low" = "integer"),
                 na.strings = c("", "NA"),
                 nrows = -1L)
ucdp <- ucdp[, c("country_id", "year", "best", "high", "low")]
cat("Raw events:", nrow(ucdp), " | Year range:", range(ucdp$year, na.rm = TRUE), "\n")
```

```{r}
#| label: aggregate
ucdp$iso3c <- cow_to_iso3(ucdp$country_id)
ucdp <- ucdp[!is.na(ucdp$iso3c), ]

ucdp_annual <- aggregate(
  cbind(n_events       = rep(1, nrow(ucdp)),
        fatalities_best = ucdp$best,
        fatalities_high = ucdp$high) ~ iso3c + year,
  data = ucdp, FUN = sum
)
ucdp_annual$year <- as.integer(ucdp_annual$year)
cat("Country-year aggregates:", nrow(ucdp_annual), "\n")
head(ucdp_annual, 5)
```

## Write

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "ucdp_annual", ucdp_annual, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year FROM ucdp_annual")
```

## Top fatality years

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT iso3c, year, n_events, fatalities_best
  FROM ucdp_annual
  WHERE year >= 2020
  ORDER BY fatalities_best DESC
  LIMIT 10
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
