---
title: "03c — Load Cline Center Coup d'État Project"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Hand-coded coup events 1945–2026 from the Cline Center at University of
Illinois. Each row = one coup event. We aggregate to **country-year**:

- `n_realized` — successful coups
- `n_attempted` — attempted-but-failed coups
- `n_conspiracy` — uncovered conspiracies that didn't reach the attempt stage

Outcome use (in `09_label_events.qmd`): a country-year is flagged as unrest
if there's any `realized OR attempted` coup next year. Conspiracies are
informational but not part of the binary outcome.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(countrycode)
source("_helpers.R")
```

## Download

The Illinois Data Bank serves files at predictable `/datafiles/{id}/download`
URLs. ID `lpt58` is the main CSV for v2.1.2.

```{r}
#| label: fetch
dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
coup_local <- "../data/raw/cline_coups.csv"
if (!file.exists(coup_local) || file.info(coup_local)$size < 1000) {
  download.file("https://databank.illinois.edu/datafiles/lpt58/download",
                coup_local, mode = "wb", quiet = TRUE)
}
```

## Parse + aggregate

```{r}
#| label: parse
coups <- read.csv(coup_local, stringsAsFactors = FALSE)
cat("Raw coup events:", nrow(coups), " | Year range:", range(coups$year, na.rm = TRUE), "\n")
cat("Event types:\n")
table(coups$event_type, useNA = "ifany")
```

```{r}
#| label: aggregate
coups$iso3c <- cow_to_iso3(coups$cowcode)
coups <- coups[!is.na(coups$iso3c), ]

# Cline's binary flag columns: realized=1 for successful, attempt=1 for
# failed-but-attempted, conspiracy=1 for uncovered plot. A given event row
# carries exactly one of these flags = 1.
coups_annual <- aggregate(
  cbind(n_realized   = coups$realized,
        n_attempted  = coups$attempt,
        n_conspiracy = coups$conspiracy) ~ iso3c + year,
  data = coups, FUN = sum
)
coups_annual$year <- as.integer(coups_annual$year)
cat("Country-year aggregates:", nrow(coups_annual), "\n")
head(coups_annual, 5)
```

## Write

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "coups_annual", coups_annual, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year FROM coups_annual")
```

## Recent coups (2018+)

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT iso3c, year, n_realized, n_attempted
  FROM coups_annual
  WHERE year >= 2018 AND (n_realized > 0 OR n_attempted > 0)
  ORDER BY year DESC, iso3c
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
