---
title: "03d — Load Fariss latent human-rights scores"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Christopher Fariss's Latent Human Rights Protection Scores v4.02 — an
item-response-theory model that combines multiple human-rights coding
sources (Amnesty, State Dept, Hathaway, CIRI, PTS) into a single latent
"physical-integrity rights" score. Coverage: 1946–2021.

Key column: `theta_mean` — latent score (z-like; positive = better human
rights). Lower = more repression. Independent measure complementing
V-Dem's expert-coded civil_liberties.

Source: Harvard Dataverse `doi:10.7910/DVN/RQ85GK` file `LHRS-v4.02-2021.csv`.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(countrycode)
```

## Download

```{r}
#| label: fetch
dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
fariss_local <- "../data/raw/fariss_lhrs_v4.csv"
if (!file.exists(fariss_local) || file.info(fariss_local)$size < 1000) {
  download.file("https://dataverse.harvard.edu/api/access/datafile/7209241",
                fariss_local, mode = "wb", quiet = TRUE)
}
```

## Parse + crosswalk

```{r}
#| label: parse
fariss <- read.csv(fariss_local, stringsAsFactors = FALSE)
cat("Columns:\n"); print(names(fariss))
cat("\nRows:", nrow(fariss), " | Year range:", range(fariss$YEAR, na.rm = TRUE), "\n")

# Convert COW code to ISO3. Fariss uses `COW` column.
fariss$iso3c <- countrycode(fariss$COW, "cown", "iso3c", warn = FALSE)
fariss <- fariss[!is.na(fariss$iso3c), c("iso3c", "YEAR", "theta_mean", "theta_sd")]
names(fariss) <- c("iso3c", "year", "hr_score", "hr_score_sd")
fariss$year <- as.integer(fariss$year)
head(fariss, 3)
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "fariss_hr", fariss, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year FROM fariss_hr")
```

## Spot check: most-repressive countries 2020

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT iso3c, year, ROUND(hr_score, 2) AS hr_score
  FROM fariss_hr
  WHERE year = 2020
  ORDER BY hr_score ASC
  LIMIT 10
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
