---
title: "03e — Load Polity5 democracy-autocracy scores"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** Polity is the oldest cross-country regime
classification, used in political science for 30+ years. The `polity2`
score (−10 full autocracy to +10 full democracy) is a standard
regression covariate. We load it for analysis but exclude it from the
v0.4+ model fit for two reasons: **coverage ends in 2018** and it's
**~0.9 correlated with V-Dem's civil-liberties index** (same
collinearity issue as Fariss).

**Data-analytics concepts shown here.**

- *Composite scores from sub-scores*. Polity's `democ` (0-10) minus
  `autoc` (0-10) gives the headline `polity2` (-10 to +10). Two-axis
  scales reduced to one for downstream use.
- *Excel as a data format*. Polity ships as an XLS file (older Excel
  binary format). `readxl::read_excel` handles both `.xls` and `.xlsx`
  transparently.
- *Coverage cliffs*. When a dataset ends in 2018 and your panel runs
  to 2025, you have to decide: forward-fill the last known value? Use
  the variable as historical-only? Drop the variable? We chose
  "exclude from fit" because forward-filling 7 years feels worse than
  losing the variable.
- *Comparing competing measures of the same concept*. Polity, V-Dem
  regime classification, and Fariss all measure aspects of political
  freedom. They mostly agree. Picking one and excluding the rest is a
  judgment call.

**Domain terms used here.**

- **Polity / Polity5** — the project name and its latest release.
  Maintained by the Center for Systemic Peace.
- **`polity2`** — the composite score (democ − autoc), range -10 to +10.
- **Interregnum / interruption / transition** — special codes in Polity
  for regime transitions where the score is missing or ambiguous.

**What success looks like.** The `polity5` table loads with one row
per (country, year) for 1800-2018. Excluded from the v0.4+ model fit.
If V-Dem's coverage ever fails, Polity could fill in.
:::

Polity5 from Marshall et al. / Center for Systemic Peace. Annual democracy
(0–10) and autocracy (0–10) scores combined into the `polity2` score
ranging −10 (full autocracy) to +10 (full democracy). Coverage: 1800–2018.

Independent of V-Dem's expert-coding methodology — useful as a cross-check
on regime classification.

Source: `https://www.systemicpeace.org/inscr/p5v2018.xls`.

## Libraries

```{r}
#| label: setup
library(readxl)
library(DBI)
library(duckdb)
library(countrycode)
source("_helpers.R")
```

## Download

```{r}
#| label: fetch
dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
polity_local <- "../data/raw/polity5_v2018.xls"
if (!file.exists(polity_local) || file.info(polity_local)$size < 1000) {
  download.file("https://www.systemicpeace.org/inscr/p5v2018.xls",
                polity_local, mode = "wb", quiet = TRUE)
}
```

## Parse + crosswalk

```{r}
#| label: parse
polity <- read_excel(polity_local)
cat("Columns (first 20):\n"); print(names(polity)[1:min(20, ncol(polity))])
cat("\nRows:", nrow(polity), " | Year range:", range(polity$year, na.rm = TRUE), "\n")

polity$iso3c <- cow_to_iso3(polity$ccode)
polity <- as.data.frame(polity)
polity <- polity[!is.na(polity$iso3c), c("iso3c", "year", "polity2", "durable")]
polity$year <- as.integer(polity$year)
# polity2 sometimes carries -66/-77/-88 to indicate transitions/interregnum;
# treat those as NA.
polity$polity2[polity$polity2 < -10 | polity$polity2 > 10] <- NA
head(polity, 3)
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "polity5", polity, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year FROM polity5")
```

## Spot check: latest polity2 distribution

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT iso3c, polity2, durable
  FROM polity5
  WHERE year = 2018 AND polity2 IS NOT NULL
  ORDER BY polity2 ASC
  LIMIT 10
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
