---
title: "03e — Load Polity5 democracy-autocracy scores"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Polity5 from Marshall et al. / Center for Systemic Peace. Annual democracy
(0–10) and autocracy (0–10) scores combined into the `polity2` score
ranging −10 (full autocracy) to +10 (full democracy). Coverage: 1800–2018.

Independent of V-Dem's expert-coding methodology — useful as a cross-check
on regime classification.

Source: `https://www.systemicpeace.org/inscr/p5v2018.xls`.

## Libraries

```{r}
#| label: setup
library(readxl)
library(DBI)
library(duckdb)
library(countrycode)
```

## Download

```{r}
#| label: fetch
dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
polity_local <- "../data/raw/polity5_v2018.xls"
if (!file.exists(polity_local) || file.info(polity_local)$size < 1000) {
  download.file("https://www.systemicpeace.org/inscr/p5v2018.xls",
                polity_local, mode = "wb", quiet = TRUE)
}
```

## Parse + crosswalk

```{r}
#| label: parse
polity <- read_excel(polity_local)
cat("Columns (first 20):\n"); print(names(polity)[1:min(20, ncol(polity))])
cat("\nRows:", nrow(polity), " | Year range:", range(polity$year, na.rm = TRUE), "\n")

polity$iso3c <- countrycode(polity$ccode, "cown", "iso3c", warn = FALSE)
polity <- as.data.frame(polity)
polity <- polity[!is.na(polity$iso3c), c("iso3c", "year", "polity2", "durable")]
polity$year <- as.integer(polity$year)
# polity2 sometimes carries -66/-77/-88 to indicate transitions/interregnum;
# treat those as NA.
polity$polity2[polity$polity2 < -10 | polity$polity2 > 10] <- NA
head(polity, 3)
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "polity5", polity, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year FROM polity5")
```

## Spot check: latest polity2 distribution

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT iso3c, polity2, durable
  FROM polity5
  WHERE year = 2018 AND polity2 IS NOT NULL
  ORDER BY polity2 ASC
  LIMIT 10
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
