---
title: "03f — Load ACLED regional event-aggregated data"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** ACLED (Armed Conflict Location & Event Data) is
the canonical real-time event source in political science. We use it
for two purposes: as an OR-fused channel in both outcome labels
(deadly mobilization and total violence ≥100 fatalities), and as
continuous-fatality-lag features in the model fit. ACLED narrows
(though doesn't eliminate) the media-availability bias that affects
all event data.

**Data-analytics concepts shown here.**

- *Choosing the right level of aggregation*. The six regional files
  are (week × admin1 × event_type × sub_event_type) cells. That's
  fine enough that we can roll up to country-year however we want,
  but pre-aggregated enough to be tractable (~950K rows total). The
  five country-year aggregate files we initially used (deleted in
  v0.7.1) had collapsed `sub_event_type` and broke the outcome label.
  **The level of aggregation in your source data determines what
  analyses you can do.**
- *Sub-event types matter*. ACLED's "Violent demonstration",
  "Excessive force against protesters", "Peaceful protest", and
  "Mob violence" are different things. Collapsing them into "demos"
  loses the discrimination we need.
- *Fatalities as the bias-resistant currency*. Open democracies get
  more news coverage so more *events* get coded; closed regimes are
  under-counted in event counts but death counts are more robust
  (hospitals, families, opposition track them). The pipeline only
  uses ACLED *fatalities* as outcome-label triggers, never raw event
  counts — this is the design lesson from v0.7-rev0's failure.
- *License-aware data handling*. ACLED's terms restrict
  redistribution. We gitignore the raw files in `data/raw/acled/` and
  only commit derived country-year aggregates and binary outcome
  flags. Same legal posture matters in industry contracts.

**Domain terms used here.**

- **ACLED** — Armed Conflict Location & Event Data Project.
  https://acleddata.com
- **Sub-event type** — ACLED's fine-grained categorization of events
  (25+ categories). Examples: Peaceful protest, Violent demonstration,
  Protest with intervention, Excessive force against protesters,
  Armed clash, Air/drone strike.
- **Admin1** — ACLED's geographic unit one level below country
  (provinces, states). Useful for "spread across the country vs.
  localized" features.
- **POPULATION_EXPOSURE** — ACLED's estimate of population in the
  area exposed to events. Built-in denominator for per-capita
  features.
- **Media-availability bias** — the systematic tendency for event-data
  sources to over-count events in places with active media and
  under-count them where media is suppressed.

**What success looks like.** The `acled_annual` table has one row per
(country, year) with fatality and event counts broken out by sub-event
type plus flags for coverage and year-completeness. ~12,000 rows
across ~245 countries. Diagnostics confirm Iran 2022's 421
violent-demonstration fatalities and DPRK's near-zero coverage (which
ACLED can't fix because the censoring is upstream).
:::

ACLED Research-tier API access was denied; ACLED support clarified that
**dashboard read access** (and AI agents reading on the user's behalf) is
permitted — only programmatic API use is restricted. We pull the six
**regional weekly+admin1 aggregate files** from the ACLED dashboard's
"Aggregated Data" page (one per region: Africa, Asia-Pacific, Europe &
Central Asia, Latin America & Caribbean, Middle East, US & Canada).

Each row in a regional file is a (week × admin1 × event_type ×
sub_event_type) cell with EVENTS, FATALITIES, and POPULATION_EXPOSURE
counts. The 950K combined rows are essentially event-level data
pre-aggregated to a level fine enough that we can roll up however we want.

**Why this replaces the v0.7-rev0 country-year aggregates** (now deleted):
those files collapsed all sub-event types into a single count, which hid
the discriminating signal between "country has active civil society"
(thousands of Peaceful protests) vs. "country had a state-protester
confrontation episode" (Violent demonstration with fatalities, Excessive
force against protesters). The regional files preserve `SUB_EVENT_TYPE`
which lets us build the distinction directly.

**Coverage** (varies by region — ACLED's geographic expansion timeline):

| Region | Coverage starts |
|---|---|
| Africa | 1996 |
| Asia-Pacific (incl. South Asia, SE Asia, Oceania) | 2009-2020 (country-varying) |
| Middle East | 2014 |
| Europe & Central Asia | 2017 |
| Latin America & Caribbean | 2017 |
| US & Canada | 2019 |

**Files live in `data/raw/acled/regional/`, gitignored.** ACLED's
no-redistribution rule applies to raw rows; only country-year derived
aggregates land in DuckDB and only binary outcome flags / continuous
features land in `artifacts/`.

**Bias note (documented for downstream readers):** event-data sources
are subject to media-availability bias. Open democracies have more
journalism covering protests so individual events get coded
more granularly; closed regimes (PRK, Eritrea, Turkmenistan; partly
Cuba, partly Iran) have fewer events coded because the news pipeline
underrepresents them. **Fatalities are the bias-resistant currency** —
bodies are tracked across regimes more uniformly than events, because
hospitals, families, and opposition groups document deaths even when
the state suppresses event-level reports. This pipeline therefore uses
fatality-based thresholds and features wherever possible, and treats
event counts as auxiliary (high-event + low-fatality = civil-society
activity, not regime-threatening mobilization).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(readxl)
library(countrycode)
source("_helpers.R")
```

## Locate + read all six regional files

```{r}
#| label: locate-and-read
acled_dir <- "../data/raw/acled/regional"
stopifnot(dir.exists(acled_dir))

files <- list.files(acled_dir, pattern = "\\.xlsx$", full.names = TRUE)
cat("Regional files found:\n")
print(basename(files))
stopifnot(length(files) >= 6)

# Each regional file shares the same 13-column schema:
#   WEEK, REGION, COUNTRY, ADMIN1, EVENT_TYPE, SUB_EVENT_TYPE,
#   EVENTS, FATALITIES, POPULATION_EXPOSURE, DISORDER_TYPE,
#   ID, CENTROID_LATITUDE, CENTROID_LONGITUDE
all_frames <- lapply(files, function(f) {
  df <- readxl::read_excel(f)
  cat(sprintf("  %s: %d rows, %d countries\n",
              basename(f), nrow(df), length(unique(df$COUNTRY))))
  df
})
acled_raw <- do.call(rbind, all_frames)

# Coerce date and extract year (the panel grain).
acled_raw$WEEK <- as.Date(acled_raw$WEEK)
acled_raw$year <- as.integer(format(acled_raw$WEEK, "%Y"))

cat(sprintf("\nCombined: %d rows, %d countries, date range %s -> %s\n",
            nrow(acled_raw), length(unique(acled_raw$COUNTRY)),
            min(acled_raw$WEEK), max(acled_raw$WEEK)))
```

## Map country names → iso3c

```{r}
#| label: iso3c-map
acled_raw$iso3c <- countrycode::countrycode(
  acled_raw$COUNTRY, origin = "country.name", destination = "iso3c", warn = FALSE
)

unmapped <- unique(acled_raw$COUNTRY[is.na(acled_raw$iso3c)])
cat("Unmapped country names (dropped — typically marine areas, Antarctica):\n")
print(unmapped)

acled_raw <- acled_raw[!is.na(acled_raw$iso3c), ]
cat(sprintf("Rows after iso3c map: %d (countries=%d)\n",
            nrow(acled_raw), length(unique(acled_raw$iso3c))))
```

## Build country-year aggregates

We aggregate from (week × admin1 × event_type × sub_event_type) up to
country-year, separately for each sub_event_type that matters
substantively. The result is a wide table with one row per (iso3c, year)
and many count/fatality columns — predictable for downstream joins.

```{r}
#| label: aggregate

# Helper: sum EVENTS and FATALITIES across rows matching a sub-event-type filter
sum_by <- function(df, subtype_pattern) {
  subset_df <- df[grepl(subtype_pattern, df$SUB_EVENT_TYPE, fixed = TRUE), ]
  if (nrow(subset_df) == 0) {
    return(data.frame(iso3c = character(), year = integer(),
                      events = integer(), fatalities = integer()))
  }
  agg <- aggregate(
    cbind(events = subset_df$EVENTS, fatalities = subset_df$FATALITIES) ~ iso3c + year,
    data = subset_df, FUN = sum
  )
  agg
}

# Master grid: every (iso3c, year) present in the data
grid <- unique(acled_raw[, c("iso3c", "year")])
grid <- grid[order(grid$iso3c, grid$year), ]

# Helper: left-join an aggregate onto grid, prefix the count/fatality cols
merge_in <- function(grid, agg, prefix) {
  if (nrow(agg) == 0) {
    grid[[paste0("n_", prefix)]] <- 0L
    grid[[paste0("fat_", prefix)]] <- 0L
    return(grid)
  }
  names(agg)[names(agg) == "events"]    <- paste0("n_",   prefix)
  names(agg)[names(agg) == "fatalities"] <- paste0("fat_", prefix)
  out <- merge(grid, agg, by = c("iso3c", "year"), all.x = TRUE)
  out[[paste0("n_",   prefix)]][is.na(out[[paste0("n_",   prefix)]])] <- 0L
  out[[paste0("fat_", prefix)]][is.na(out[[paste0("fat_", prefix)]])] <- 0L
  out
}

# ----- DEMONSTRATION sub-types (Protests + Riots event_types) -----
# Peaceful protest:                            "civil society activity" — uses fatalities only as signal
# Violent demonstration:                        protesters used force or protest turned deadly
# Protest with intervention:                    state forces actively dispersed/contained
# Excessive force against protesters:           state response went beyond standard policing
# Mob violence:                                 disorganized crowd violence (often not anti-regime)
# Riots:                                        catch-all for any Riot-classified event_type

grid <- merge_in(grid, sum_by(acled_raw, "Peaceful protest"),                  "peaceful_protest")
grid <- merge_in(grid, sum_by(acled_raw, "Violent demonstration"),             "violent_demo")
grid <- merge_in(grid, sum_by(acled_raw, "Protest with intervention"),         "protest_intervention")
grid <- merge_in(grid, sum_by(acled_raw, "Excessive force against protesters"), "excessive_force")
grid <- merge_in(grid, sum_by(acled_raw, "Mob violence"),                      "mob_violence")

# ----- POLITICAL VIOLENCE sub-types (Battles, Explosions, VAC) -----
# Battle (state armed clash):                   conventional conflict
# Air/drone strike, Shelling, Remote IED:       remote violence
# Attack:                                       targeted attack (mostly insurgent)
# Civilian targeting (special: a flag, not event_type):
grid <- merge_in(grid, sum_by(acled_raw, "Armed clash"),         "armed_clash")
grid <- merge_in(grid, sum_by(acled_raw, "Air/drone strike"),    "air_strike")
grid <- merge_in(grid, sum_by(acled_raw, "Shelling/artillery"),  "shelling")
grid <- merge_in(grid, sum_by(acled_raw, "Attack"),              "attack")
grid <- merge_in(grid, sum_by(acled_raw, "Abduction"),           "abduction")
grid <- merge_in(grid, sum_by(acled_raw, "Sexual violence"),     "sexual_violence")

# ----- DISORDER_TYPE-level totals (the broadest buckets) -----
# These match what the country-year aggregate files had — useful as a
# sanity check that the new method reproduces the old totals.
dt_agg <- aggregate(
  cbind(events = acled_raw$EVENTS, fatalities = acled_raw$FATALITIES) ~ iso3c + year + DISORDER_TYPE,
  data = acled_raw, FUN = sum
)

dt_pv  <- dt_agg[dt_agg$DISORDER_TYPE == "Political violence", c("iso3c","year","events","fatalities")]
dt_dem <- dt_agg[dt_agg$DISORDER_TYPE == "Demonstrations",      c("iso3c","year","events","fatalities")]

names(dt_pv)[3:4]  <- c("n_political_violence_events", "fat_political_violence_events")
names(dt_dem)[3:4] <- c("n_demonstration_events",      "fat_demonstration_events")

grid <- merge(grid, dt_pv,  by = c("iso3c","year"), all.x = TRUE)
grid <- merge(grid, dt_dem, by = c("iso3c","year"), all.x = TRUE)
for (col in c("n_political_violence_events", "fat_political_violence_events",
              "n_demonstration_events", "fat_demonstration_events")) {
  grid[[col]][is.na(grid[[col]])] <- 0
}

# ----- Total all-event fatalities (replaces old `n_fatalities` aggregate) -----
total_fat <- aggregate(
  cbind(n_events = acled_raw$EVENTS, fatalities = acled_raw$FATALITIES) ~ iso3c + year,
  data = acled_raw, FUN = sum
)
names(total_fat)[3:4] <- c("n_total_events", "fat_total")
grid <- merge(grid, total_fat, by = c("iso3c","year"), all.x = TRUE)
grid$n_total_events[is.na(grid$n_total_events)] <- 0
grid$fat_total[is.na(grid$fat_total)] <- 0

# ----- Geographic concentration: how many distinct admin1 had violent-demos? -----
# Spread across many provinces vs. one is the country-wide-vs-localized signal.
adm1_vd <- acled_raw[acled_raw$SUB_EVENT_TYPE == "Violent demonstration" & acled_raw$EVENTS > 0, ]
if (nrow(adm1_vd) > 0) {
  adm1_count <- aggregate(ADMIN1 ~ iso3c + year, data = adm1_vd,
                          FUN = function(x) length(unique(x)))
  names(adm1_count)[3] <- "n_admin1_violent_demo"
  grid <- merge(grid, adm1_count, by = c("iso3c","year"), all.x = TRUE)
} else {
  grid$n_admin1_violent_demo <- 0L
}
grid$n_admin1_violent_demo[is.na(grid$n_admin1_violent_demo)] <- 0L

# ----- Coverage flags -----
# acled_covered: 1 for any country-year where ACLED has any row (since grid
# was built from acled_raw, every grid row is covered by construction).
grid$acled_covered <- 1L

# acled_year_complete: 0 for 2026 (partial-year extract, ~through May)
ACLED_EXTRACT_YEAR <- 2026L
grid$acled_year_complete <- as.integer(grid$year < ACLED_EXTRACT_YEAR)

# Order columns predictably
acled_annual <- grid[order(grid$iso3c, grid$year), ]
cat(sprintf("\nacled_annual: %d rows × %d columns\n",
            nrow(acled_annual), ncol(acled_annual)))
str(acled_annual)
```

## Write

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "acled_annual", acled_annual, overwrite = TRUE)
dbGetQuery(con, "
  SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year,
         COUNT(DISTINCT iso3c) AS countries
  FROM acled_annual
")
```

## Coverage timeline by region

```{r}
#| label: coverage-timeline
dbGetQuery(con, "
  SELECT year,
         COUNT(DISTINCT iso3c) AS n_countries_covered,
         SUM(n_demonstration_events)        AS total_demos,
         SUM(n_violent_demo)                AS total_violent_demos,
         SUM(fat_violent_demo)              AS total_violent_demo_fatalities,
         SUM(fat_political_violence_events) AS total_pv_fatalities
  FROM acled_annual
  GROUP BY year
  ORDER BY year
")
```

## Sanity — canonical mobilization episodes

Confirm `fat_violent_demo` is the discriminating signal that the
fatalities-only outcome label will key on.

```{r}
#| label: sanity-mob
dbGetQuery(con, "
  SELECT iso3c, year,
         n_peaceful_protest, n_violent_demo, fat_violent_demo,
         n_protest_intervention, n_excessive_force, fat_excessive_force
  FROM acled_annual
  WHERE (iso3c = 'IRN' AND year BETWEEN 2020 AND 2024)
     OR (iso3c = 'CUB' AND year BETWEEN 2020 AND 2024)
     OR (iso3c = 'USA' AND year BETWEEN 2019 AND 2024)
     OR (iso3c = 'FRA' AND year BETWEEN 2020 AND 2024)
  ORDER BY iso3c, year
")
```

## Sanity — closed-regime censoring is visible

PRK / Eritrea / Turkmenistan should have near-zero coverage, which is
the censoring problem ACLED can't fix (but does make measurable). Cuba
2021 11J is a good example: only 1 violent-demo fatality coded, but
38 protest-with-intervention events flagged → the channel is detecting
state-protester confrontation, even when fatality count is low.

```{r}
#| label: sanity-censoring
dbGetQuery(con, "
  SELECT iso3c, year, n_peaceful_protest, n_violent_demo, fat_violent_demo,
         n_protest_intervention, fat_excessive_force
  FROM acled_annual
  WHERE iso3c IN ('PRK','ERI','TKM','CUB','BLR','TJK')
    AND year >= 2020
  ORDER BY iso3c, year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
