---
title: "04 — Load UNHCR refugee statistics"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Pulls annual end-of-year refugee + asylum-seeker counts by country of origin
from UNHCR's public Refugee Statistics API. Lands as `unhcr` in
`data/panel.duckdb`. One row per (origin, year) — refugee outflow is the
origin-side political-stress signal we want, so we aggregate across the
asylum dimension at the API level.

## Install

```{r}
#| label: install
#| eval: false
install.packages(c("httr2", "jsonlite"))
```

The API is fully public — no key, no OAuth.

## Libraries

```{r}
#| label: setup
library(httr2)
library(jsonlite)
library(DBI)
library(duckdb)
library(ggplot2)
```

## Fetch all origins, all years

Setting `coo_all=true` without `coa_all=true` makes UNHCR collapse the
asylum dimension. ~200 origin countries × ~35 years ≈ 7K rows.

```{r}
#| label: fetch
base <- "https://api.unhcr.org/population/v1/population/"

fetch_page <- function(page, page_size = 1000) {
  request(base) |>
    req_url_query(coo_all = "true",
                  yearFrom = 1990,
                  yearTo   = format(Sys.Date(), "%Y"),
                  limit    = page_size,
                  page     = page) |>
    req_perform() |>
    resp_body_json(simplifyVector = TRUE)
}

first <- fetch_page(1)
all_rows <- if (first$maxPages > 1) {
  rest <- lapply(2:first$maxPages, function(p) fetch_page(p)$items)
  do.call(rbind, c(list(first$items), rest))
} else first$items

cat("Pages:", first$maxPages, " | Total rows:", nrow(all_rows), "\n")
```

## Tidy

UNHCR returns count fields as character because `"-"` distinguishes "not
collected" from a real zero. Convert to integer with `"-"` → `NA`.

```{r}
#| label: tidy
count_cols <- c("refugees", "asylum_seekers", "returned_refugees",
                "idps", "returned_idps", "stateless", "ooc")

to_int <- function(x) { x[x == "-"] <- NA; suppressWarnings(as.integer(x)) }

unhcr <- all_rows
for (col in count_cols) if (col %in% names(unhcr)) unhcr[[col]] <- to_int(unhcr[[col]])
unhcr$year <- as.integer(unhcr$year)
unhcr <- unhcr[, c("year", "coo", "coo_iso", "coo_name",
                   "refugees", "asylum_seekers", "idps", "stateless", "ooc")]
head(unhcr, 3)
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "unhcr", unhcr, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows FROM unhcr")
```

## Coverage

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT
    COUNT(DISTINCT coo_iso) AS origin_countries,
    MIN(year) AS earliest_year,
    MAX(year) AS latest_year,
    SUM(refugees)       AS total_refugees_all_years,
    SUM(asylum_seekers) AS total_asylum_seekers_all_years
  FROM unhcr
")
```

## Cuba sanity check

```{r}
#| label: cuba-recent
cuba_recent <- dbGetQuery(con, "
  SELECT year, refugees, asylum_seekers,
         (COALESCE(refugees,0) + COALESCE(asylum_seekers,0)) AS total_displaced
  FROM unhcr
  WHERE coo_iso = 'CUB' AND year >= 2015
  ORDER BY year
")
cuba_recent
```

```{r}
#| label: cuba-plot
#| fig-cap: "Cubans seeking refuge or asylum abroad, 2015 onward (UNHCR)"
#| fig-width: 7
#| fig-height: 4
parchment <- "#F8F3E8"
ink       <- "#2A2620"
accent    <- "#1F3A60"

ggplot(cuba_recent, aes(x = year, y = total_displaced)) +
  geom_area(fill = accent, alpha = 0.25) +
  geom_line(color = accent, linewidth = 0.9) +
  geom_point(color = accent, size = 2.5) +
  scale_x_continuous(breaks = cuba_recent$year) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Refugees + asylum-seekers (origin: Cuba)",
       caption = "Source: UNHCR Refugee Statistics API") +
  theme_minimal(base_family = "serif", base_size = 12) +
  theme(plot.background  = element_rect(fill = parchment, color = NA),
        panel.background = element_rect(fill = parchment, color = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "#D8CFC0"),
        axis.text        = element_text(color = ink),
        axis.title       = element_text(color = ink),
        plot.caption     = element_text(color = "#6B5D4F", face = "italic"))
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
