---
title: "04 — Load UNHCR refugee statistics"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Pulls annual end-of-year refugee + asylum-seeker counts from UNHCR's public
Refugee Statistics API along two dimensions and writes two tables to
`data/panel.duckdb`:

- `unhcr` — keyed by **country of origin** (one row per origin × year), the
  outflow side. Refugee outflow is the political-stress signal that v0
  originally used.
- `unhcr_inflow` — keyed by **country of asylum** (one row per destination ×
  year), the inflow side. Needed so we can build *net* outflow in
  `08_build_panel`, which kills a reverse-causal contamination where
  large refugee-receiving democracies (USA, Germany) look like Syria
  on a log scale.

## Install

```{r}
#| label: install
#| eval: false
install.packages(c("httr2", "jsonlite"))
```

The API is fully public — no key, no OAuth.

## Libraries

```{r}
#| label: setup
library(httr2)
library(jsonlite)
library(DBI)
library(duckdb)
library(ggplot2)
source("_helpers.R")
```

## Fetch all origins, all years (outflow side)

Setting `coo_all=true` without `coa_all=true` makes UNHCR collapse the
asylum dimension. ~200 origin countries × ~35 years ≈ 7K rows.

```{r}
#| label: fetch
base <- "https://api.unhcr.org/population/v1/population/"

fetch_page <- function(page, page_size = 1000, side = c("coo", "coa")) {
  side <- match.arg(side)
  q <- list(yearFrom = 1990,
            yearTo   = format(Sys.Date(), "%Y"),
            limit    = page_size,
            page     = page)
  q[[paste0(side, "_all")]] <- "true"
  do.call(req_url_query, c(list(request(base)), q)) |>
    req_perform() |>
    resp_body_json(simplifyVector = TRUE)
}

fetch_all <- function(side) {
  first <- fetch_page(1, side = side)
  if (first$maxPages > 1) {
    rest <- lapply(2:first$maxPages, function(p) fetch_page(p, side = side)$items)
    do.call(rbind, c(list(first$items), rest))
  } else first$items
}

all_rows         <- fetch_all("coo")
all_rows_inflow  <- fetch_all("coa")

cat("Origin-side rows:     ", nrow(all_rows), "\n")
cat("Destination-side rows:", nrow(all_rows_inflow), "\n")
```

## Tidy

UNHCR returns count fields as character because `"-"` distinguishes "not
collected" from a real zero. Convert to integer with `"-"` → `NA`.

```{r}
#| label: tidy
count_cols <- c("refugees", "asylum_seekers", "returned_refugees",
                "idps", "returned_idps", "stateless", "ooc")

to_int <- function(x) { x[x == "-"] <- NA; suppressWarnings(as.integer(x)) }

tidy_side <- function(df, side) {
  for (col in count_cols) if (col %in% names(df)) df[[col]] <- to_int(df[[col]])
  df$year <- as.integer(df$year)
  iso_col <- paste0(side, "_iso")
  name_col <- paste0(side, "_name")
  keep <- c("year", side, iso_col, name_col,
            "refugees", "asylum_seekers", "idps", "stateless", "ooc")
  df[, intersect(keep, names(df)), drop = FALSE]
}

unhcr        <- tidy_side(all_rows,        "coo")
unhcr_inflow <- tidy_side(all_rows_inflow, "coa")
head(unhcr, 3)
head(unhcr_inflow, 3)
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "unhcr",        unhcr,        overwrite = TRUE)
dbWriteTable(con, "unhcr_inflow", unhcr_inflow, overwrite = TRUE)
dbGetQuery(con, "SELECT
  (SELECT COUNT(*) FROM unhcr)        AS origin_rows,
  (SELECT COUNT(*) FROM unhcr_inflow) AS destination_rows")
```

## Coverage

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT
    COUNT(DISTINCT coo_iso) AS origin_countries,
    MIN(year) AS earliest_year,
    MAX(year) AS latest_year,
    SUM(refugees)       AS total_refugees_all_years,
    SUM(asylum_seekers) AS total_asylum_seekers_all_years
  FROM unhcr
")
```

## Cuba sanity check

```{r}
#| label: cuba-recent
cuba_recent <- dbGetQuery(con, "
  SELECT year, refugees, asylum_seekers,
         (COALESCE(refugees,0) + COALESCE(asylum_seekers,0)) AS total_displaced
  FROM unhcr
  WHERE coo_iso = 'CUB' AND year >= 2015
  ORDER BY year
")
cuba_recent
```

```{r}
#| label: cuba-plot
#| fig-cap: "Cubans seeking refuge or asylum abroad, 2015 onward (UNHCR)"
#| fig-width: 7
#| fig-height: 4
ggplot(cuba_recent, aes(x = year, y = total_displaced)) +
  geom_area(fill = srs_palette$accent, alpha = 0.25) +
  geom_line(color = srs_palette$accent, linewidth = 0.9) +
  geom_point(color = srs_palette$accent, size = 2.5) +
  scale_x_continuous(breaks = cuba_recent$year) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Refugees + asylum-seekers (origin: Cuba)",
       caption = "Source: UNHCR Refugee Statistics API") +
  srs_theme()
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
