---
title: "05 — Load FAO Food Price Index"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Downloads FAO's monthly Food Price Index (FFPI), 1990–present, plus the five
commodity sub-indices (Meat, Dairy, Cereals, Oils, Sugar). Lands as `fao` in
`data/panel.duckdb`. Unlike the other sources this is a **global** time series,
not country-level — at panel build time it broadcasts to every country on
year+month.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(ggplot2)
source("_helpers.R")
```

## Download

Two-step (download then parse) because R's base `file()` can't open HTTPS
URLs through `read.csv()` on Windows. Persisting the raw CSV at `data/raw/`
also gives us a reproducibility checkpoint if FAO ever changes the URL.

```{r}
#| label: fetch
fao_url <- paste0(
  "https://www.fao.org/media/docs/worldfoodsituationlibraries/",
  "default-document-library/food_price_indices_data.csv",
  "?sfvrsn=523ebd2a_79&download=true"
)

dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
fao_local <- "../data/raw/fao_food_price_indices.csv"
download.file(fao_url, fao_local, mode = "wb", quiet = TRUE)
```

## Parse

The CSV has a publication-style layout: title row, base-period note, header
row, blank separator, then monthly data. We skip the first two rows so
`read.csv()` reads the column headers correctly, then drop the all-NA
separator row downstream.

```{r}
#| label: parse
fao_raw <- read.csv(fao_local, skip = 2, header = TRUE,
                    na.strings = "", check.names = FALSE,
                    stringsAsFactors = FALSE)

# Drop the 50+ trailing empty padding columns; keep the seven real ones.
fao <- fao_raw[, c("Date", "Food Price Index", "Meat", "Dairy",
                   "Cereals", "Oils", "Sugar")]
names(fao) <- c("date", "ffpi", "meat", "dairy", "cereals", "oils", "sugar")
fao <- fao[!is.na(fao$date) & fao$date != "", ]

# Split YYYY-MM into integer year + month plus a Date column at month-start.
parts <- strsplit(fao$date, "-", fixed = TRUE)
fao$year  <- as.integer(vapply(parts, `[`, character(1), 1))
fao$month <- as.integer(vapply(parts, `[`, character(1), 2))
fao$date  <- as.Date(paste0(fao$date, "-01"))

for (col in c("ffpi", "meat", "dairy", "cereals", "oils", "sugar"))
  fao[[col]] <- suppressWarnings(as.numeric(fao[[col]]))

fao <- fao[, c("date", "year", "month", "ffpi", "meat", "dairy", "cereals", "oils", "sugar")]
cat("Rows:", nrow(fao), " | Range:",
    as.character(min(fao$date)), "to", as.character(max(fao$date)), "\n")
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "fao", fao, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows, MIN(year) AS first_year, MAX(year) AS last_year FROM fao")
```

## Peak years sanity check

The three major spikes since 1990: 2008 food-price crisis, 2011 Arab Spring
run-up, 2022 Russia-Ukraine grain disruption.

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT year, ROUND(MAX(ffpi),1) AS peak_ffpi, ROUND(AVG(ffpi),1) AS avg_ffpi
  FROM fao
  WHERE year IN (2008, 2011, 2022)
  GROUP BY year
  ORDER BY year
")
```

```{r}
#| label: ffpi-plot
#| fig-cap: "FAO Food Price Index (2014–2016 = 100), monthly, 1990 onward"
#| fig-width: 8
#| fig-height: 4.2
ffpi <- dbGetQuery(con, "SELECT date, ffpi FROM fao ORDER BY date")
ffpi$date <- as.Date(ffpi$date)

peaks <- data.frame(
  date  = as.Date(c("2008-06-01", "2011-02-01", "2022-03-01")),
  ffpi  = c(132.7, 137.6, 159.7),
  label = c("2008", "2011", "2022")
)

ggplot(ffpi, aes(x = date, y = ffpi)) +
  geom_line(color = srs_palette$accent, linewidth = 0.7) +
  geom_point(data = peaks, color = srs_palette$crisis, size = 3) +
  geom_text(data = peaks, aes(label = label),
            color = srs_palette$crisis, nudge_y = 8, family = "serif", fontface = "bold") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(x = NULL, y = "FAO Food Price Index (2014–2016 = 100)",
       caption = "Source: FAO Food Price Index") +
  srs_theme()
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
