---
title: "06 — Load Pilster-Böhmelt counterbalancing data"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** Authoritarian leaders fragment their militaries to
prevent coups — a regime with multiple competing security branches is
harder to overthrow because no single commander has enough force.
Pilster-Böhmelt measure this with a Herfindahl-Hirschman-style index
of "effective number of forces." Higher = more fragmented = harder
to coup. The model uses it as `coup_proof_fragmentation`.

**Data-analytics concepts shown here.**

- *Harvard Dataverse as a data repository*. Academic datasets often
  live on Dataverse, a Harvard-run repository with stable DOIs.
  `download.file()` with the Dataverse API URL fetches the latest
  version programmatically.
- *Effective-number / Herfindahl-style indices*. These are diversity/
  concentration measures from economics (the Herfindahl-Hirschman
  Index for market concentration). Same math, different application
  — political-science adapts the technique to measure military
  fragmentation.
- *ID system conversion is universal*. P-B uses Correlates-of-War
  (`ccode`) numeric codes. We convert to `iso3c` via `countrycode`.
  Same pattern as every other loader.

**Domain terms used here.**

- **Pilster-Böhmelt** — Ulrich Pilster and Tobias Böhmelt, political
  scientists who built the counterbalancing dataset.
- **Counterbalancing** — strategy of fragmenting the military into
  multiple competing branches to prevent any one commander from
  successfully launching a coup.
- **Effective number of forces** — Herfindahl-style index of military
  fragmentation. 1 = one unified military, 5 = five equally-sized
  branches.
- **Herfindahl-Hirschman Index (HHI)** — economic measure of market
  concentration adapted here for military structure.

**What success looks like.** The `pb_counterbalancing` table has one
row per (country, year). Used in the model as
`coup_proof_fragmentation` — one of the consistently nonzero
coefficients in the violence-tier model.
:::

Downloads the Pilster-Böhmelt "Counterbalancing / Coup-Proofing Data" from
Harvard Dataverse and lands it as the `pb_counterbalancing` table.

The headline variable is **`effectivenumber`** — an effective-number-of-forces
fragmentation index (HHI-style). Higher = more fragmented security
architecture = harder to coordinate a coup (the "counterbalancing" intuition
from coup-proofing literature).

Coverage: 1970–2025, ~200 countries. Identified by Correlates-of-War code
(`ccode`); we convert to `iso3c` via `countrycode` so the join through the
master `countries` table works.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(countrycode)
source("_helpers.R")
```

## Download

Harvard Dataverse exposes a stable REST API. File ID 10980318 is the current
version of `Coup-Proofing 1970-2025.tab`. We persist to `data/raw/` so the
parser step is reproducible without re-hitting the network.

```{r}
#| label: fetch
dir.create("../data/raw", showWarnings = FALSE, recursive = TRUE)
pb_local <- "../data/raw/pilster_bohmelt_coup_proofing.tab"

if (!file.exists(pb_local) || file.info(pb_local)$size < 1000) {
  download.file(
    "https://dataverse.harvard.edu/api/access/datafile/10980318",
    pb_local, mode = "wb", quiet = TRUE
  )
}
cat("File:", pb_local, "(", file.info(pb_local)$size, "bytes)\n")
```

## Parse and crosswalk to ISO3

```{r}
#| label: parse
pb <- read.delim(pb_local, sep = "\t", stringsAsFactors = FALSE)
cat("Raw rows:", nrow(pb), " | Year range:", range(pb$year, na.rm = TRUE), "\n")

# Convert COW code to ISO3 so we can join to `countries`.
pb$iso3c <- cow_to_iso3(pb$ccode)
n_lost <- sum(is.na(pb$iso3c))
cat("Rows without ISO3 (dropped):", n_lost, "\n")

pb <- pb[!is.na(pb$iso3c), c("iso3c", "year", "abbrev", "ccode", "effectivenumber")]
head(pb, 3)
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "pb_counterbalancing", pb, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows FROM pb_counterbalancing")
```

## Coverage

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT COUNT(DISTINCT iso3c) AS countries,
         MIN(year) AS earliest, MAX(year) AS latest,
         ROUND(AVG(effectivenumber), 2) AS avg_effective_n,
         ROUND(STDDEV(effectivenumber), 2) AS sd_effective_n
  FROM pb_counterbalancing
  WHERE effectivenumber IS NOT NULL
")
```

## Cuba sanity check

```{r}
#| label: cuba-check
dbGetQuery(con, "
  SELECT year, effectivenumber
  FROM pb_counterbalancing
  WHERE iso3c = 'CUB' AND year >= 2015
  ORDER BY year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
