---
title: "07 — Load de Bruin State Security Forces data"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** Erica de Bruin's State Security Forces dataset
measures the *internal* composition of security architectures across
countries — counterbalancing and politicization indices, finer-grained
than Pilster-Böhmelt's effective-number measure. Sounds great. But
**coverage ends in 2010** and the model needs predictors through the
current year. So this dataset is loaded for analysis but **excluded
from the fit** for sparsity.

**Data-analytics concepts shown here.**

- *Recognizing when a dataset is too sparse to use*. A perfect feature
  with 15-year-stale data is worse than a noisy feature with current
  data. We could forward-fill from 2010 to 2025, but that fills 15
  years of synthetic values that the model would learn from. Forward-
  filling that long is dishonest.
- *Documenting the road not taken*. This loader stays in the pipeline
  with a clear note about why we're not using it. If someone wants to
  redesign around historical analysis, the loader is ready. **Code
  isn't just executable — it's a written record of analytic choices.**
- *Industry parallel*: same pattern when you have an old proprietary
  data source that's frozen. Either commit to using it for historical
  analysis only, or replace it.

**Domain terms used here.**

- **de Bruin** — Erica de Bruin, political scientist at Hamilton
  College. Author of *How to Prevent Coups d'État* (Cornell, 2020).
- **State security forces** — military, police, paramilitary,
  intelligence services; the apparatus of state violence.
- **Counterbalancing** (de Bruin's version) — finer-grained than
  Pilster-Böhmelt; measures how many parallel security structures
  exist that could check each other.
- **Politicization** — extent to which security forces are co-opted
  for political (rather than military) purposes.

**What success looks like.** The `security_forces` table loads
cleanly. Flagged in the model card as **excluded from v0.4+ fit**
because coverage ends 2010. If de Bruin updates the dataset, this
loader picks up the new data automatically.
:::

Downloads Erica de Bruin's State Security Forces dataset (the country-year
file from her 2021 JPR article) and lands it as the `security_forces` table.

Coverage: **1960–2010**, ~110 countries with full coverage and another ~90
with partial. Two model-relevant columns:

- **`counterbalancing`** — index of how many security forces structurally
  counter each other (a more detailed measure than Pilster-Böhmelt's
  effective-number)
- **`politicization`** — index of how many forces have political
  (regime-aligned) appointment / control

de Bruin ends at 2010. The panel build (08) will forward-fill each country's
last-available value through 2024 — coup-proofing structure changes slowly.

## Libraries

```{r}
#| label: setup
library(readxl)
library(DBI)
library(duckdb)
library(countrycode)
source("_helpers.R")
```

## Download

The replication files are hosted in a Dropbox shared folder. The `?dl=1`
parameter forces the folder to download as a zip.

```{r}
#| label: fetch
raw_dir <- "../data/raw"
dir.create(raw_dir, showWarnings = FALSE, recursive = TRUE)
zip_path <- file.path(raw_dir, "debruin_ssf.zip")
extract_dir <- file.path(raw_dir, "debruin_ssf")
xls_path <- file.path(extract_dir, "SSFD_cy_jul2020.xls")

if (!file.exists(xls_path)) {
  download.file(
    "https://www.dropbox.com/sh/sqilrelbaqa6qpm/AADk3FUPsUapW0sDbHByXmoZa?dl=1",
    zip_path, mode = "wb", quiet = TRUE
  )
  unzip(zip_path, exdir = extract_dir)
}
cat("Country-year file:", xls_path, "\n")
```

## Parse and crosswalk to ISO3

```{r}
#| label: parse
db_raw <- read_excel(xls_path)
cat("Raw rows:", nrow(db_raw), " | Year range:", range(db_raw$year, na.rm = TRUE), "\n")

# Convert COW to ISO3; keep only the columns we want.
db_raw$iso3c <- cow_to_iso3(db_raw$ccode)
n_lost <- sum(is.na(db_raw$iso3c))
cat("Rows without ISO3 (dropped):", n_lost, "\n")

security_forces <- as.data.frame(db_raw)[, c("iso3c", "year", "ccode",
                                              "counterbalancing", "politicization")]
security_forces <- security_forces[!is.na(security_forces$iso3c), ]
security_forces$year <- as.integer(security_forces$year)
head(security_forces, 3)
```

## Write to DuckDB

```{r}
#| label: write
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
dbWriteTable(con, "security_forces", security_forces, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows FROM security_forces")
```

## Coverage

```{r}
#| label: inspect
dbGetQuery(con, "
  SELECT COUNT(DISTINCT iso3c) AS countries,
         MIN(year) AS earliest, MAX(year) AS latest,
         ROUND(AVG(counterbalancing), 2) AS avg_counterbalancing,
         ROUND(AVG(politicization), 2) AS avg_politicization
  FROM security_forces
  WHERE counterbalancing IS NOT NULL
")
```

## Cuba sanity check

```{r}
#| label: cuba-check
dbGetQuery(con, "
  SELECT year, counterbalancing, politicization
  FROM security_forces
  WHERE iso3c = 'CUB' AND year >= 2000
  ORDER BY year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
