---
title: "08 — Build the country-year panel"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Joins V-Dem, WDI, UNHCR, and FAO through the `countries` master table into a
single country-year panel. One row per (iso3c, year). All numeric features
are z-scored **within country** so the model learns "this country's shock
relative to its own history."

Output: `panel_features` table in `data/panel.duckdb`.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(ggplot2)
source("_helpers.R")
```

## Open the panel db

```{r}
#| label: connect
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
```

## Aggregate FAO to annual

FAO is monthly and global. We average to annual so it lines up with the
country-year grain. (Year-over-year change would be richer; keeping it simple
for v0.)

```{r}
#| label: fao-annual
fao_annual <- dbGetQuery(con, "
  SELECT year, AVG(ffpi) AS ffpi_avg
  FROM fao
  GROUP BY year
  ORDER BY year
")
dbWriteTable(con, "fao_annual", fao_annual, overwrite = TRUE)
head(fao_annual, 3)
```

## Build the raw country-year join

Pull all V-Dem core variables we care about, left-join WDI on (iso3c, year),
UNHCR on (iso3c, year), and FAO on year alone (it's global).

```{r}
#| label: raw-join
raw <- dbGetQuery(con, "
  SELECT
    c.iso3c,
    c.name,
    c.region,
    c.subregion,
    v.year,
    v.v2x_regime         AS regime_type,
    v.v2x_civlib         AS civil_liberties,
    v.v2x_freexp_altinf  AS freedom_expression,
    v.v2regsuploc        AS regime_support_locus,
    v.v2cagenmob_ord     AS mass_mobilization_ord,
    v.v2caautmob_ord     AS autoc_mobilization_ord,
    v.v2cademmob_ord     AS demoz_mobilization_ord,
    v.v2caviol_ord       AS political_violence_ord,
    -- CPI priority chain: WDI -> hand-curated IMF -> bulk IMF WEO. Anything
    -- still NULL falls to region-median imputation in R below, and gets
    -- flagged via cpi_imputed.
    COALESCE(w.cpi_inflation, imf.cpi_inflation, weo.cpi_inflation) AS cpi_inflation,
    -- Tier-3 indicator: 1 when no real CPI source exists (will be region-median
    -- imputed below). Lets the model learn that 'imputed-CPI countries' are a
    -- distinct subpopulation rather than treating the synthetic value as signal.
    CASE WHEN w.cpi_inflation IS NULL
           AND imf.cpi_inflation IS NULL
           AND weo.cpi_inflation IS NULL
         THEN 1 ELSE 0 END AS cpi_imputed,
    w.gdp_growth,
    w.gdp_per_cap,
    w.trade_pct_gdp,
    w.unemployment,
    (COALESCE(u.refugees, 0) + COALESCE(u.asylum_seekers, 0)) AS refugee_outflow,
    f.ffpi_avg,
    pb.effectivenumber   AS coup_proof_fragmentation,
    sf.counterbalancing,
    sf.politicization,
    fh.hr_score          AS fariss_hr_score,
    pol.polity2          AS polity2
  FROM countries c
  -- Cap at 2024: WDI data for most countries ends here. V-Dem v16 has
  -- partial 2025 stub data with NAs in WDI/UNHCR/etc., which gets filled
  -- by region-median imputation. The snapshot would then pick 2025 (mostly
  -- imputed) over 2024 (mostly real). Filter to keep 2024 as the snapshot
  -- year for now. Revisit when next year's data releases land.
  JOIN vdem v       ON c.vdem_id    = v.country_id  AND v.year >= 1995 AND v.year <= 2024
  LEFT JOIN wdi w   ON c.iso3c      = w.iso3c       AND v.year = w.year
  LEFT JOIN unhcr u ON c.iso3c      = u.coo_iso     AND v.year = u.year
  LEFT JOIN fao_annual f         ON v.year  = f.year
  LEFT JOIN pb_counterbalancing pb ON c.iso3c = pb.iso3c AND v.year = pb.year
  LEFT JOIN security_forces sf   ON c.iso3c = sf.iso3c AND v.year = sf.year
  LEFT JOIN fariss_hr fh         ON c.iso3c = fh.iso3c AND v.year = fh.year
  LEFT JOIN polity5 pol          ON c.iso3c = pol.iso3c AND v.year = pol.year
  LEFT JOIN imf_inflation imf    ON c.iso3c = imf.iso3c AND v.year = imf.year
  LEFT JOIN imf_weo weo          ON c.iso3c = weo.iso3c AND v.year = weo.year
  WHERE c.region IS NOT NULL
  ORDER BY c.iso3c, v.year
")

# Forward-fill late-ending series with the country's last non-NA value:
# de Bruin ends 2010, Fariss ends 2021, Polity5 ends 2018. Structural
# variables change slowly so short carry-forward is defensible. For the two
# features that now enter the v0.2 ridge fit (fariss_hr_score, polity2) we
# cap carry-forward at +5 years and let downstream region-median imputation
# fill anything past that — stale governance scores get the most wrong on
# countries that just had a regime change (Myanmar 2021, Niger 2023, etc).
locf_within <- function(x) {
  for (i in seq_along(x)) if (is.na(x[i]) && i > 1) x[i] <- x[i - 1]
  x
}
locf_within_capped <- function(x, max_carry = 5) {
  last_real <- NA_real_
  carry <- 0L
  for (i in seq_along(x)) {
    if (!is.na(x[i])) {
      last_real <- x[i]
      carry <- 0L
    } else {
      carry <- carry + 1L
      if (!is.na(last_real) && carry <= max_carry) x[i] <- last_real
    }
  }
  x
}
# de Bruin: still excluded from the fit, keep unconditional fill so it stays
# usable for ad-hoc panel queries.
for (col in c("counterbalancing", "politicization")) {
  raw[[col]] <- ave(raw[[col]], raw$iso3c, FUN = locf_within)
}
# Fariss + Polity5: capped fill so 6+ year carries fall through to imputation.
for (col in c("fariss_hr_score", "polity2")) {
  raw[[col]] <- ave(raw[[col]], raw$iso3c, FUN = locf_within_capped)
}
cat("After forward-fill: non-NA rows per feature:\n")
print(c(counterbalancing  = sum(!is.na(raw$counterbalancing)),
        fariss_hr_score   = sum(!is.na(raw$fariss_hr_score)),
        polity2           = sum(!is.na(raw$polity2)),
        total_rows        = nrow(raw)))
cat("Raw panel rows:", nrow(raw), "\n")
cat("Countries:     ", length(unique(raw$iso3c)), "\n")
cat("Year range:    ", range(raw$year), "\n")
head(raw, 2)
```

## Within-country z-scoring

For each numeric feature, subtract the country's own historical mean and
divide by its own historical SD. This makes "Cuba's GDP shock in 2020"
comparable to "Argentina's GDP shock in 2002" even though their absolute
GDP-growth distributions are different. Skip countries with too-few obs.

```{r}
#| label: z-score
features <- c("civil_liberties", "freedom_expression", "regime_support_locus",
              "political_violence_ord",
              "cpi_inflation", "gdp_growth", "gdp_per_cap",
              "trade_pct_gdp", "unemployment",
              "refugee_outflow", "ffpi_avg",
              "coup_proof_fragmentation", "counterbalancing", "politicization",
              "fariss_hr_score", "polity2")

# z_within is defined in _helpers.R (sourced above). Same function used
# downstream by 12_export_artifacts.qmd if needed, and the dashboard's
# `effectiveZValue` JS function mirrors the formula + winsorize bound.

# Region-median imputation: for each (year, region), fill missing values
# with the median of available values in that (year, region). Falls back to
# global-median if the (year, region) cell is entirely empty. This keeps
# countries like Cuba (which doesn't report CPI to the World Bank) in the
# panel by giving them a plausible regional baseline rather than dropping
# them entirely. Done BEFORE z-scoring so within-country z's reflect a
# consistent imputed time series.
panel <- raw
imputation_log <- data.frame(feature = character(), n_imputed = integer())

for (feat in features) {
  na_idx <- which(is.na(panel[[feat]]))
  if (length(na_idx) == 0) next

  # Try region-median per (year, region)
  for (i in na_idx) {
    yr <- panel$year[i]
    rgn <- panel$region[i]
    pool <- panel[[feat]][panel$year == yr & panel$region == rgn]
    pool <- pool[!is.na(pool)]
    if (length(pool) >= 3) {
      panel[[feat]][i] <- median(pool)
    }
  }

  # Anything still NA: fall back to global median for that feature
  still_na <- is.na(panel[[feat]])
  if (any(still_na)) {
    panel[[feat]][still_na] <- median(panel[[feat]], na.rm = TRUE)
  }

  imputation_log <- rbind(imputation_log,
                          data.frame(feature = feat, n_imputed = length(na_idx)))
}
cat("Region-median imputation log:\n")
print(imputation_log)

# Some features are heavily skewed (e.g. refugee_outflow: most countries
# near zero, a few in the hundreds of thousands). Log-transform before
# z-scoring so the z reflects meaningful variation across the whole range.
log_features <- c("refugee_outflow")

for (feat in features) {
  z_col <- paste0("z_", feat)
  x <- panel[[feat]]
  if (feat %in% log_features) x <- log1p(pmax(x, 0))
  panel[[z_col]] <- ave(x, panel$iso3c, FUN = z_within)
}

# Quick sanity: a country's z-scores should average ~0 across history.
cuba <- subset(panel, iso3c == "CUB")
cat("Cuba z-score means (should be ~0):\n")
print(round(sapply(cuba[, paste0("z_", features)], mean, na.rm = TRUE), 3))
```

## Write the panel

```{r}
#| label: write
dbWriteTable(con, "panel_features", panel, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows FROM panel_features")
```

## Panel coverage

```{r}
#| label: coverage
dbGetQuery(con, "
  SELECT
    year,
    COUNT(*) AS rows,
    SUM(CASE WHEN regime_type IS NOT NULL THEN 1 ELSE 0 END) AS regime_obs,
    SUM(CASE WHEN gdp_growth  IS NOT NULL THEN 1 ELSE 0 END) AS gdp_obs,
    SUM(CASE WHEN mass_mobilization_ord IS NOT NULL THEN 1 ELSE 0 END) AS mob_obs
  FROM panel_features
  WHERE year >= 2018
  GROUP BY year
  ORDER BY year
")
```

## Sanity plot: Cuba indicators 2015 onward

```{r}
#| label: cuba-plot
#| fig-cap: "Cuba indicator z-scores, 2015 onward — built panel coherence check"
#| fig-width: 8
#| fig-height: 5
cuba_plot <- dbGetQuery(con, "
  SELECT year,
         z_civil_liberties   AS civil_liberties,
         z_gdp_growth        AS gdp_growth,
         z_refugee_outflow   AS refugee_outflow,
         z_ffpi_avg          AS food_price
  FROM panel_features
  WHERE iso3c = 'CUB' AND year >= 2015
  ORDER BY year
")

cuba_long <- stack(cuba_plot[, -1])
cuba_long$year <- rep(cuba_plot$year, ncol(cuba_plot) - 1)
names(cuba_long) <- c("z", "indicator", "year")

ggplot(cuba_long, aes(x = year, y = z, color = indicator)) +
  geom_hline(yintercept = 0, color = "#A89880", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(cuba_long$year)) +
  scale_color_manual(values = c(civil_liberties = srs_palette$accent,
                                gdp_growth      = "#5C7A93",
                                refugee_outflow = srs_palette$crisis,
                                food_price      = "#8B6F2B")) +
  labs(x = NULL, y = "z-score (within-country)",
       caption = "Source: V-Dem + WDI + UNHCR + FAO via panel.duckdb",
       color = NULL) +
  srs_theme() +
  ggplot2::theme(legend.position = "bottom")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
