---
title: "08 — Build the country-year panel"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Joins V-Dem, WDI, UNHCR, and FAO through the `countries` master table into a
single country-year panel. One row per (iso3c, year). All numeric features
are z-scored **within country** so the model learns "this country's shock
relative to its own history."

Output: `panel_features` table in `data/panel.duckdb`.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(ggplot2)
```

## Open the panel db

```{r}
#| label: connect
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
```

## Aggregate FAO to annual

FAO is monthly and global. We average to annual so it lines up with the
country-year grain. (Year-over-year change would be richer; keeping it simple
for v0.)

```{r}
#| label: fao-annual
fao_annual <- dbGetQuery(con, "
  SELECT year, AVG(ffpi) AS ffpi_avg
  FROM fao
  GROUP BY year
  ORDER BY year
")
dbWriteTable(con, "fao_annual", fao_annual, overwrite = TRUE)
head(fao_annual, 3)
```

## Build the raw country-year join

Pull all V-Dem core variables we care about, left-join WDI on (iso3c, year),
UNHCR on (iso3c, year), and FAO on year alone (it's global).

```{r}
#| label: raw-join
raw <- dbGetQuery(con, "
  SELECT
    c.iso3c,
    c.name,
    c.region,
    c.subregion,
    v.year,
    v.v2x_regime         AS regime_type,
    v.v2x_civlib         AS civil_liberties,
    v.v2x_freexp_altinf  AS freedom_expression,
    v.v2regsuploc        AS regime_support_locus,
    v.v2cagenmob_ord     AS mass_mobilization_ord,
    v.v2caautmob_ord     AS autoc_mobilization_ord,
    v.v2cademmob_ord     AS demoz_mobilization_ord,
    v.v2caviol_ord       AS political_violence_ord,
    w.cpi_inflation,
    w.gdp_growth,
    w.gdp_per_cap,
    w.trade_pct_gdp,
    w.unemployment,
    (COALESCE(u.refugees, 0) + COALESCE(u.asylum_seekers, 0)) AS refugee_outflow,
    f.ffpi_avg,
    pb.effectivenumber   AS coup_proof_fragmentation,
    sf.counterbalancing,
    sf.politicization
  FROM countries c
  JOIN vdem v       ON c.vdem_id    = v.country_id  AND v.year >= 1995
  LEFT JOIN wdi w   ON c.iso3c      = w.iso3c       AND v.year = w.year
  LEFT JOIN unhcr u ON c.iso3c      = u.coo_iso     AND v.year = u.year
  LEFT JOIN fao_annual f         ON v.year  = f.year
  LEFT JOIN pb_counterbalancing pb ON c.iso3c = pb.iso3c AND v.year = pb.year
  LEFT JOIN security_forces sf   ON c.iso3c = sf.iso3c AND v.year = sf.year
  WHERE c.region IS NOT NULL
  ORDER BY c.iso3c, v.year
")

# de Bruin ends at 2010; forward-fill each country's last non-NA value
# through 2024. Coup-proofing structure changes slowly so this is defensible.
locf_within <- function(x) {
  for (i in seq_along(x)) if (is.na(x[i]) && i > 1) x[i] <- x[i - 1]
  x
}
for (col in c("counterbalancing", "politicization")) {
  raw[[col]] <- ave(raw[[col]], raw$iso3c, FUN = locf_within)
}
cat("After forward-fill: counterbalancing non-NA rows:",
    sum(!is.na(raw$counterbalancing)), "/", nrow(raw), "\n")
cat("Raw panel rows:", nrow(raw), "\n")
cat("Countries:     ", length(unique(raw$iso3c)), "\n")
cat("Year range:    ", range(raw$year), "\n")
head(raw, 2)
```

## Within-country z-scoring

For each numeric feature, subtract the country's own historical mean and
divide by its own historical SD. This makes "Cuba's GDP shock in 2020"
comparable to "Argentina's GDP shock in 2002" even though their absolute
GDP-growth distributions are different. Skip countries with too-few obs.

```{r}
#| label: z-score
features <- c("civil_liberties", "freedom_expression", "regime_support_locus",
              "political_violence_ord",
              "cpi_inflation", "gdp_growth", "gdp_per_cap",
              "trade_pct_gdp", "unemployment",
              "refugee_outflow", "ffpi_avg",
              "coup_proof_fragmentation", "counterbalancing", "politicization")

z_within <- function(x) {
  if (sum(!is.na(x)) < 5) return(rep(NA_real_, length(x)))
  m <- mean(x, na.rm = TRUE); s <- sd(x, na.rm = TRUE)
  if (is.na(s) || s == 0) return(rep(0, length(x)))
  # Winsorize at +/-3 SDs so a country with a tiny historical baseline
  # (e.g. UK's near-zero refugee outflow) doesn't blow up to +3.5 when a
  # small absolute number arrives. Caps extreme outliers without changing
  # ordering for the bulk of the distribution.
  z <- (x - m) / s
  pmin(pmax(z, -3), 3)
}

# Some features are heavily skewed (e.g. refugee_outflow: most countries
# near zero, a few in the hundreds of thousands). Log-transform before
# z-scoring so the z reflects meaningful variation across the whole range.
log_features <- c("refugee_outflow")

panel <- raw
for (feat in features) {
  z_col <- paste0("z_", feat)
  x <- panel[[feat]]
  if (feat %in% log_features) x <- log1p(pmax(x, 0))
  panel[[z_col]] <- ave(x, panel$iso3c, FUN = z_within)
}

# Quick sanity: a country's z-scores should average ~0 across history.
cuba <- subset(panel, iso3c == "CUB")
cat("Cuba z-score means (should be ~0):\n")
print(round(sapply(cuba[, paste0("z_", features)], mean, na.rm = TRUE), 3))
```

## Write the panel

```{r}
#| label: write
dbWriteTable(con, "panel_features", panel, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows FROM panel_features")
```

## Panel coverage

```{r}
#| label: coverage
dbGetQuery(con, "
  SELECT
    year,
    COUNT(*) AS rows,
    SUM(CASE WHEN regime_type IS NOT NULL THEN 1 ELSE 0 END) AS regime_obs,
    SUM(CASE WHEN gdp_growth  IS NOT NULL THEN 1 ELSE 0 END) AS gdp_obs,
    SUM(CASE WHEN mass_mobilization_ord IS NOT NULL THEN 1 ELSE 0 END) AS mob_obs
  FROM panel_features
  WHERE year >= 2018
  GROUP BY year
  ORDER BY year
")
```

## Sanity plot: Cuba indicators 2015 onward

```{r}
#| label: cuba-plot
#| fig-cap: "Cuba indicator z-scores, 2015 onward — built panel coherence check"
#| fig-width: 8
#| fig-height: 5
parchment <- "#F8F3E8"
ink       <- "#2A2620"
accent    <- "#1F3A60"
crisis    <- "#8B2E2E"

cuba_plot <- dbGetQuery(con, "
  SELECT year,
         z_civil_liberties   AS civil_liberties,
         z_gdp_growth        AS gdp_growth,
         z_refugee_outflow   AS refugee_outflow,
         z_ffpi_avg          AS food_price
  FROM panel_features
  WHERE iso3c = 'CUB' AND year >= 2015
  ORDER BY year
")

cuba_long <- stack(cuba_plot[, -1])
cuba_long$year <- rep(cuba_plot$year, ncol(cuba_plot) - 1)
names(cuba_long) <- c("z", "indicator", "year")

ggplot(cuba_long, aes(x = year, y = z, color = indicator)) +
  geom_hline(yintercept = 0, color = "#A89880", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(cuba_long$year)) +
  scale_color_manual(values = c(civil_liberties = accent,
                                gdp_growth      = "#5C7A93",
                                refugee_outflow = crisis,
                                food_price      = "#8B6F2B")) +
  labs(x = NULL, y = "z-score (within-country)",
       caption = "Source: V-Dem + WDI + UNHCR + FAO via panel.duckdb",
       color = NULL) +
  theme_minimal(base_family = "serif", base_size = 12) +
  theme(plot.background  = element_rect(fill = parchment, color = NA),
        panel.background = element_rect(fill = parchment, color = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "#D8CFC0"),
        axis.text        = element_text(color = ink),
        axis.title       = element_text(color = ink),
        legend.position  = "bottom",
        plot.caption     = element_text(color = "#6B5D4F", face = "italic"))
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
