---
title: "08 — Build the country-year panel"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Joins V-Dem, WDI, UNHCR, and FAO through the `countries` master table into a
single country-year panel. One row per (iso3c, year).

**v0.4 simplification:** features are **raw cross-country levels**, no
within-country z-scoring. The model answers "given the structural and
economic characteristics of this country, what's the world-historical
unrest probability?" — the PITF/Goldstone framing.

**v0.4.1 honesty principle:** model-input features are **not** region-median
imputed. If a country (e.g., PRK, Cuba pre-2018, Somalia) has no real source
data for a feature, the row stays NA and the country is dropped from
training and from the inference snapshot. The dashboard surfaces this as
"no prediction available" rather than inventing a regional baseline.

Heavy-tailed features (`refugee_outflow`, `cpi_inflation`) get
log-transformed or winsorized so the regression isn't dominated by
Venezuela/Zimbabwe hyperinflation or Afghanistan/Syria refugee outflows.

Output: `panel_features` table in `data/panel.duckdb`.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(ggplot2)
source("_helpers.R")  # srs_palette / srs_theme — z_within() is gone in v0.4
```

## Open the panel db

```{r}
#| label: connect
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
```

## Aggregate FAO to annual

FAO is monthly and global. We average to annual so it lines up with the
country-year grain. (Year-over-year change would be richer; keeping it simple
for v0.)

```{r}
#| label: fao-annual
fao_annual <- dbGetQuery(con, "
  SELECT year, AVG(ffpi) AS ffpi_avg
  FROM fao
  GROUP BY year
  ORDER BY year
")
dbWriteTable(con, "fao_annual", fao_annual, overwrite = TRUE)
head(fao_annual, 3)
```

## Build the raw country-year join

Pull all V-Dem core variables we care about, left-join WDI on (iso3c, year),
UNHCR on (iso3c, year), and FAO on year alone (it's global).

```{r}
#| label: raw-join
raw <- dbGetQuery(con, "
  SELECT
    c.iso3c,
    c.name,
    c.region,
    c.subregion,
    v.year,
    v.v2x_regime         AS regime_type,
    v.v2x_civlib         AS civil_liberties,
    v.v2x_freexp_altinf  AS freedom_expression,
    v.v2regsuploc        AS regime_support_locus,
    v.v2cagenmob_ord     AS mass_mobilization_ord,
    v.v2caautmob_ord     AS autoc_mobilization_ord,
    v.v2cademmob_ord     AS demoz_mobilization_ord,
    v.v2caviol_ord       AS political_violence_ord,
    v.v2elaccept_ord     AS loser_consent,     -- 0..4, V-Dem election-losers consent (factionalism inverse)
    -- CPI priority chain: WDI -> hand-curated IMF -> bulk IMF WEO. Anything
    -- still NULL falls to region-median imputation in R below, and gets
    -- flagged via cpi_imputed.
    COALESCE(w.cpi_inflation, imf.cpi_inflation, weo.cpi_inflation) AS cpi_inflation,
    -- Tier-3 indicator: 1 when no real CPI source exists (will be region-median
    -- imputed below). Lets the model learn that 'imputed-CPI countries' are a
    -- distinct subpopulation rather than treating the synthetic value as signal.
    CASE WHEN w.cpi_inflation IS NULL
           AND imf.cpi_inflation IS NULL
           AND weo.cpi_inflation IS NULL
         THEN 1 ELSE 0 END AS cpi_imputed,
    -- GDP growth + GDP per capita: backfill WDI with IMF WEO. WDI runs
    -- 12–18 months behind real time; WEO has IMF country-team estimates
    -- for the current year and projections through year+6. For 2025/2026
    -- forecasts this is the only way to score countries before WDI's
    -- annual update lands. Marked here as a source-backfill (not synthetic
    -- imputation): both WDI and WEO are real published data.
    COALESCE(w.gdp_growth, weo_g.gdp_growth) AS gdp_growth,
    COALESCE(w.gdp_per_cap, weo_pc.gdp_per_cap) AS gdp_per_cap,
    w.trade_pct_gdp,
    -- Unemployment: WDI ILO modelled estimate by default, but for state-
    -- managed economies (Cuba, Belarus, PRK, Venezuela, etc.) the official
    -- rate doesn't reflect informal-sector / underemployment reality. We
    -- override with hand-curated IMF Article IV / NGO estimates where they
    -- exist; otherwise fall back to WDI.
    COALESCE(un.unemployment_curated, w.unemployment) AS unemployment,
    (COALESCE(u.refugees, 0)   + COALESCE(u.asylum_seekers, 0))   AS refugee_outflow,
    (COALESCE(uin.refugees, 0) + COALESCE(uin.asylum_seekers, 0)) AS refugee_inflow,
    f.ffpi_avg,
    pb.effectivenumber   AS coup_proof_fragmentation,
    sf.counterbalancing,
    sf.politicization,
    fh.hr_score          AS fariss_hr_score,
    pol.polity2          AS polity2
  FROM countries c
  -- v0.4.1: panel goes through 2025. WDI/V-Dem/UNHCR/PB all have 2025
  -- data; UCDP currently ends 2024 so conflict_neighbors_count for 2025
  -- will be NA for most countries, which means most 2025 rows fall back
  -- to 2024 at inference time. That's the honest answer — we no longer
  -- region-median fill phantom 2025 values. When UCDP 2025 lands (mid-2026
  -- release cadence), more countries will get a real 2025 prediction
  -- (which forecasts unrest in 2026).
  JOIN vdem v       ON c.vdem_id    = v.country_id  AND v.year >= 1995 AND v.year <= 2025
  LEFT JOIN wdi w   ON c.iso3c      = w.iso3c       AND v.year = w.year
  LEFT JOIN unhcr        u   ON c.iso3c = u.coo_iso   AND v.year = u.year
  LEFT JOIN unhcr_inflow uin ON c.iso3c = uin.coa_iso AND v.year = uin.year
  LEFT JOIN fao_annual f         ON v.year  = f.year
  LEFT JOIN pb_counterbalancing pb ON c.iso3c = pb.iso3c AND v.year = pb.year
  LEFT JOIN security_forces sf   ON c.iso3c = sf.iso3c AND v.year = sf.year
  LEFT JOIN fariss_hr fh         ON c.iso3c = fh.iso3c AND v.year = fh.year
  LEFT JOIN polity5 pol          ON c.iso3c = pol.iso3c AND v.year = pol.year
  LEFT JOIN imf_inflation imf      ON c.iso3c = imf.iso3c    AND v.year = imf.year
  LEFT JOIN imf_weo weo            ON c.iso3c = weo.iso3c    AND v.year = weo.year
  LEFT JOIN imf_weo_gdp_growth   weo_g  ON c.iso3c = weo_g.iso3c   AND v.year = weo_g.year
  LEFT JOIN imf_weo_gdp_pc       weo_pc ON c.iso3c = weo_pc.iso3c  AND v.year = weo_pc.year
  LEFT JOIN unemployment_curated un     ON c.iso3c = un.iso3c      AND v.year = un.year
  WHERE c.region IS NOT NULL
  ORDER BY c.iso3c, v.year
")

# v0.4.1 honesty principle: NO forward-fill for any source. de Bruin
# (ends 2010), Polity5 (ends 2018), and Fariss (ends 2021) leave their NAs
# intact. These features are excluded from the v0.4 model fit anyway; in
# panel queries / dashboard display they show as NA for post-source-end
# years, which is the honest answer. Earlier versions had locf carry-forward
# that created phantom "data" the model treated as real.
cat("After forward-fill: non-NA rows per feature:\n")
print(c(counterbalancing  = sum(!is.na(raw$counterbalancing)),
        fariss_hr_score   = sum(!is.na(raw$fariss_hr_score)),
        polity2           = sum(!is.na(raw$polity2)),
        total_rows        = nrow(raw)))
cat("Raw panel rows:", nrow(raw), "\n")
cat("Countries:     ", length(unique(raw$iso3c)), "\n")
cat("Year range:    ", range(raw$year), "\n")
head(raw, 2)
```

## v0.3 derived features

Three new "structural anchor" features grounded in the PITF (Goldstone 2010)
and ViEWS methodology:

- `log_gdp_per_cap` — development anchor. Rich countries are structurally
  less unrest-prone; this is the second-strongest predictor in PITF after
  regime type. Used as a level (not z-scored within country) so cross-country
  variation in wealth survives.
- `conflict_neighbors_count` — PITF's "conflict-ridden neighborhoods" feature.
  Count of bordering / same-region countries with UCDP fatalities ≥100 this
  year. Same-region approximation since we don't have a true borders table.
- `loser_consent` — V-Dem `v2elaccept_ord` (0..4). Inverse of factional
  politics: low values mean election losers don't accept results
  (Bolsonaro 2022, Trump 2020-2023). PITF found this to be the most powerful
  predictor among partial-democracy regimes.

```{r}
#| label: v03-features
# log_gdp_per_cap: log1p is safe when gdp_per_cap is 0 or NA.
raw$log_gdp_per_cap <- log1p(pmax(raw$gdp_per_cap, 0))

# conflict_neighbors_count: for each (region, year), how many OTHER countries
# in the same UN region had UCDP fatalities >= 100 that year?
ucdp_annual_local <- dbGetQuery(con, "
  SELECT iso3c, year, fatalities_best FROM ucdp_annual
  WHERE fatalities_best >= 100
")
# Tag each UCDP row with its region by joining to countries
ucdp_annual_local <- merge(
  ucdp_annual_local,
  unique(raw[, c("iso3c", "region")]),
  by = "iso3c", all.x = TRUE
)
# Count conflict countries per (region, year). Use UCDP year as-is — we'll
# join it as the LAG-1 feature so 2025 panel rows pick up 2024 UCDP data
# (academically defensible — conflict spillover takes time to manifest, and
# this also means 2025 features are forecastable today even though UCDP
# only releases a year behind real time).
region_conflict_counts <- aggregate(
  iso3c ~ region + year,
  data = ucdp_annual_local,
  FUN = function(x) length(unique(x))
)
names(region_conflict_counts)[3] <- "n_in_region"
# Lag: a row for panel year=t pulls UCDP from year=t-1
region_conflict_counts$panel_year <- region_conflict_counts$year + 1L
region_conflict_counts$year <- NULL

raw <- merge(raw, region_conflict_counts,
             by.x = c("region", "year"),
             by.y = c("region", "panel_year"),
             all.x = TRUE)
raw$n_in_region[is.na(raw$n_in_region)] <- 0
# Subtract 1 if THIS country itself was in conflict last year (don't count self)
this_country_conflict <- ucdp_annual_local[, c("iso3c", "year")]
this_country_conflict$panel_year <- this_country_conflict$year + 1L
this_country_conflict$year <- NULL
this_country_conflict$self_in_conflict <- 1L
raw <- merge(raw, this_country_conflict,
             by.x = c("iso3c", "year"),
             by.y = c("iso3c", "panel_year"), all.x = TRUE)
raw$self_in_conflict[is.na(raw$self_in_conflict)] <- 0L
raw$conflict_neighbors_count <- pmax(raw$n_in_region - raw$self_in_conflict, 0)
raw$n_in_region <- NULL
raw$self_in_conflict <- NULL

# loser_consent already in raw via SQL join (v2elaccept_ord).

cat("v0.3 anchor features:\n")
print(summary(raw[, c("log_gdp_per_cap", "conflict_neighbors_count", "loser_consent")]))

# Re-sort because the merges reorder rows.
raw <- raw[order(raw$iso3c, raw$year), ]
```

## Heavy-tail transforms (no imputation for v0.4 fit features)

**v0.4.1 honesty principle:** the v0.4 model-input features are **not**
region-median imputed. If a country has no real source data for a feature,
the row is NA and gets listwise-dropped from training; at inference, the
dashboard reports "no prediction available" rather than scoring a country
on synthetic regional medians. This is what protects PRK / Cuba-pre-2018 /
Somalia / Eritrea from getting a fake prediction built on imputed data.

We still region-median impute *ancillary* features (`fariss_hr_score`,
`polity2`, `counterbalancing`, `freedom_expression`, etc.) for dashboard
display purposes only — they aren't model inputs.

Two heavy-tailed transforms for cross-country comparability:

- `log_refugee_outflow = log1p(max(0, refugee_outflow − refugee_inflow))` —
  **net** outflow, not gross. Built from UNHCR origin- and destination-side
  counts. The v0 fit used gross outflow and learned a strong positive
  coefficient because Syria/Afghanistan dominate the high end. Applied to
  the USA, gross outflow (~4K asylum seekers) was being treated as a
  weak-but-real instability signal — and because the model has so few
  active features for established democracies, it dominated the prediction
  (~+3 on the logit, ~17% violence headline). Switching to net outflow
  zeros it out for receiving democracies (USA, Germany, Canada, UK) which
  is the desired semantics: refugee outflow is a *symptom* of instability,
  not a cause, and rich receiving countries clearly don't have that symptom.
- `cpi_inflation` is **clipped to [-10, 200]%** — hyperinflation outliers
  (Venezuela 2017, Zimbabwe 2008) would otherwise swing the coefficient.

```{r}
#| label: transform
# v0.4.1 honesty principle: NO imputation anywhere. Model-input features
# stay NA where source data is missing; rows get listwise-dropped in fitting
# and skipped at inference. Ancillary panel features also stay NA — earlier
# versions region-median imputed them, which created phantom values that
# could leak into dashboard displays or future regressions by accident.
panel <- raw

# Cross-country log/clip transforms for the heavy-tailed model inputs.
# (Applied AFTER joining; NA stays NA through these transforms.)
panel$net_refugee_outflow <- pmax(0, panel$refugee_outflow - panel$refugee_inflow)
panel$log_refugee_outflow <- log1p(panel$net_refugee_outflow)
panel$cpi_inflation       <- pmin(pmax(panel$cpi_inflation, -10), 200)

# v0.5: lag political_violence_ord by 1 year per country. Concurrent
# political_violence_ord (V-Dem v2caviol_ord) at year T predicted our
# year-T+1 outcome, which is partly DEFINED by V-Dem violence at year T+1
# (the multi-source fusion uses v2cagenmob_ord >= 4 as a fallback). Lagging
# breaks the autocorrelation: year-T-1 violence is in the past relative to
# year-T+1 outcome, two-year gap, no overlap with the outcome label.
panel <- panel[order(panel$iso3c, panel$year), ]
panel$political_violence_ord_lag1 <- ave(panel$political_violence_ord, panel$iso3c,
                                          FUN = function(x) c(NA, x[-length(x)]))

# v0.8: lag mass_mobilization_ord by 1 year. Concurrent mass_mobilization_ord
# at year T would leak into the outcome label (which uses V-Dem mob at T+1),
# but T-1 is a clean 2-year gap. Gives the model a "chronic mobilization
# potential" signal — without it, countries like Iran that sit at V-Dem 2-3
# every year look identical to countries with no protest history at all.
panel$mass_mobilization_ord_lag1 <- ave(panel$mass_mobilization_ord, panel$iso3c,
                                         FUN = function(x) c(NA, x[-length(x)]))

# v0.8 model-input feature set. Adds mass_mobilization_ord_lag1 to give the
# mobilization tier a chronic-protest-potential signal it was previously blind to.
features <- c("civil_liberties", "log_gdp_per_cap", "loser_consent",
              "conflict_neighbors_count",
              "gdp_growth", "cpi_inflation", "unemployment",
              "political_violence_ord_lag1",
              "mass_mobilization_ord_lag1",
              "log_refugee_outflow", "coup_proof_fragmentation")

cat("v0.4 features (raw levels, ", length(features), " total):\n", sep = "")
print(features)

# Per-feature NA count for 2024 — which countries get dropped at inference time?
y2024 <- panel[panel$year == 2024, ]
cat("\n2024 NA counts per model-input feature (n=", nrow(y2024),
    " country rows):\n", sep = "")
print(sapply(features, function(f) sum(is.na(y2024[[f]]))))
cat("\nCountries with ANY NA on a v0.4 model feature in 2024:\n")
y2024_na_any <- y2024[!complete.cases(y2024[, features]), c("iso3c", "name")]
print(y2024_na_any)
cat("\nv0.4 feature summary across all years (cross-country distribution):\n")
print(summary(panel[, features]))
```

## Write the panel

```{r}
#| label: write
dbWriteTable(con, "panel_features", panel, overwrite = TRUE)
dbGetQuery(con, "SELECT COUNT(*) AS rows FROM panel_features")
```

## Panel coverage

```{r}
#| label: coverage
dbGetQuery(con, "
  SELECT
    year,
    COUNT(*) AS rows,
    SUM(CASE WHEN regime_type IS NOT NULL THEN 1 ELSE 0 END) AS regime_obs,
    SUM(CASE WHEN gdp_growth  IS NOT NULL THEN 1 ELSE 0 END) AS gdp_obs,
    SUM(CASE WHEN mass_mobilization_ord IS NOT NULL THEN 1 ELSE 0 END) AS mob_obs
  FROM panel_features
  WHERE year >= 2018
  GROUP BY year
  ORDER BY year
")
```

## Sanity plot: Cuba raw indicators 2015 onward

```{r}
#| label: cuba-plot
#| fig-cap: "Cuba indicators (raw cross-country levels), 2015 onward"
#| fig-width: 8
#| fig-height: 5
cuba_plot <- dbGetQuery(con, "
  SELECT year, civil_liberties, gdp_growth, log_refugee_outflow, cpi_inflation
  FROM panel_features
  WHERE iso3c = 'CUB' AND year >= 2015
  ORDER BY year
")

cuba_long <- stack(cuba_plot[, -1])
cuba_long$year <- rep(cuba_plot$year, ncol(cuba_plot) - 1)
names(cuba_long) <- c("value", "indicator", "year")

ggplot(cuba_long, aes(x = year, y = value, color = indicator)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ indicator, scales = "free_y") +
  scale_x_continuous(breaks = unique(cuba_long$year)) +
  scale_color_manual(values = c(civil_liberties     = srs_palette$accent,
                                gdp_growth          = "#5C7A93",
                                log_refugee_outflow = srs_palette$crisis,
                                cpi_inflation       = "#8B6F2B")) +
  labs(x = NULL, y = NULL,
       caption = "Source: V-Dem + WDI + UNHCR via panel.duckdb",
       color = NULL) +
  srs_theme() +
  ggplot2::theme(legend.position = "none")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
