---
title: "09 — Label outcome events (next-year mass mobilization)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

For each (country, year), look up V-Dem's mass-mobilization ordinal at
(country, year+1) and threshold to a binary outcome:

- `unrest_next_year = 1` if `v2cagenmob_ord >= 4` next year
- `unrest_next_year = 0` otherwise

V-Dem's `v2cagenmob_ord` ordinal scale (5 levels in v16):

| Value | Meaning | % of country-years |
|---|---|---|
| 0 | No mass mobilization | 18% |
| 1 | Few small events | 27% |
| 2 | Several small events | 13% |
| 3 | Several large events | 31% |
| 4 | Many large events | 11% |

The `>= 4` threshold captures "sustained mass mobilization at scale" —
matches the original spec resolution criteria (>100k cumulative participants,
multiple cities, >2 weeks). This is our v0 outcome proxy until ACLED comes
online with event-level frequency.

Output: `panel_labeled` table (same as `panel_features` + the `unrest_next_year`
column).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Construct the lead label

DuckDB's `LEAD()` window function pulls the next year's value within each
country's series. We then threshold.

```{r}
#| label: label
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

dbExecute(con, "
  CREATE OR REPLACE TABLE panel_labeled AS
  SELECT *,
         LEAD(mass_mobilization_ord) OVER (PARTITION BY iso3c ORDER BY year) AS mass_mob_next,
         CASE WHEN LEAD(mass_mobilization_ord) OVER (PARTITION BY iso3c ORDER BY year) >= 4
              THEN 1
              WHEN LEAD(mass_mobilization_ord) OVER (PARTITION BY iso3c ORDER BY year) IS NULL
              THEN NULL
              ELSE 0 END AS unrest_next_year
  FROM panel_features
")

dbGetQuery(con, "SELECT COUNT(*) AS rows FROM panel_labeled")
```

## Outcome distribution

```{r}
#| label: distribution
dbGetQuery(con, "
  SELECT unrest_next_year, COUNT(*) AS n,
         ROUND(100.0 * COUNT(*) / (SELECT COUNT(*) FROM panel_labeled WHERE unrest_next_year IS NOT NULL), 1) AS pct
  FROM panel_labeled
  WHERE unrest_next_year IS NOT NULL
  GROUP BY unrest_next_year
  ORDER BY unrest_next_year
")
```

## Outcome rate by regime category

V-Dem RoW: 0=closed autocracy, 1=electoral autocracy, 2=electoral democracy,
3=liberal democracy. We expect autocracies to show higher mobilization rates
(more repression triggers more visible protest).

```{r}
#| label: by-regime
dbGetQuery(con, "
  SELECT regime_type,
         COUNT(*) AS n_obs,
         SUM(unrest_next_year) AS n_unrest,
         ROUND(100.0 * AVG(unrest_next_year), 1) AS pct_unrest
  FROM panel_labeled
  WHERE unrest_next_year IS NOT NULL AND regime_type IS NOT NULL
  GROUP BY regime_type
  ORDER BY regime_type
")
```

## Outcome rate over time

```{r}
#| label: by-year
dbGetQuery(con, "
  SELECT year,
         COUNT(*) AS n_obs,
         SUM(unrest_next_year) AS n_unrest,
         ROUND(100.0 * AVG(unrest_next_year), 1) AS pct_unrest
  FROM panel_labeled
  WHERE unrest_next_year IS NOT NULL AND year BETWEEN 2018 AND 2023
  GROUP BY year
  ORDER BY year
")
```

## Cuba check

```{r}
#| label: cuba-check
dbGetQuery(con, "
  SELECT year, mass_mobilization_ord, mass_mob_next, unrest_next_year, regime_type
  FROM panel_labeled
  WHERE iso3c = 'CUB' AND year BETWEEN 2018 AND 2023
  ORDER BY year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
