---
title: "09 — Label outcome events (two-tier: mobilization + violence)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** `08_build_panel.qmd` produced the predictors.
This notebook produces the **outcomes** — the binary labels the model
will learn to predict. Outcomes are the most important and often
hardest part of a predictive analysis. Defining them carelessly is
where projects most often go wrong.

**Data-analytics concepts shown here.**

- *Outcome definition is a design decision*. There's no objective
  ground truth for "country had a mass mobilization episode" — you
  pick a definition and live with it. Different choices produce
  different models. We split into two tiers (mobilization, violence)
  because they're empirically different processes.
- *Multi-source OR-fusion*. No single source catches every event.
  MMP records large protests but ends in 2020. V-Dem has expert
  ordinals but updates slowly. ACLED has dense recent coverage but
  pre-2018 geographic gaps. We `OR-fuse` them: a country-year is
  labeled 1 if ANY source flags it. This is **data ensembling at the
  outcome level**, not just the model level.
- *Threshold choices are calibration choices*. We use ACLED violent-
  demonstration fatalities ≥ 10 (catches Iran 2022's 421 deaths,
  USA BLM's 14) and UCDP fatalities ≥ 100 (the canonical threshold
  for "serious armed conflict"). Lower or higher thresholds change
  the label distribution. Document and justify every threshold.
- *Bias-aware feature design*. Event counts have media-availability
  bias (open democracies get more news coverage so more events get
  coded). Fatalities are the bias-resistant currency. v0.7-rev0 used
  event counts and broke catastrophically; v0.7.1 uses fatalities only.
- *Future-leakage avoidance via lagged outcomes*. The label at panel
  row T predicts what happens at year T+1 (`LEAD(...) OVER ...` in
  SQL). The features stay at year T. Two-year gap between feature
  lags (T-1) and outcomes (T+1) keeps the prediction task honest.
- *NULL vs. 0 vs. 1*. "No data about next year" is different from
  "we know nothing happened." The CASE WHEN logic handles this
  explicitly: NULL when sources don't have data, 0 when all sources
  agree on no event, 1 when any source flags an event.

**Domain terms used here.**

- **Outcome label** — the variable we're trying to predict. In
  classification, it's a 0/1 flag per observation.
- **OR-fusion** — combining multiple sources with logical OR. Any
  source flagging "yes" → outcome = 1.
- **Future-leakage** — using information that wouldn't be available
  at prediction time. Cardinal sin in predictive modeling.
- **Lag-1** — using year T-1's value to predict year T+1's outcome.
- **`LEAD(...) OVER (PARTITION BY ... ORDER BY ...)`** — SQL window
  function that pulls the next row's value within a country.

**What success looks like.** The `panel_labeled` table adds three
columns to `panel_features`: `outcome_mobilization`, `outcome_violence`,
and `unrest_next_year` (legacy fused). Base rates are around 11-15%
for mobilization, 25-30% for violence — within the range where a
model can usefully discriminate. If the base rate were 1% or 99%,
the prediction task would be much harder.
:::

**v0.6:** stratify the outcome into two tiers. The previous single
`unrest_next_year` flag fused four channels into one — which made the
prediction ambiguous ("USA 26%" could mean coup OR protest OR armed
conflict; same number meant different things in different countries).
The two tiers separate **disruptive mobilization** from
**regime-threatening violence**.

**v0.7.1 (revised Phase A): ACLED uses FATALITIES, not event counts.**

The v0.7-rev0 attempt used an event-count threshold on ACLED demonstrations
(10 per million population). Diagnostic rendering showed that threshold
catastrophically over-fires on active democracies: Sweden, Japan, Germany,
Canada, France, USA all flipped to mobilization=1 every year 2019-2024
because ACLED counts every protest event — a 50-person climate rally
outside Swedish parliament is "1 event" the same way a million-person
march in Tehran is. Mobilization base rate in the 2023 holdout jumped from
11.8% to 63.2% and Brier skill score collapsed from −0.05 to −0.81. That
threshold rule has been reverted.

**Why fatalities, not events?** ACLED's event-data methodology has known
media-availability bias: open democracies get more journalism so individual
events get coded more granularly; closed regimes are under-coded. **Deaths
are the bias-resistant currency** — bodies are tracked across regimes
more uniformly than events (hospitals, families, opposition document
deaths even when state suppresses news reports). Plus, "mobilization that
turned deadly" is a sharper, more honest outcome than "mobilization
however measured": Iran 2022 (421 violent-demo fatalities), USA 2020 BLM
(14 fatalities) are unambiguous; Sweden 2024 (0) is unambiguous. France
2023 pension protests (0 fatalities) genuinely was a chronic-democracy
mobilization year, but it falls outside this *deadly mobilization* outcome
— captured by the V-Dem ord ≥ 4 channel if applicable.

ACLED in this revision provides one new OR-channel per tier:

| Tier | Outcome column | Channels (OR-fused) |
|---|---|---|
| 1 | `outcome_mobilization` | MMP per-capita large protest OR V-Dem `v2cagenmob_ord ≥ 4` OR **ACLED violent-demonstration fatalities ≥ 10** |
| 2 | `outcome_violence` | UCDP fatalities ≥ 100 OR Cline coup attempt OR **ACLED total fatalities ≥ 100** |

**Mobilization threshold rationale (≥ 10 violent-demo fatalities):**
Iran 2022 (421), Iran 2021 (10), USA 2020 BLM (14) → flagged. Cuba 2021
11J (1 fatality) → not flagged by this channel but caught by the existing
MMP-or-V-Dem channels. Sweden / Japan / Germany / chronic-protest
democracies → never flagged.

**Violence threshold rationale (≥ 100 fatalities):** Same threshold as
UCDP, so this adds a complementary measurement (different coding pipeline,
different release cadence) rather than a different threshold. Critically,
fills UCDP's release-lag gap (UCDP coverage ends 2024; ACLED runs through
2026-05).

**Pre-coverage country-years** (ACLED hadn't yet started coding the
country in question — Latin America / Europe / North America pre-2018)
have NULL ACLED counts. `COALESCE(..., 0)` pulls them to 0, meaning
"ACLED doesn't flag" — the OR-fusion falls back to MMP / V-Dem / UCDP /
Cline for those rows. Same pattern as the existing channels.

We keep `unrest_next_year` as a legacy column (OR of both tiers) for
backwards compatibility with anything that still depends on it.

**Source coverage:**

- Mobilization: V-Dem v16 runs through 2025 ✓; MMP ends 2020 (still usable for training)
- Violence: UCDP runs through 2024 ✓; Cline coups end 2022 (still usable for training)

Output: `panel_labeled` table (`panel_features` + 3 outcome columns).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Build the two-tier label

```{r}
#| label: label
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

dbExecute(con, "
  CREATE OR REPLACE TABLE panel_labeled AS
  WITH fused AS (
    SELECT
      p.*,
      LEAD(mass_mobilization_ord) OVER (PARTITION BY p.iso3c ORDER BY p.year) AS mass_mob_next,

      -- Per-source next-year flags. NULL = no data at all (post-source-end),
      -- 0 = source ran but recorded no event, >0 = event(s) recorded.
      COALESCE(c.n_realized, 0) + COALESCE(c.n_attempted, 0)   AS coup_flag,
      COALESCE(u.fatalities_best, 0)                            AS ucdp_fatal_next,
      COALESCE(m.n_large_protests, 0)                           AS mmp_large_next,
      LEAD(mass_mobilization_ord) OVER (PARTITION BY p.iso3c ORDER BY p.year) AS vdem_mob_next,

      -- v0.7.1: ACLED next-year channels (fatalities-only).
      --   acled_violent_demo_fat_next  = violent-demonstration FATALITIES at year+1
      --   acled_total_fat_next         = total fatalities (all event types) at year+1
      --   acled_covered_next           = 1 if ACLED was actually covering year+1
      --   acled_year_complete_next     = 1 if year+1 ended before the ACLED extract
      -- Event counts (acled_n_*) are deliberately NOT pulled here — they
      -- were the bias-prone signal that broke v0.7-rev0. We use fatalities
      -- only as the OR-channel trigger.
      COALESCE(a.fat_violent_demo, 0)              AS acled_violent_demo_fat_next,
      COALESCE(a.fat_total, 0)                     AS acled_total_fat_next,
      COALESCE(a.acled_covered, 0)                 AS acled_covered_next,
      COALESCE(a.acled_year_complete, 0)           AS acled_year_complete_next
    FROM panel_features p
    LEFT JOIN coups_annual c ON c.iso3c = p.iso3c AND c.year = p.year + 1
    LEFT JOIN ucdp_annual  u ON u.iso3c = p.iso3c AND u.year = p.year + 1
    LEFT JOIN mmp_annual   m ON m.iso3c = p.iso3c AND m.year = p.year + 1
    LEFT JOIN acled_annual a ON a.iso3c = p.iso3c AND a.year = p.year + 1
  )
  SELECT *,
    -- TIER 1: mass mobilization (protest/demonstration at country-defining scale)
    -- v0.7.1: third OR-channel = ACLED violent-demonstration FATALITIES ≥ 10.
    -- Fatalities-only by design — event counts have media-availability bias
    -- that broke v0.7-rev0 (Sweden, Japan, etc. all flipped to mobilization=1).
    CASE
      WHEN mmp_large_next >= 1
        OR (vdem_mob_next >= 4 AND year >= 2020)   -- V-Dem fallback for post-MMP years
        OR acled_violent_demo_fat_next >= 10        -- ACLED deadly-protest channel
      THEN 1
      WHEN vdem_mob_next IS NULL AND mmp_large_next = 0
      THEN NULL                                    -- no info about next year (V-Dem drives this)
      ELSE 0
    END AS outcome_mobilization,

    -- TIER 2: regime-threatening violence (armed conflict or coup)
    -- v0.7.1: third OR-channel = ACLED total fatalities ≥ 100. Same threshold
    -- as UCDP, so this adds a complementary measurement (different coders,
    -- different release cadence) rather than a different threshold. Critically
    -- fills UCDP's release-lag gap (UCDP ends 2024; ACLED runs through 2026).
    CASE
      WHEN coup_flag >= 1
        OR ucdp_fatal_next >= 100
        OR acled_total_fat_next >= 100
      THEN 1
      -- Zero-state = all sources are BELOW their threshold (not strictly = 0).
      -- Permissive: a country-year with 5 ACLED fatalities and 0 UCDP/coups
      -- is still no-major-political-violence = 0, not unknown = NULL.
      -- (The strict = 0 rule shrank the violence holdout from 122 to 76 in
      -- v0.7.1 testing — bug fixed here.)
      WHEN coup_flag = 0 AND ucdp_fatal_next < 100 AND acled_total_fat_next < 100
      THEN 0
      ELSE NULL
    END AS outcome_violence,

    -- LEGACY: OR of both tiers, preserved for backwards compatibility
    -- with anything that still reads unrest_next_year.
    CASE
      WHEN coup_flag >= 1
        OR ucdp_fatal_next >= 100
        OR acled_total_fat_next >= 100
        OR mmp_large_next >= 1
        OR (vdem_mob_next >= 4 AND year >= 2020)
        OR acled_violent_demo_fat_next >= 10
      THEN 1
      WHEN vdem_mob_next IS NULL AND coup_flag = 0 AND ucdp_fatal_next = 0
        AND mmp_large_next = 0 AND acled_violent_demo_fat_next = 0 AND acled_total_fat_next = 0
      THEN NULL
      ELSE 0
    END AS unrest_next_year
  FROM fused
")

dbGetQuery(con, "SELECT COUNT(*) AS rows FROM panel_labeled")
```

## Manual label overrides

Country-years where ground truth from news/events has reached us before
the upstream sources updated. Each row needs a citation in `source` and a
plain-language `reason`. Rows here override whatever the source-fusion
logic would have produced; `manual_label = 1` lets the fit step keep these
rows in training even when they fall past the standard train cutoff.

Current overrides (see `data/curated/manual_mobilization_labels.csv`):

- **IRN 2025** — Late Dec 2025 – Jan 2026 Grand Bazaar shutdown spread to
  all 31 provinces. V-Dem 2025 was coded before these events; will likely
  upgrade to level 4 in the 2027 V-Dem release.
- **CUB 2024** — Labels the 2025 wave: Cuban Conflict Observatory recorded
  11,268 protests during 2025, the largest mobilization in Havana since the
  July 2021 11J wave. V-Dem rated 2025 at level 1, plausibly under-coded
  given state-managed-economy information access constraints.

```{r}
#| label: manual-overrides
manual_labels <- read.csv("../data/curated/manual_mobilization_labels.csv",
                          stringsAsFactors = FALSE)
cat("Manual labels file: ", nrow(manual_labels), " row(s)\n", sep = "")
print(manual_labels[, c("iso3c", "year", "outcome_mobilization", "reason")])

# Add a flag column so the fit step can include these rows in training
# regardless of the year cutoff.
dbExecute(con, "ALTER TABLE panel_labeled ADD COLUMN manual_label INTEGER DEFAULT 0")

dbWriteTable(con, "manual_mobilization_labels", manual_labels, overwrite = TRUE)
dbExecute(con, "
  UPDATE panel_labeled p
  SET outcome_mobilization = m.outcome_mobilization,
      manual_label         = 1
  FROM manual_mobilization_labels m
  WHERE p.iso3c = m.iso3c AND p.year = m.year
")

cat("\nAfter override, manual-label row count: ",
    dbGetQuery(con, "SELECT COUNT(*) AS n FROM panel_labeled WHERE manual_label = 1")$n,
    "\n", sep = "")
```

## Outcome distributions

```{r}
#| label: distribution-mobilization
cat("Tier 1: outcome_mobilization\n")
dbGetQuery(con, "
  SELECT outcome_mobilization, COUNT(*) AS n,
         ROUND(100.0 * COUNT(*) / (SELECT COUNT(*) FROM panel_labeled WHERE outcome_mobilization IS NOT NULL), 1) AS pct
  FROM panel_labeled
  WHERE outcome_mobilization IS NOT NULL
  GROUP BY outcome_mobilization
  ORDER BY outcome_mobilization
")
```

```{r}
#| label: distribution-violence
cat("Tier 2: outcome_violence\n")
dbGetQuery(con, "
  SELECT outcome_violence, COUNT(*) AS n,
         ROUND(100.0 * COUNT(*) / (SELECT COUNT(*) FROM panel_labeled WHERE outcome_violence IS NOT NULL), 1) AS pct
  FROM panel_labeled
  WHERE outcome_violence IS NOT NULL
  GROUP BY outcome_violence
  ORDER BY outcome_violence
")
```

## Channel-by-channel source contribution

```{r}
#| label: source-contribution
dbGetQuery(con, "
  SELECT
    SUM(CASE WHEN coup_flag                    >= 1                  THEN 1 ELSE 0 END) AS n_coups,
    SUM(CASE WHEN ucdp_fatal_next              >= 100                THEN 1 ELSE 0 END) AS n_ucdp,
    SUM(CASE WHEN acled_total_fat_next         >= 100                THEN 1 ELSE 0 END) AS n_acled_violence,
    SUM(CASE WHEN mmp_large_next               >= 1                  THEN 1 ELSE 0 END) AS n_mmp,
    SUM(CASE WHEN vdem_mob_next                >= 4 AND year >= 2020 THEN 1 ELSE 0 END) AS n_vdem_fallback,
    SUM(CASE WHEN acled_violent_demo_fat_next  >= 10                 THEN 1 ELSE 0 END) AS n_acled_mobilization,
    SUM(outcome_mobilization) AS n_mobilization_total,
    SUM(outcome_violence)     AS n_violence_total,
    COUNT(*) AS n_total
  FROM panel_labeled
")
```

### How much does ACLED uniquely contribute vs. the existing channels?

Country-years flagged by ACLED that the prior channels (pre-v0.7) would have
missed. Honest measure of the marginal information ACLED adds.

```{r}
#| label: acled-marginal-contribution
dbGetQuery(con, "
  SELECT
    SUM(CASE WHEN acled_violent_demo_fat_next >= 10
              AND NOT (mmp_large_next >= 1 OR (vdem_mob_next >= 4 AND year >= 2020))
             THEN 1 ELSE 0 END) AS n_mob_acled_only,
    SUM(CASE WHEN acled_total_fat_next >= 100
              AND NOT (coup_flag >= 1 OR ucdp_fatal_next >= 100)
             THEN 1 ELSE 0 END) AS n_vio_acled_only
  FROM panel_labeled
")
```

## Outcome rate by regime category

```{r}
#| label: by-regime
dbGetQuery(con, "
  SELECT regime_type,
         COUNT(*) AS n_obs,
         ROUND(100.0 * AVG(outcome_mobilization), 1) AS pct_mobilization,
         ROUND(100.0 * AVG(outcome_violence), 1)     AS pct_violence
  FROM panel_labeled
  WHERE regime_type IS NOT NULL
    AND outcome_mobilization IS NOT NULL AND outcome_violence IS NOT NULL
  GROUP BY regime_type
  ORDER BY regime_type
")
```

## Cuba check

```{r}
#| label: cuba-check
dbGetQuery(con, "
  SELECT year, coup_flag, ucdp_fatal_next, mmp_large_next, vdem_mob_next,
         outcome_mobilization, outcome_violence
  FROM panel_labeled
  WHERE iso3c = 'CUB' AND year BETWEEN 2015 AND 2023
  ORDER BY year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
