---
title: "09 — Label outcome events (two-tier: mobilization + violence)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

**v0.6:** stratify the outcome into two tiers. The previous single
`unrest_next_year` flag fused four channels into one — which made the
prediction ambiguous ("USA 26%" could mean coup OR protest OR armed
conflict; same number meant different things in different countries).
The two tiers separate **disruptive mobilization** from
**regime-threatening violence**.

| Tier | Outcome column | Channels (OR-fused) | What "yes" means |
|---|---|---|---|
| 1 | `outcome_mobilization` | MMP per-capita large protest (`n_large_protests >= 1`) OR V-Dem `v2cagenmob_ord >= 4` | People in the streets at country-defining scale: Yellow Vests, BLM, Iran 2022, HK 2019, Tahrir, Iceland 2008 |
| 2 | `outcome_violence` | UCDP fatalities ≥100 OR Cline Center coup attempt | Regime-threatening violence: armed-conflict deaths or coup attempts |

We keep `unrest_next_year` as a legacy column (OR of both tiers) for
backwards compatibility with anything that still depends on it.

**Source coverage:**

- Mobilization: V-Dem v16 runs through 2025 ✓; MMP ends 2020 (still usable for training)
- Violence: UCDP runs through 2024 ✓; Cline coups end 2022 (still usable for training)

Output: `panel_labeled` table (`panel_features` + 3 outcome columns).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Build the two-tier label

```{r}
#| label: label
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

dbExecute(con, "
  CREATE OR REPLACE TABLE panel_labeled AS
  WITH fused AS (
    SELECT
      p.*,
      LEAD(mass_mobilization_ord) OVER (PARTITION BY p.iso3c ORDER BY p.year) AS mass_mob_next,

      -- Per-source next-year flags. NULL = no data at all (post-source-end),
      -- 0 = source ran but recorded no event, >0 = event(s) recorded.
      COALESCE(c.n_realized, 0) + COALESCE(c.n_attempted, 0)   AS coup_flag,
      COALESCE(u.fatalities_best, 0)                            AS ucdp_fatal_next,
      COALESCE(m.n_large_protests, 0)                           AS mmp_large_next,
      LEAD(mass_mobilization_ord) OVER (PARTITION BY p.iso3c ORDER BY p.year) AS vdem_mob_next
    FROM panel_features p
    LEFT JOIN coups_annual c ON c.iso3c = p.iso3c AND c.year = p.year + 1
    LEFT JOIN ucdp_annual  u ON u.iso3c = p.iso3c AND u.year = p.year + 1
    LEFT JOIN mmp_annual   m ON m.iso3c = p.iso3c AND m.year = p.year + 1
  )
  SELECT *,
    -- TIER 1: mass mobilization (protest/demonstration at country-defining scale)
    CASE
      WHEN mmp_large_next >= 1
        OR (vdem_mob_next >= 4 AND year >= 2020)  -- V-Dem fallback for post-MMP years
      THEN 1
      WHEN vdem_mob_next IS NULL AND mmp_large_next = 0
      THEN NULL                                   -- no info about next year
      ELSE 0
    END AS outcome_mobilization,

    -- TIER 2: regime-threatening violence (armed conflict or coup)
    CASE
      WHEN coup_flag >= 1
        OR ucdp_fatal_next >= 100
      THEN 1
      WHEN coup_flag = 0 AND ucdp_fatal_next = 0
      THEN 0
      ELSE NULL
    END AS outcome_violence,

    -- LEGACY: OR of both tiers, preserved for backwards compatibility
    -- with anything that still reads unrest_next_year.
    CASE
      WHEN coup_flag >= 1
        OR ucdp_fatal_next >= 100
        OR mmp_large_next >= 1
        OR (vdem_mob_next >= 4 AND year >= 2020)
      THEN 1
      WHEN vdem_mob_next IS NULL AND coup_flag = 0 AND ucdp_fatal_next = 0 AND mmp_large_next = 0
      THEN NULL
      ELSE 0
    END AS unrest_next_year
  FROM fused
")

dbGetQuery(con, "SELECT COUNT(*) AS rows FROM panel_labeled")
```

## Outcome distributions

```{r}
#| label: distribution-mobilization
cat("Tier 1: outcome_mobilization\n")
dbGetQuery(con, "
  SELECT outcome_mobilization, COUNT(*) AS n,
         ROUND(100.0 * COUNT(*) / (SELECT COUNT(*) FROM panel_labeled WHERE outcome_mobilization IS NOT NULL), 1) AS pct
  FROM panel_labeled
  WHERE outcome_mobilization IS NOT NULL
  GROUP BY outcome_mobilization
  ORDER BY outcome_mobilization
")
```

```{r}
#| label: distribution-violence
cat("Tier 2: outcome_violence\n")
dbGetQuery(con, "
  SELECT outcome_violence, COUNT(*) AS n,
         ROUND(100.0 * COUNT(*) / (SELECT COUNT(*) FROM panel_labeled WHERE outcome_violence IS NOT NULL), 1) AS pct
  FROM panel_labeled
  WHERE outcome_violence IS NOT NULL
  GROUP BY outcome_violence
  ORDER BY outcome_violence
")
```

## Channel-by-channel source contribution

```{r}
#| label: source-contribution
dbGetQuery(con, "
  SELECT
    SUM(CASE WHEN coup_flag      >= 1                       THEN 1 ELSE 0 END) AS n_coups,
    SUM(CASE WHEN ucdp_fatal_next >= 100                    THEN 1 ELSE 0 END) AS n_ucdp,
    SUM(CASE WHEN mmp_large_next  >= 1                      THEN 1 ELSE 0 END) AS n_mmp,
    SUM(CASE WHEN vdem_mob_next   >= 4 AND year >= 2020     THEN 1 ELSE 0 END) AS n_vdem_fallback,
    SUM(outcome_mobilization) AS n_mobilization_total,
    SUM(outcome_violence)     AS n_violence_total,
    COUNT(*) AS n_total
  FROM panel_labeled
")
```

## Outcome rate by regime category

```{r}
#| label: by-regime
dbGetQuery(con, "
  SELECT regime_type,
         COUNT(*) AS n_obs,
         ROUND(100.0 * AVG(outcome_mobilization), 1) AS pct_mobilization,
         ROUND(100.0 * AVG(outcome_violence), 1)     AS pct_violence
  FROM panel_labeled
  WHERE regime_type IS NOT NULL
    AND outcome_mobilization IS NOT NULL AND outcome_violence IS NOT NULL
  GROUP BY regime_type
  ORDER BY regime_type
")
```

## Cuba check

```{r}
#| label: cuba-check
dbGetQuery(con, "
  SELECT year, coup_flag, ucdp_fatal_next, mmp_large_next, vdem_mob_next,
         outcome_mobilization, outcome_violence
  FROM panel_labeled
  WHERE iso3c = 'CUB' AND year BETWEEN 2015 AND 2023
  ORDER BY year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
