---
title: "09 — Label outcome events (multi-source fusion)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

For each (country, year), look at **next year's events** across four
independent sources and flag the country-year as unrest if **any** of these
fire:

| Source | Threshold | Captures |
|---|---|---|
| Cline Center coups | `n_realized + n_attempted >= 1` | Regime-change attempts |
| UCDP GED | `fatalities_best >= 100` next year | Organized armed-violence episodes |
| MMP large protests | `n_large_protests >= 1` next year | Sustained mass mobilization at scale (≥1,000 participants) |
| V-Dem (fallback) | `v2cagenmob_ord >= 4` next year | When MMP doesn't cover (post-2020) |

Why a fusion: V-Dem alone gave us flat base rates (47–60% across regimes)
because the ordinal is too broad. Event-based sources are sparser and more
specific — they should let closed-autocracy rates and liberal-democracy
rates diverge as they should.

V-Dem stays as a **post-2020 fallback** because MMP ends in 2020.

Output: `panel_labeled` table (`panel_features` + outcome columns).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Build the lead-label panel

DuckDB does the heavy lifting in one SQL pass. For each (iso3c, year) row in
`panel_features`, we left-join the next-year row from each event source and
OR the thresholds.

```{r}
#| label: label
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

dbExecute(con, "
  CREATE OR REPLACE TABLE panel_labeled AS
  WITH next_yr AS (
    SELECT iso3c, year, year + 1 AS forecast_year
    FROM panel_features
  ),
  fused AS (
    SELECT
      p.*,
      LEAD(mass_mobilization_ord) OVER (PARTITION BY p.iso3c ORDER BY p.year) AS mass_mob_next,

      -- per-source flags for next year, defaulting NULL to 0
      COALESCE(c.n_realized, 0)  + COALESCE(c.n_attempted, 0)   AS coup_flag,
      COALESCE(u.fatalities_best, 0)                            AS ucdp_fatal_next,
      COALESCE(m.n_large_protests, 0)                           AS mmp_large_next,
      LEAD(mass_mobilization_ord) OVER (PARTITION BY p.iso3c ORDER BY p.year) AS vdem_mob_next
    FROM panel_features p
    LEFT JOIN coups_annual c ON c.iso3c = p.iso3c AND c.year = p.year + 1
    LEFT JOIN ucdp_annual  u ON u.iso3c = p.iso3c AND u.year = p.year + 1
    LEFT JOIN mmp_annual   m ON m.iso3c = p.iso3c AND m.year = p.year + 1
  )
  SELECT *,
    CASE
      WHEN coup_flag >= 1
        OR ucdp_fatal_next >= 100
        OR mmp_large_next >= 1
        OR (vdem_mob_next >= 4 AND year >= 2020) -- V-Dem fallback for post-MMP years
      THEN 1
      WHEN vdem_mob_next IS NULL AND coup_flag = 0 AND ucdp_fatal_next = 0 AND mmp_large_next = 0
      THEN NULL    -- no info about next year for this country
      ELSE 0
    END AS unrest_next_year
  FROM fused
")

dbGetQuery(con, "SELECT COUNT(*) AS rows FROM panel_labeled")
```

## Outcome distribution

```{r}
#| label: distribution
dbGetQuery(con, "
  SELECT unrest_next_year, COUNT(*) AS n,
         ROUND(100.0 * COUNT(*) / (SELECT COUNT(*) FROM panel_labeled WHERE unrest_next_year IS NOT NULL), 1) AS pct
  FROM panel_labeled
  WHERE unrest_next_year IS NOT NULL
  GROUP BY unrest_next_year
  ORDER BY unrest_next_year
")
```

## Which source fires most often?

Useful for diagnosing whether one source is dominating the label.

```{r}
#| label: source-contribution
dbGetQuery(con, "
  SELECT
    SUM(CASE WHEN coup_flag      >= 1 THEN 1 ELSE 0 END) AS n_coups,
    SUM(CASE WHEN ucdp_fatal_next >= 100 THEN 1 ELSE 0 END) AS n_ucdp,
    SUM(CASE WHEN mmp_large_next  >= 1 THEN 1 ELSE 0 END) AS n_mmp,
    SUM(CASE WHEN vdem_mob_next   >= 4 AND year >= 2020 THEN 1 ELSE 0 END) AS n_vdem,
    SUM(unrest_next_year) AS n_any_unrest,
    COUNT(*) AS n_total
  FROM panel_labeled
  WHERE unrest_next_year IS NOT NULL
")
```

## Outcome rate by regime category

```{r}
#| label: by-regime
dbGetQuery(con, "
  SELECT regime_type,
         COUNT(*) AS n_obs,
         SUM(unrest_next_year) AS n_unrest,
         ROUND(100.0 * AVG(unrest_next_year), 1) AS pct_unrest
  FROM panel_labeled
  WHERE unrest_next_year IS NOT NULL AND regime_type IS NOT NULL
  GROUP BY regime_type
  ORDER BY regime_type
")
```

## Cuba check

```{r}
#| label: cuba-check
dbGetQuery(con, "
  SELECT year, coup_flag, ucdp_fatal_next, mmp_large_next, vdem_mob_next, unrest_next_year
  FROM panel_labeled
  WHERE iso3c = 'CUB' AND year BETWEEN 2015 AND 2023
  ORDER BY year
")
```

## Cleanup

```{r}
#| label: cleanup
dbDisconnect(con, shutdown = TRUE)
```
