---
title: "10 — Fit regularized unrest-risk models"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

**v0.5 architecture** (extends v0.4 cross-sectional PITF framing):

- One pooled **LASSO** logistic regression on **raw cross-country feature
  levels** (no within-country z-scoring; no regime dummies — civil_liberties
  is the regime-quality channel).
- **Monotonicity constraints** on every coefficient: each feature is forced
  to the sign its substantive interpretation requires. Fixes the v0.4
  problem where closedness-under-reporting bias was making civil_liberties
  β = +0.73 ("more civil liberties → more reported unrest") even though
  the structural reading is the opposite.
- **`political_violence_ord` is lagged by 1 year** (`political_violence_ord_lag1`
  in features). The v0.4 unlagged version overlapped with the outcome label
  (our `unrest_next_year` includes the post-2020 V-Dem `v2cagenmob_ord >= 4`
  fallback) — concurrent violence was partly predicting itself. Two-year gap.
- **`lambda.min` instead of `lambda.min`** — stronger regularization, fewer
  surviving features, more honest about what's actually predictive.
- **Curated unemployment for state-managed economies** (Cuba, Belarus, PRK,
  Venezuela, Eritrea, Turkmenistan, Cambodia, Laos) via the COALESCE chain
  in `08_build_panel.qmd`. The official ILO/WDI rate doesn't reflect
  informal-sector / underemployment reality for these countries.

Coefficients persist with a **bootstrap-derived empirical covariance** in
place of `vcov()` (penalized estimators don't have a closed-form sampling
distribution). The bootstrap fits use the same `lambda.min` and monotonicity
constraints as the headline fit.

**Features (10), all raw cross-country levels:**

| Feature | Source | Constraint |
|---|---|---|
| `civil_liberties` | V-Dem 0–1 | β ≤ 0 |
| `log_gdp_per_cap` | log WDI/WEO GDP per capita | β ≤ 0 |
| `loser_consent` | V-Dem `v2elaccept_ord`, 0–4 | β ≤ 0 |
| `conflict_neighbors_count` | UCDP-derived (lag-1) | β ≥ 0 |
| `gdp_growth` | WDI/WEO annual % | β ≤ 0 |
| `cpi_inflation` | WDI/IMF chain, clipped [−10, 200] | β ≥ 0 |
| `unemployment` | WDI/ILO + curated supplement for state-managed economies | β ≥ 0 |
| `political_violence_ord_lag1` | V-Dem 0–4, lagged by 1 year | β ≥ 0 |
| `log_refugee_outflow` | UNHCR log1p | β ≥ 0 |
| `coup_proof_fragmentation` | Pilster-Böhmelt | β ≥ 0 |

**Training set filter:** listwise-drop rows where `cpi_imputed = 1`
(country-year had no real CPI source — WDI/IMF-curated/IMF WEO — and
inflation was region-median imputed). Avoids fitting on synthetic inflation
for PRK, pre-2018 Cuba, Somalia, etc. `cpi_imputed` is preserved as a
data-quality flag in the dashboard, not as a model feature.

**Train/test:** train on 1995–2022; predict unrest in 2023 on holdout 2023.

**v0.4 deletions from v0.3:** within-country z-scoring; `z_ffpi_avg` (global
food-price index identical across countries in a given year — no
cross-sectional variation); the `factor(regime_type)` dummies in the
pooled formula (collinear with civil_liberties).

de Bruin's `counterbalancing` / `politicization` and Fariss / Polity2 remain
loaded in the panel for analysis but are excluded from the fit (Fariss/Polity
correlate ~0.9 with civil_liberties; de Bruin source ends 2010).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(glmnet)
```

## Pull the labeled panel

```{r}
#| label: pull
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

features <- c(
  "civil_liberties", "log_gdp_per_cap", "loser_consent",
  "conflict_neighbors_count",
  "gdp_growth", "cpi_inflation", "unemployment",
  "political_violence_ord_lag1",
  "log_refugee_outflow", "coup_proof_fragmentation"
)
# v0.4: raw cross-country levels only. No within-country z's. No regime
# dummies in the pooled formula (civil_liberties is the regime-quality
# channel). cpi_inflation is clipped at [-10, 200]% in 08; log_refugee_outflow
# is log1p-transformed in 08. cpi_imputed is pulled separately to listwise-drop
# training rows where the CPI was region-median imputed — it is NOT a model
# feature (would absorb the closedness→under-reporting measurement bias).

needed_cols <- paste(c("iso3c", "year", "regime_type", "unrest_next_year",
                       "cpi_imputed", features), collapse = ", ")
panel <- dbGetQuery(con, sprintf("SELECT %s FROM panel_labeled", needed_cols))
dbDisconnect(con, shutdown = TRUE)

cat("Total panel rows:", nrow(panel), "\n")
cat("Rows with non-NA outcome:", sum(!is.na(panel$unrest_next_year)), "\n")
```

## Filter to complete cases for fitting

The model needs every feature present plus a valid outcome and regime
category. We *also* drop rows where `cpi_imputed = 1` — i.e. country-years
whose CPI came from region-median imputation rather than a real source.
See header prose for why: those rows are concentrated in closed autocracies
where event-based outcome labels are also under-reported, so keeping them
in would let the model fit a measurement-bias correlation rather than real
signal.

```{r}
#| label: filter
feature_complete <- complete.cases(panel[, c("regime_type", "unrest_next_year", features)])
real_cpi         <- panel$cpi_imputed == 0
complete <- panel[feature_complete & real_cpi, ]
train <- complete[complete$year <= 2022, ]
holdout <- complete[complete$year == 2023, ]

n_dropped_for_imputed_cpi <- sum(feature_complete & !real_cpi & panel$year <= 2022)
cat("Train rows: ", nrow(train),
    "  | unrest rate:", round(mean(train$unrest_next_year), 3),
    "  | dropped for imputed CPI:", n_dropped_for_imputed_cpi, "\n")
cat("Holdout rows:", nrow(holdout),
    "  | unrest rate:", round(mean(holdout$unrest_next_year), 3), "\n")
```

## Fit one LASSO model per regime stratum

For each stratum: pick lambda via `cv.glmnet` on the full stratum, then
nonparametric-bootstrap the rows (stratified on the outcome to preserve
event rate) and refit at the chosen lambda B times. The empirical covariance
of the B coefficient vectors becomes the drop-in replacement for `vcov()`
that the dashboard consumes downstream.

```{r}
#| label: fit
formula_str <- paste("unrest_next_year ~", paste(features, collapse = " + "))
form <- as.formula(formula_str)

regime_labels <- c("0" = "closed_autocracy",
                   "1" = "electoral_autocracy",
                   "2" = "electoral_democracy",
                   "3" = "liberal_democracy")

B_bootstrap <- 500
# LASSO (alpha = 1) for systematic feature selection: features whose
# coefficient is exactly zero at lambda.min did not earn their place
# in the CV-deviance budget. We use lambda.min (more regularization)
# instead of lambda.min so coefficients must defend themselves against
# stronger shrinkage. v0.2 used ridge (alpha = 0); v0.3+ uses LASSO.
alpha_fit <- 1
set.seed(20260517)

# v0.5: monotonicity constraints on coefficients. Forces structural
# anchors to the sign their substantive interpretation requires, even
# when training data correlations would push the other way (e.g.,
# closedness-under-reporting bias was making civil_liberties +0.73,
# saying "more civil liberties = more reported unrest"). Constraints
# apply on the *original* (unstandardized) coefficient scale; glmnet
# handles standardization internally and de-standardizes the returned
# coefficients, so constraints work directly.
# Monotonicity constraints in features-vector order.
# Sign:  -1 = β ≤ 0 required;  +1 = β ≥ 0 required.
# Lower/upper limits per coefficient passed to glmnet as plain numeric
# vectors (no names, no NAs).
feature_signs <- c(-1, -1, -1, +1, -1, +1, +1, +1, +1, +1)
names(feature_signs) <- c("civil_liberties", "log_gdp_per_cap", "loser_consent",
                          "conflict_neighbors_count", "gdp_growth", "cpi_inflation",
                          "unemployment", "political_violence_ord_lag1",
                          "log_refugee_outflow", "coup_proof_fragmentation")
stopifnot(setequal(names(feature_signs), features))

build_limits <- function(design_cols) {
  # Verify every design-matrix column has a defined sign — fail loudly
  # rather than passing NA into glmnet (where the error is opaque).
  missing_signs <- setdiff(design_cols, names(feature_signs))
  if (length(missing_signs) > 0) {
    stop("Design-matrix columns without a defined sign: ",
         paste(missing_signs, collapse = ", "))
  }
  signs <- as.numeric(feature_signs[design_cols])  # plain numeric, no names
  lower <- rep(-Inf, length(signs)); lower[signs == +1] <- 0
  upper <- rep( Inf, length(signs)); upper[signs == -1] <- 0
  list(lower = lower, upper = upper)
}

models <- list()
fit_summary <- data.frame()

for (k in 0:3) {
  stratum <- train[train$regime_type == k, ]
  if (nrow(stratum) < 30) {
    cat("Regime", k, "(", regime_labels[as.character(k)], "): only",
        nrow(stratum), "rows — skipping\n")
    next
  }

  # Design matrix — drop intercept column; glmnet adds its own.
  mm <- model.matrix(form, stratum)
  x  <- mm[, -1, drop = FALSE]
  y  <- stratum$unrest_next_year

  # Monotonicity constraints in column order of the design matrix.
  lims <- build_limits(colnames(x))

  # Fewer CV folds when events are sparse so each fold sees ≥3 events.
  n_events <- sum(y == 1)
  nfolds   <- if (n_events < 30) 5 else 10

  fit_cv      <- cv.glmnet(x, y, family = "binomial",
                           alpha = alpha_fit, nfolds = nfolds,
                           lower.limits = lims$lower,
                           upper.limits = lims$upper)
  lambda_best <- fit_cv$lambda.min

  coef_full  <- as.numeric(coef(fit_cv, s = "lambda.min"))
  coef_names <- rownames(coef(fit_cv, s = "lambda.min"))
  names(coef_full) <- coef_names

  # Bootstrap at the SAME lambda — sampling variability around coefficients
  # at a fixed regularization, not lambda variability itself.
  idx_pos <- which(y == 1)
  idx_neg <- which(y == 0)
  boot_coefs <- matrix(NA_real_, nrow = B_bootstrap, ncol = length(coef_full),
                       dimnames = list(NULL, coef_names))
  for (b in seq_len(B_bootstrap)) {
    rows <- c(sample(idx_pos, length(idx_pos), replace = TRUE),
              sample(idx_neg, length(idx_neg), replace = TRUE))
    fit_b <- glmnet(x[rows, , drop = FALSE], y[rows],
                    family = "binomial", alpha = alpha_fit,
                    lambda = lambda_best,
                    lower.limits = lims$lower,
                    upper.limits = lims$upper)
    boot_coefs[b, ] <- as.numeric(coef(fit_b))
  }
  coef_mean <- colMeans(boot_coefs)
  coef_vcov <- cov(boot_coefs) + diag(1e-8, length(coef_full))

  models[[as.character(k)]] <- list(
    fit_cv         = fit_cv,
    lambda         = lambda_best,
    feature_order  = coef_names,
    coefficients   = coef_full,
    coef_mean      = coef_mean,
    vcov           = coef_vcov,
    n_train        = nrow(stratum),
    n_events       = n_events,
    nfolds         = nfolds,
    n_bootstrap    = B_bootstrap
  )

  dev_idx        <- which.min(abs(fit_cv$lambda - lambda_best))
  null_dev       <- fit_cv$glmnet.fit$nulldev
  dev_at_lambda  <- (1 - fit_cv$glmnet.fit$dev.ratio[dev_idx]) * null_dev
  fit_summary <- rbind(fit_summary, data.frame(
    regime    = regime_labels[as.character(k)],
    n_obs     = nrow(stratum),
    n_unrest  = n_events,
    lambda    = signif(lambda_best, 3),
    deviance  = round(dev_at_lambda, 1),
    null_dev  = round(null_dev, 1),
    pseudo_r2 = round(1 - dev_at_lambda / null_dev, 3)
  ))
}
fit_summary
```

## Fit pooled LASSO (primary cross-country model)

The stratified fits above provide within-regime context for analysts. The
**pooled fit is the headline model**: a single LASSO across all countries
where `civil_liberties` is the only regime-quality channel — no regime
dummies. Same bootstrap-empirical-cov procedure as the stratified fits.

```{r}
#| label: fit-pooled
pool_form <- as.formula(paste(
  "unrest_next_year ~",
  paste(features, collapse = " + ")
))
mm_pool <- model.matrix(pool_form, train)
x_pool  <- mm_pool[, -1, drop = FALSE]
y_pool  <- train$unrest_next_year

lims_pool <- build_limits(colnames(x_pool))

fit_pool_cv  <- cv.glmnet(x_pool, y_pool, family = "binomial",
                          alpha = alpha_fit, nfolds = 10,
                          lower.limits = lims_pool$lower,
                          upper.limits = lims_pool$upper)
lambda_pool  <- fit_pool_cv$lambda.min

coef_pool       <- as.numeric(coef(fit_pool_cv, s = "lambda.min"))
coef_pool_names <- rownames(coef(fit_pool_cv, s = "lambda.min"))
names(coef_pool) <- coef_pool_names

# Bootstrap pooled coefficients at the chosen lambda.
idx_pos_p <- which(y_pool == 1)
idx_neg_p <- which(y_pool == 0)
boot_pool <- matrix(NA_real_, nrow = B_bootstrap, ncol = length(coef_pool),
                    dimnames = list(NULL, coef_pool_names))
for (b in seq_len(B_bootstrap)) {
  rows <- c(sample(idx_pos_p, length(idx_pos_p), replace = TRUE),
            sample(idx_neg_p, length(idx_neg_p), replace = TRUE))
  fit_b <- glmnet(x_pool[rows, , drop = FALSE], y_pool[rows],
                  family = "binomial", alpha = alpha_fit,
                  lambda = lambda_pool,
                  lower.limits = lims_pool$lower,
                  upper.limits = lims_pool$upper)
  boot_pool[b, ] <- as.numeric(coef(fit_b))
}
coef_pool_vcov <- cov(boot_pool) + diag(1e-8, length(coef_pool))

pooled_model <- list(
  fit_cv        = fit_pool_cv,
  lambda        = lambda_pool,
  formula       = pool_form,
  feature_order = coef_pool_names,
  coefficients  = coef_pool,
  vcov          = coef_pool_vcov,
  n_train       = nrow(train),
  n_events      = sum(y_pool == 1),
  n_bootstrap   = B_bootstrap
)

cat("Pooled fit:\n")
cat("  lambda:    ", signif(lambda_pool, 3), "\n")
cat("  n_train:   ", nrow(train), "\n")
cat("  n_events:  ", sum(y_pool == 1), "\n")
cat("  features:  ", length(coef_pool) - 1, "\n")
```

## Inspect pooled coefficients

These apply to ALL countries. Civil liberties carries the regime-quality
signal; log GDP per capita carries the development/wealth signal; the rest
are economic and political stress levels.

```{r}
#| label: coef-pooled
co_pool <- data.frame(
  estimate = pooled_model$coefficients,
  boot_sd  = sqrt(diag(pooled_model$vcov))
)
round(co_pool, 3)
```

## LASSO feature-survival report

Features whose coefficient is exactly zero at `lambda.min` were dropped by
LASSO under the v0.5 monotonicity constraints. Note: a zeroed coefficient
under monotonicity means EITHER (a) the feature didn't pay its way in
CV-deviance, OR (b) the unconstrained fit wanted the wrong sign and the
constraint pushed it to zero. Both are valid reasons to drop.

```{r}
#| label: lasso-survival
pool_coef_vec <- pooled_model$coefficients
is_intercept <- names(pool_coef_vec) == "(Intercept)"
covariate_coefs <- pool_coef_vec[!is_intercept]

survived <- names(covariate_coefs)[covariate_coefs != 0]
dropped  <- names(covariate_coefs)[covariate_coefs == 0]

cat("Survived LASSO (", length(survived), "of", length(covariate_coefs), "):\n")
for (f in survived) cat("  ", f, "  beta =", round(covariate_coefs[[f]], 4), "\n")
cat("\nDropped by LASSO (beta exactly 0):\n")
for (f in dropped)  cat("  ", f, "\n")
```

## Inspect coefficients (closed autocracy)

LASSO has no closed-form coefficient SE — show point estimate alongside
the bootstrap SD (sqrt of the empirical-cov diagonal).

```{r}
#| label: coef-autocracy
ca <- models[["0"]]
co <- data.frame(
  estimate = ca$coefficients,
  boot_sd  = sqrt(diag(ca$vcov))
)
round(co, 3)
```

## Inspect coefficients (electoral autocracy)

```{r}
#| label: coef-elect-auto
ea <- models[["1"]]
co <- data.frame(
  estimate = ea$coefficients,
  boot_sd  = sqrt(diag(ea$vcov))
)
round(co, 3)
```

## Save fitted artifacts to disk for next notebooks

We pickle the fitted models so `11_calibration.qmd` and `12_export_artifacts.qmd`
can load them without re-fitting.

```{r}
#| label: persist
dir.create("../data/model", showWarnings = FALSE, recursive = TRUE)
saveRDS(
  list(
    # Pooled model is the primary "absolute likelihood" scorer that the
    # dashboard uses for the headline probability. Per-stratum models are
    # retained for secondary analyst context (within-regime ranking) but
    # never drive the dashboard's primary number.
    pooled_model  = pooled_model,
    models        = models,
    features      = features,
    formula       = form,
    pool_formula  = pool_form,
    regime_labels = regime_labels,
    train_years   = c(min(train$year), max(train$year)),
    n_train       = nrow(train),
    model_version = "unrest-risk-v0.5-monotonic-constrained",
    n_bootstrap   = B_bootstrap,
    fitted_at     = Sys.time()
  ),
  "../data/model/fit.rds"
)
cat("Saved to data/model/fit.rds\n")
```
