---
title: "10 ŌĆö Fit the regime-stratified ridge logistic regression"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Four separate ridge logistic regressions (one per V-Dem RoW regime category)
predicting `unrest_next_year`. Lambda chosen per stratum via `cv.glmnet`.
Coefficients persist with a **bootstrap-derived empirical covariance** in
place of `vcov()` (penalized estimators don't have a closed-form sampling
distribution); the dashboard still samples 90% CIs from `MVN(coef, cov)`,
just from the empirical cov.

**Features (11):**

*Absolute-level* ŌĆö "where this country sits structurally":
- `civil_liberties` ŌĆö V-Dem 0ŌĆō1 index
- `fariss_hr_score` ŌĆö Fariss latent human-rights protection (added v0.2)
- `polity2` ŌĆö Polity5 democracyŌĆōautocracy score (added v0.2)

*Within-country z* ŌĆö "how stressed vs the country's own history":
- `z_gdp_growth` ŌĆö economic shock
- `z_cpi_inflation` ŌĆö inflation shock (CPI sourced via WDI ŌåÆ IMF-curated ŌåÆ IMF WEO chain in `08_build_panel.qmd`; region-median imputed only for ~5% of training rows, flagged below)
- `z_unemployment` ŌĆö labor-market shock
- `z_refugee_outflow` ŌĆö population-leaving shock (log1p before z)
- `z_political_violence_ord` ŌĆö recent violence vs own norm
- `z_ffpi_avg` ŌĆö global food-price pressure
- `z_coup_proof_fragmentation` ŌĆö Pilster-B├Čhmelt effective-number of forces

*Imputation indicator* ŌĆö added v0.3 after we found the listwise filter below
was a silent no-op (every NA had been region-median imputed upstream, so
countries with no real CPI source ŌĆö PRK/SOM pre-2013/parts of ERI/TKM ŌĆö
were entering training with synthetic neighborhood values masquerading as
signal):
- `cpi_imputed` ŌĆö 1 when no real CPI source (WDI/IMF-curated/IMF-WEO)
  existed for this country-year. Lets the model learn that imputed-CPI
  rows are a distinct subpopulation.

Z-scores are winsorized at ┬▒3 SDs (in `08_build_panel.qmd`) so countries
with tiny historical baselines (e.g., UK's near-zero refugee outflow) don't
blow up when a small absolute number arrives.

The hybrid (levels for structural regime signals, z-scores for dynamic
economic/conflict signals) captures both "where this country sits" and "how
it's moving." Three regime-quality levels (civil_liberties, Fariss, polity2)
are correlated; ridge keeps them all without the collinear sign-flip that
killed an unpenalized fit (skill 0.044 ŌåÆ 0.007 with Fariss naively added).

de Bruin's `counterbalancing` and `politicization` remain excluded: source
coverage ends 2010 and a 14-year carry to 2024 isn't defensible.

**Train/test:** train on 1995ŌĆō2022; predict unrest in 2023 on holdout 2023.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(glmnet)
```

## Pull the labeled panel

```{r}
#| label: pull
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

features <- c(
  # Absolute level ŌĆö structural (three correlated regime-quality signals;
  # ridge keeps all three without collinear sign-flip)
  "civil_liberties", "fariss_hr_score", "polity2",
  # Economic stress (within-country z, region-median-imputed)
  "z_gdp_growth", "z_cpi_inflation", "z_unemployment",
  # Political stress
  "z_political_violence_ord",
  # External / structural
  "z_refugee_outflow", "z_ffpi_avg", "z_coup_proof_fragmentation",
  # Imputation indicator (binary, used raw ŌĆö not z-scored)
  "cpi_imputed"
)

needed_cols <- paste(c("iso3c", "year", "regime_type", "unrest_next_year", features), collapse = ", ")
panel <- dbGetQuery(con, sprintf("SELECT %s FROM panel_labeled", needed_cols))
dbDisconnect(con, shutdown = TRUE)

cat("Total panel rows:", nrow(panel), "\n")
cat("Rows with non-NA outcome:", sum(!is.na(panel$unrest_next_year)), "\n")
```

## Filter to complete cases for fitting

The model needs every feature present plus a valid outcome and regime
category. Listwise deletion is the simplest reasonable default for v0.

```{r}
#| label: filter
complete <- panel[complete.cases(panel[, c("regime_type", "unrest_next_year", features)]), ]
train <- complete[complete$year <= 2022, ]
holdout <- complete[complete$year == 2023, ]

cat("Train rows: ", nrow(train),
    "  | unrest rate:", round(mean(train$unrest_next_year), 3), "\n")
cat("Holdout rows:", nrow(holdout),
    "  | unrest rate:", round(mean(holdout$unrest_next_year), 3), "\n")
```

## Fit one ridge model per regime stratum

For each stratum: pick lambda via `cv.glmnet` on the full stratum, then
nonparametric-bootstrap the rows (stratified on the outcome to preserve
event rate) and refit at the chosen lambda B times. The empirical covariance
of the B coefficient vectors becomes the drop-in replacement for `vcov()`
that the dashboard consumes downstream.

```{r}
#| label: fit
formula_str <- paste("unrest_next_year ~", paste(features, collapse = " + "))
form <- as.formula(formula_str)

regime_labels <- c("0" = "closed_autocracy",
                   "1" = "electoral_autocracy",
                   "2" = "electoral_democracy",
                   "3" = "liberal_democracy")

B_bootstrap <- 500
alpha_ridge <- 0  # pure ridge
set.seed(20260517)

models <- list()
fit_summary <- data.frame()

for (k in 0:3) {
  stratum <- train[train$regime_type == k, ]
  if (nrow(stratum) < 30) {
    cat("Regime", k, "(", regime_labels[as.character(k)], "): only",
        nrow(stratum), "rows ŌĆö skipping\n")
    next
  }

  # Design matrix ŌĆö drop intercept column; glmnet adds its own.
  mm <- model.matrix(form, stratum)
  x  <- mm[, -1, drop = FALSE]
  y  <- stratum$unrest_next_year

  # Fewer CV folds when events are sparse so each fold sees Ōēź3 events.
  n_events <- sum(y == 1)
  nfolds   <- if (n_events < 30) 5 else 10

  fit_cv      <- cv.glmnet(x, y, family = "binomial",
                           alpha = alpha_ridge, nfolds = nfolds)
  lambda_best <- fit_cv$lambda.min

  coef_full  <- as.numeric(coef(fit_cv, s = "lambda.min"))
  coef_names <- rownames(coef(fit_cv, s = "lambda.min"))
  names(coef_full) <- coef_names

  # Bootstrap at the SAME lambda ŌĆö sampling variability around coefficients
  # at a fixed regularization, not lambda variability itself.
  idx_pos <- which(y == 1)
  idx_neg <- which(y == 0)
  boot_coefs <- matrix(NA_real_, nrow = B_bootstrap, ncol = length(coef_full),
                       dimnames = list(NULL, coef_names))
  for (b in seq_len(B_bootstrap)) {
    rows <- c(sample(idx_pos, length(idx_pos), replace = TRUE),
              sample(idx_neg, length(idx_neg), replace = TRUE))
    fit_b <- glmnet(x[rows, , drop = FALSE], y[rows],
                    family = "binomial", alpha = alpha_ridge,
                    lambda = lambda_best)
    boot_coefs[b, ] <- as.numeric(coef(fit_b))
  }
  coef_mean <- colMeans(boot_coefs)
  coef_vcov <- cov(boot_coefs) + diag(1e-8, length(coef_full))

  models[[as.character(k)]] <- list(
    fit_cv         = fit_cv,
    lambda         = lambda_best,
    feature_order  = coef_names,
    coefficients   = coef_full,
    coef_mean      = coef_mean,
    vcov           = coef_vcov,
    n_train        = nrow(stratum),
    n_events       = n_events,
    nfolds         = nfolds,
    n_bootstrap    = B_bootstrap
  )

  dev_idx        <- which.min(abs(fit_cv$lambda - lambda_best))
  null_dev       <- fit_cv$glmnet.fit$nulldev
  dev_at_lambda  <- (1 - fit_cv$glmnet.fit$dev.ratio[dev_idx]) * null_dev
  fit_summary <- rbind(fit_summary, data.frame(
    regime    = regime_labels[as.character(k)],
    n_obs     = nrow(stratum),
    n_unrest  = n_events,
    lambda    = signif(lambda_best, 3),
    deviance  = round(dev_at_lambda, 1),
    null_dev  = round(null_dev, 1),
    pseudo_r2 = round(1 - dev_at_lambda / null_dev, 3)
  ))
}
fit_summary
```

## Inspect coefficients (closed autocracy)

Ridge has no closed-form coefficient SE ŌĆö show point estimate alongside
the bootstrap SD (sqrt of the empirical-cov diagonal).

```{r}
#| label: coef-autocracy
ca <- models[["0"]]
co <- data.frame(
  estimate = ca$coefficients,
  boot_sd  = sqrt(diag(ca$vcov))
)
rownames(co) <- gsub("^z_", "", rownames(co))
round(co, 3)
```

## Inspect coefficients (electoral autocracy)

```{r}
#| label: coef-elect-auto
ea <- models[["1"]]
co <- data.frame(
  estimate = ea$coefficients,
  boot_sd  = sqrt(diag(ea$vcov))
)
rownames(co) <- gsub("^z_", "", rownames(co))
round(co, 3)
```

## Save fitted artifacts to disk for next notebooks

We pickle the fitted models so `11_calibration.qmd` and `12_export_artifacts.qmd`
can load them without re-fitting.

```{r}
#| label: persist
dir.create("../data/model", showWarnings = FALSE, recursive = TRUE)
saveRDS(
  list(
    models        = models,
    features      = features,
    formula       = form,
    regime_labels = regime_labels,
    train_years   = c(min(train$year), max(train$year)),
    n_train       = nrow(train),
    model_version = "unrest-risk-v0.2-ridge",
    n_bootstrap   = B_bootstrap,
    fitted_at     = Sys.time()
  ),
  "../data/model/fit.rds"
)
cat("Saved to data/model/fit.rds\n")
```
