---
title: "10 — Fit two-tier risk models (mobilization + violence)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

**v0.6 architecture** (extends v0.5 with two-tier outcome stratification):

- **Two outcome tiers fit separately**:
  - `outcome_mobilization` = MMP large protest OR V-Dem `v2cagenmob_ord ≥ 4` —
    the "people in the streets at country-defining scale" channel
  - `outcome_violence` = UCDP fatalities ≥100 OR Cline coup —
    the "regime-threatening armed event" channel

  Each tier gets its own pooled LASSO + 4 stratified-by-regime LASSOs.
  The dashboard reports both probabilities separately.

- Per-tier model architecture is unchanged from v0.5:
  - Pooled LASSO on raw cross-country feature levels (no within-country z; no
    regime dummies — civil_liberties is the regime-quality channel)
  - Monotonicity constraints on every coefficient (each feature forced to the
    sign its substantive interpretation requires)
  - `political_violence_ord` lagged by 1 year (kills concurrent-V-Dem-violence
    autocorrelation with the outcome label)
  - Curated unemployment for state-managed economies via the 08 COALESCE chain
  - Coefficients persist with bootstrap-derived empirical covariance

The two-tier split changes the *outcome*, not the *features*. Same 10 features
predict both. The interesting empirical question is whether economic features
matter more for `outcome_violence` (armed conflict / coup) than for
`outcome_mobilization` (protest) — they should, based on civil-war literature.

**Features (10), all raw cross-country levels:**

| Feature | Source | Constraint |
|---|---|---|
| `civil_liberties` | V-Dem 0–1 | β ≤ 0 |
| `log_gdp_per_cap` | log WDI/WEO GDP per capita | β ≤ 0 |
| `loser_consent` | V-Dem `v2elaccept_ord`, 0–4 | β ≤ 0 |
| `conflict_neighbors_count` | UCDP-derived (lag-1) | β ≥ 0 |
| `gdp_growth` | WDI/WEO annual % | β ≤ 0 |
| `cpi_inflation` | WDI/IMF chain, clipped [−10, 200] | β ≥ 0 |
| `unemployment` | WDI/ILO + curated supplement for state-managed economies | β ≥ 0 |
| `political_violence_ord_lag1` | V-Dem 0–4, lagged by 1 year | β ≥ 0 |
| `log_refugee_outflow` | UNHCR log1p | β ≥ 0 |
| `coup_proof_fragmentation` | Pilster-Böhmelt | β ≥ 0 |

**Training set filter:** listwise-drop rows where `cpi_imputed = 1`
(country-year had no real CPI source — WDI/IMF-curated/IMF WEO — and
inflation was region-median imputed). Avoids fitting on synthetic inflation
for PRK, pre-2018 Cuba, Somalia, etc. `cpi_imputed` is preserved as a
data-quality flag in the dashboard, not as a model feature.

**Train/test:** train on 1995–2022; predict unrest in 2023 on holdout 2023.

**v0.4 deletions from v0.3:** within-country z-scoring; `z_ffpi_avg` (global
food-price index identical across countries in a given year — no
cross-sectional variation); the `factor(regime_type)` dummies in the
pooled formula (collinear with civil_liberties).

de Bruin's `counterbalancing` / `politicization` and Fariss / Polity2 remain
loaded in the panel for analysis but are excluded from the fit (Fariss/Polity
correlate ~0.9 with civil_liberties; de Bruin source ends 2010).

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(glmnet)
```

## Pull the labeled panel

```{r}
#| label: pull
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

features <- c(
  "civil_liberties", "log_gdp_per_cap", "loser_consent",
  "conflict_neighbors_count",
  "gdp_growth", "cpi_inflation", "unemployment",
  "political_violence_ord_lag1",
  "log_refugee_outflow", "coup_proof_fragmentation"
)
# v0.4: raw cross-country levels only. No within-country z's. No regime
# dummies in the pooled formula (civil_liberties is the regime-quality
# channel). cpi_inflation is clipped at [-10, 200]% in 08; log_refugee_outflow
# is log1p-transformed in 08. cpi_imputed is pulled separately to listwise-drop
# training rows where the CPI was region-median imputed — it is NOT a model
# feature (would absorb the closedness→under-reporting measurement bias).

needed_cols <- paste(c("iso3c", "year", "regime_type",
                       "outcome_mobilization", "outcome_violence",
                       "cpi_imputed", features), collapse = ", ")
panel <- dbGetQuery(con, sprintf("SELECT %s FROM panel_labeled", needed_cols))
dbDisconnect(con, shutdown = TRUE)

cat("Total panel rows:", nrow(panel), "\n")
cat("Rows with non-NA outcome_mobilization:", sum(!is.na(panel$outcome_mobilization)), "\n")
cat("Rows with non-NA outcome_violence:    ", sum(!is.na(panel$outcome_violence)), "\n")
```

## Filter helpers

Per-outcome complete-cases filter. We keep the same listwise-drop rule
across both tiers: drop rows where any feature is NA, where regime_type
is missing, where the outcome column is NA, OR where `cpi_imputed = 1`
(CPI came from region-median imputation in a closed regime — the closedness
also under-reports the outcome label, so the model would learn a spurious
correlation).

```{r}
#| label: filter-helpers
build_subsets <- function(outcome_col) {
  fc <- complete.cases(panel[, c("regime_type", outcome_col, features)])
  rc <- panel$cpi_imputed == 0
  comp <- panel[fc & rc, ]
  list(
    train   = comp[comp$year <= 2022, ],
    holdout = comp[comp$year == 2023, ]
  )
}

# Inspect each tier's train/holdout shape before fitting.
for (oc in c("outcome_mobilization", "outcome_violence")) {
  s <- build_subsets(oc)
  cat(sprintf("\n%s:\n  train rows = %d (rate %.3f),  holdout rows = %d (rate %.3f)\n",
              oc, nrow(s$train), mean(s$train[[oc]]),
              nrow(s$holdout), mean(s$holdout[[oc]])))
}
```

## Fit setup (shared across both tiers)

```{r}
#| label: fit-setup
regime_labels <- c("0" = "closed_autocracy",
                   "1" = "electoral_autocracy",
                   "2" = "electoral_democracy",
                   "3" = "liberal_democracy")

B_bootstrap <- 500
alpha_fit <- 1
set.seed(20260517)

# Monotonicity constraints in features-vector order.
# -1 = β ≤ 0 required;  +1 = β ≥ 0 required.
feature_signs <- c(-1, -1, -1, +1, -1, +1, +1, +1, +1, +1)
names(feature_signs) <- c("civil_liberties", "log_gdp_per_cap", "loser_consent",
                          "conflict_neighbors_count", "gdp_growth", "cpi_inflation",
                          "unemployment", "political_violence_ord_lag1",
                          "log_refugee_outflow", "coup_proof_fragmentation")
stopifnot(setequal(names(feature_signs), features))

build_limits <- function(design_cols) {
  missing_signs <- setdiff(design_cols, names(feature_signs))
  if (length(missing_signs) > 0) {
    stop("Design-matrix columns without a defined sign: ",
         paste(missing_signs, collapse = ", "))
  }
  signs <- as.numeric(feature_signs[design_cols])
  lower <- rep(-Inf, length(signs)); lower[signs == +1] <- 0
  upper <- rep( Inf, length(signs)); upper[signs == -1] <- 0
  list(lower = lower, upper = upper)
}

# Fit one stratum or pool: returns the same list shape regardless of which.
fit_lasso <- function(x, y, B) {
  lims <- build_limits(colnames(x))
  n_events <- sum(y == 1)
  nfolds   <- if (n_events < 30) 5 else 10

  fit_cv      <- cv.glmnet(x, y, family = "binomial",
                           alpha = alpha_fit, nfolds = nfolds,
                           lower.limits = lims$lower,
                           upper.limits = lims$upper)
  lambda_best <- fit_cv$lambda.min

  coef_full  <- as.numeric(coef(fit_cv, s = "lambda.min"))
  coef_names <- rownames(coef(fit_cv, s = "lambda.min"))
  names(coef_full) <- coef_names

  idx_pos <- which(y == 1); idx_neg <- which(y == 0)
  boot_coefs <- matrix(NA_real_, nrow = B, ncol = length(coef_full),
                       dimnames = list(NULL, coef_names))
  for (b in seq_len(B)) {
    rows <- c(sample(idx_pos, length(idx_pos), replace = TRUE),
              sample(idx_neg, length(idx_neg), replace = TRUE))
    fit_b <- glmnet(x[rows, , drop = FALSE], y[rows],
                    family = "binomial", alpha = alpha_fit,
                    lambda = lambda_best,
                    lower.limits = lims$lower,
                    upper.limits = lims$upper)
    boot_coefs[b, ] <- as.numeric(coef(fit_b))
  }

  list(
    fit_cv        = fit_cv,
    lambda        = lambda_best,
    feature_order = coef_names,
    coefficients  = coef_full,
    coef_mean     = colMeans(boot_coefs),
    vcov          = cov(boot_coefs) + diag(1e-8, length(coef_full)),
    n_train       = nrow(x),
    n_events      = n_events,
    nfolds        = nfolds,
    n_bootstrap   = B
  )
}

# Fit one full tier: per-stratum models + pooled. Returns a list with both
# plus train/holdout metadata.
fit_tier <- function(outcome_col) {
  subs   <- build_subsets(outcome_col)
  train  <- subs$train

  cat(sprintf("\n=== Fitting tier: %s ===\n", outcome_col))
  cat(sprintf("  train rows: %d, events: %d (rate %.3f)\n",
              nrow(train), sum(train[[outcome_col]]), mean(train[[outcome_col]])))

  form <- as.formula(paste(outcome_col, "~", paste(features, collapse = " + ")))

  # Per-stratum fits
  models <- list()
  fit_summary <- data.frame()
  for (k in 0:3) {
    stratum <- train[train$regime_type == k, ]
    if (nrow(stratum) < 30) {
      cat(sprintf("  Regime %d (%s): only %d rows — skipping\n",
                  k, regime_labels[as.character(k)], nrow(stratum)))
      next
    }
    mm <- model.matrix(form, stratum)
    x  <- mm[, -1, drop = FALSE]
    y  <- stratum[[outcome_col]]
    fit_k <- fit_lasso(x, y, B_bootstrap)
    models[[as.character(k)]] <- fit_k

    dev_idx       <- which.min(abs(fit_k$fit_cv$lambda - fit_k$lambda))
    null_dev      <- fit_k$fit_cv$glmnet.fit$nulldev
    dev_at_lambda <- (1 - fit_k$fit_cv$glmnet.fit$dev.ratio[dev_idx]) * null_dev
    fit_summary <- rbind(fit_summary, data.frame(
      tier      = outcome_col,
      regime    = regime_labels[as.character(k)],
      n_obs     = fit_k$n_train,
      n_events  = fit_k$n_events,
      lambda    = signif(fit_k$lambda, 3),
      pseudo_r2 = round(1 - dev_at_lambda / null_dev, 3)
    ))
  }

  # Pooled fit
  mm_pool <- model.matrix(form, train)
  x_pool  <- mm_pool[, -1, drop = FALSE]
  y_pool  <- train[[outcome_col]]
  pooled  <- fit_lasso(x_pool, y_pool, B_bootstrap)
  pooled$formula <- form

  cat("  Pooled coefficients:\n")
  for (nm in names(pooled$coefficients)) {
    cat(sprintf("    %-32s %+.4f\n", nm, pooled$coefficients[[nm]]))
  }

  list(
    pooled_model  = pooled,
    models        = models,
    fit_summary   = fit_summary,
    formula       = form,
    pool_formula  = form,
    train_years   = c(min(train$year), max(train$year)),
    n_train       = nrow(train),
    n_events      = sum(y_pool == 1)
  )
}
```

## Fit both tiers

```{r}
#| label: fit-both
tier_mobilization <- fit_tier("outcome_mobilization")
tier_violence     <- fit_tier("outcome_violence")
```

## Per-tier coefficient comparison

```{r}
#| label: coef-comparison
coef_compare <- data.frame(
  feature              = features,
  mobilization_beta    = round(tier_mobilization$pooled_model$coefficients[features], 4),
  violence_beta        = round(tier_violence$pooled_model$coefficients[features], 4)
)
coef_compare$mobilization_beta[is.na(coef_compare$mobilization_beta)] <- 0
coef_compare$violence_beta[is.na(coef_compare$violence_beta)]         <- 0
coef_compare
```

## Save fitted artifacts to disk

`fit.rds` now stores a `tiers` list — each entry is one outcome tier's full
fit (pooled + stratified + metadata). Calibration (`11`) and export (`12`)
loop over the tiers.

```{r}
#| label: persist
dir.create("../data/model", showWarnings = FALSE, recursive = TRUE)
saveRDS(
  list(
    tiers = list(
      mobilization = tier_mobilization,
      violence     = tier_violence
    ),
    features      = features,
    regime_labels = regime_labels,
    model_version = "unrest-risk-v0.6-two-tier",
    n_bootstrap   = B_bootstrap,
    fitted_at     = Sys.time()
  ),
  "../data/model/fit.rds"
)
cat("Saved data/model/fit.rds with tiers: ",
    paste(c("mobilization", "violence"), collapse = ", "), "\n")
```
