---
title: "10 — Fit the regime-stratified logistic regression"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Four separate logistic regressions — one per V-Dem RoW regime category —
predicting `unrest_next_year` from six z-scored covariates. Coefficients and
the variance-covariance matrix are saved so the dashboard can reproduce
predictions *and* compute 90% CIs by sampling from `MVN(coef, vcov)`.

**Features (7):**

*Absolute-level (raw V-Dem indices)* — "where this country sits structurally":
- `civil_liberties` — V-Dem 0–1 index
- `freedom_expression` — V-Dem 0–1 index

*Within-country z-scored* — "how stressed vs this country's own history":
- `z_gdp_growth` — economic shock
- `z_refugee_outflow` — population-leaving shock
- `z_political_violence_ord` — recent violence vs own norm
- `z_ffpi_avg` — global food-price pressure
- `z_coup_proof_fragmentation` — Pilster-Böhmelt effective-number of forces

Z-scores are winsorized at ±3 SDs (in `08_build_panel.qmd`) so countries
with tiny historical baselines (e.g., UK's near-zero refugee outflow) don't
blow up when a small absolute number arrives.

The hybrid (levels for structural V-Dem signals, z-scores for dynamic
economic/conflict signals) captures both "where this country sits" and "how
it's moving." With pure within-country z-scoring, UK and Cuba look similar
on civil-liberties because each one's z-score is computed against its own
baseline — the raw levels are dropped. The level features fix that.

de Bruin's `counterbalancing` and `politicization` are loaded into the panel
but excluded from the v0 fit — coverage gaps cause excessive listwise drop.

**Train/test:** train on 1995–2022; predict unrest in 2023+ on holdout 2023.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
```

## Pull the labeled panel

```{r}
#| label: pull
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")

features <- c(
  # Absolute level - structural "where this country sits"
  # (dropped freedom_expression - too collinear with civil_liberties; the
  # joint fit produced sign-flipped coefficients on one of them. V-Dem's
  # core democracy indices correlate ~0.85+ cross-sectionally so we keep
  # just one.)
  "civil_liberties",
  # Within-country z-scores - "how stressed vs this country's own history"
  "z_gdp_growth", "z_refugee_outflow", "z_political_violence_ord",
  "z_ffpi_avg", "z_coup_proof_fragmentation"
)

needed_cols <- paste(c("iso3c", "year", "regime_type", "unrest_next_year", features), collapse = ", ")
panel <- dbGetQuery(con, sprintf("SELECT %s FROM panel_labeled", needed_cols))
dbDisconnect(con, shutdown = TRUE)

cat("Total panel rows:", nrow(panel), "\n")
cat("Rows with non-NA outcome:", sum(!is.na(panel$unrest_next_year)), "\n")
```

## Filter to complete cases for fitting

The model needs every feature present plus a valid outcome and regime
category. Listwise deletion is the simplest reasonable default for v0.

```{r}
#| label: filter
complete <- panel[complete.cases(panel[, c("regime_type", "unrest_next_year", features)]), ]
train <- complete[complete$year <= 2022, ]
holdout <- complete[complete$year == 2023, ]

cat("Train rows: ", nrow(train),
    "  | unrest rate:", round(mean(train$unrest_next_year), 3), "\n")
cat("Holdout rows:", nrow(holdout),
    "  | unrest rate:", round(mean(holdout$unrest_next_year), 3), "\n")
```

## Fit one model per regime stratum

```{r}
#| label: fit
formula_str <- paste("unrest_next_year ~", paste(features, collapse = " + "))
form <- as.formula(formula_str)

regime_labels <- c("0" = "closed_autocracy",
                   "1" = "electoral_autocracy",
                   "2" = "electoral_democracy",
                   "3" = "liberal_democracy")

models <- list()
fit_summary <- data.frame()

for (k in 0:3) {
  stratum <- train[train$regime_type == k, ]
  if (nrow(stratum) < 30) {
    cat("Regime", k, "(", regime_labels[as.character(k)], "): only",
        nrow(stratum), "rows — skipping\n")
    next
  }
  fit <- glm(form, data = stratum, family = binomial(link = "logit"))
  models[[as.character(k)]] <- fit
  fit_summary <- rbind(fit_summary, data.frame(
    regime    = regime_labels[as.character(k)],
    n_obs     = nrow(stratum),
    n_unrest  = sum(stratum$unrest_next_year),
    deviance  = round(fit$deviance, 1),
    null_dev  = round(fit$null.deviance, 1),
    pseudo_r2 = round(1 - fit$deviance / fit$null.deviance, 3)
  ))
}
fit_summary
```

## Inspect coefficients (closed autocracy)

```{r}
#| label: coef-autocracy
ca <- models[["0"]]
co <- summary(ca)$coefficients
rownames(co) <- gsub("^z_", "", rownames(co))
round(co, 3)
```

## Inspect coefficients (electoral autocracy)

```{r}
#| label: coef-elect-auto
ea <- models[["1"]]
co <- summary(ea)$coefficients
rownames(co) <- gsub("^z_", "", rownames(co))
round(co, 3)
```

## Save fitted artifacts to disk for next notebooks

We pickle the fitted models so `11_calibration.qmd` and `12_export_artifacts.qmd`
can load them without re-fitting.

```{r}
#| label: persist
dir.create("../data/model", showWarnings = FALSE, recursive = TRUE)
saveRDS(
  list(
    models       = models,
    features     = features,
    regime_labels = regime_labels,
    train_years  = c(min(train$year), max(train$year)),
    n_train      = nrow(train),
    fitted_at    = Sys.time()
  ),
  "../data/model/fit.rds"
)
cat("Saved to data/model/fit.rds\n")
```
