---
title: "11 — Holdout calibration (both tiers)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note title="For students — what this notebook teaches"}
**Why it exists.** A model with high accuracy on its training data is
worthless if it doesn't generalize. This notebook evaluates the
trained model on the **2023 holdout** — data the model never saw
during fitting — and computes how well its predicted probabilities
match observed outcomes. This is what tells us whether the model is
actually useful.

**Data-analytics concepts shown here.**

- *Holdout evaluation*. The cardinal rule of predictive modeling.
  Train on 1995-2022, test on 2023. Never let test data influence
  model fitting in any way.
- *Brier score*. A scoring rule for probabilistic forecasts.
  Mathematically: average of `(predicted_prob - actual)²`. Lower is
  better. Penalizes both over-confidence (predicting 0.9 when the
  event doesn't happen) and under-confidence (predicting 0.1 when it
  does).
- *Brier skill score (BSS)*. `1 − (your Brier) / (baseline Brier)`,
  where baseline = "always predict the base rate." Positive BSS =
  your model beats the no-features baseline. Zero = your model is as
  good as the baseline. Negative = worse than the baseline (your
  model is actively misleading). v0.7.1's mobilization tier achieved
  +0.022 — first time positive in the project's history.
- *Reliability diagrams*. A plot showing whether predicted
  probabilities match observed frequencies. If you predict 30% for
  a bin of country-years, you want ~30% of them to actually have the
  event. Deviations from the diagonal show miscalibration.
- *Why probabilistic forecasts > point predictions*. A binary "yes/no"
  forecast can't distinguish "I'm 51% confident" from "I'm 99%
  confident." Probabilities communicate uncertainty. Brier scoring
  rewards well-calibrated probabilities, not just correct binary calls.

**Domain terms used here.**

- **Brier score** — mean squared error of probabilities vs. binary
  outcomes. Lower is better.
- **Brier skill score (BSS)** — Brier score relative to a baseline.
  Higher is better.
- **Log loss / cross-entropy** — another probabilistic scoring rule.
  Penalizes confident-and-wrong predictions more heavily than Brier.
- **Reliability** (a.k.a. calibration) — does a "70% predicted"
  actually fire 70% of the time? Strong models are well-calibrated.
- **Resolution** — does the model discriminate between event and
  non-event country-years? Trivially perfect calibration ("always
  predict the base rate") has zero resolution.
- **Holdout / test set** — data set aside, not used for training or
  hyperparameter tuning, only used for final evaluation.

**What success looks like.** A table with mobilization BSS and
violence BSS, plus reliability diagrams that hopefully fall close to
the 45-degree diagonal. v0.7.1 shipped with mobilization BSS = +0.022
and violence BSS = +0.624 — published in the dashboard so users can
judge the model honestly.
:::

Score the 2023 holdout for each tier (mobilization, violence) separately.
Each tier gets its own Brier / log loss / reliability bins. The dashboard
displays both calibration metrics so users can judge each prediction.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(ggplot2)
library(glmnet)
source("_helpers.R")
```

## Load fitted model + holdout panel

```{r}
#| label: load
fit      <- readRDS("../data/model/fit.rds")
features <- fit$features
tiers    <- fit$tiers

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
needed_cols <- paste(c("iso3c", "year", "regime_type",
                       "outcome_mobilization", "outcome_violence", features),
                     collapse = ", ")
panel <- dbGetQuery(con, sprintf("SELECT %s FROM panel_labeled", needed_cols))
dbDisconnect(con, shutdown = TRUE)
```

## Per-tier scoring helper

```{r}
#| label: scorer
score_tier <- function(tier, outcome_col) {
  pooled_model <- tier$pooled_model
  pool_form    <- tier$pool_formula

  holdout <- panel[
    complete.cases(panel[, c("regime_type", outcome_col, features)]) &
    panel$year == 2023, ]

  mm_pool_h <- model.matrix(pool_form, holdout)
  x_pool_h  <- mm_pool_h[, -1, drop = FALSE]
  holdout$pred <- as.numeric(predict(pooled_model$fit_cv, newx = x_pool_h,
                                     s = "lambda.min", type = "response"))
  holdout <- holdout[!is.na(holdout$pred), ]

  y <- holdout[[outcome_col]]
  brier          <- mean((holdout$pred - y)^2)
  log_loss       <- -mean(y * log(pmax(holdout$pred, 1e-9)) +
                          (1 - y) * log(pmax(1 - holdout$pred, 1e-9)))
  base_rate      <- mean(y)
  brier_baseline <- mean((base_rate - y)^2)

  breaks <- seq(0, 1, by = 0.1)
  holdout$bin <- cut(holdout$pred, breaks = breaks, include.lowest = TRUE, right = FALSE)
  reliability <- do.call(rbind, lapply(levels(holdout$bin), function(b) {
    rows <- holdout[holdout$bin == b, ]
    if (nrow(rows) == 0) return(NULL)
    data.frame(bin = b,
               predicted = mean(rows$pred),
               actual    = mean(rows[[outcome_col]]),
               n         = nrow(rows))
  }))

  list(
    outcome_col       = outcome_col,
    holdout_year      = 2023,
    n_holdout         = nrow(holdout),
    brier             = brier,
    brier_baseline    = brier_baseline,
    brier_skill_score = 1 - brier / brier_baseline,
    log_loss          = log_loss,
    base_rate         = base_rate,
    mean_pred         = mean(holdout$pred),
    reliability       = list(
      bin_centers = seq(0.05, 0.95, by = 0.1),
      predicted   = reliability$predicted,
      actual      = reliability$actual,
      n           = reliability$n
    )
  )
}

mob_cal <- score_tier(tiers$mobilization, "outcome_mobilization")
vio_cal <- score_tier(tiers$violence,     "outcome_violence")
```

## Calibration summary

```{r}
#| label: summary
summary_tbl <- data.frame(
  tier              = c("mobilization", "violence"),
  n_holdout         = c(mob_cal$n_holdout, vio_cal$n_holdout),
  base_rate         = round(c(mob_cal$base_rate, vio_cal$base_rate), 3),
  mean_pred         = round(c(mob_cal$mean_pred, vio_cal$mean_pred), 3),
  brier             = round(c(mob_cal$brier, vio_cal$brier), 4),
  brier_baseline    = round(c(mob_cal$brier_baseline, vio_cal$brier_baseline), 4),
  brier_skill_score = round(c(mob_cal$brier_skill_score, vio_cal$brier_skill_score), 3),
  log_loss          = round(c(mob_cal$log_loss, vio_cal$log_loss), 4)
)
print(summary_tbl)
```

## Reliability — mobilization tier

```{r}
#| label: reliability-mob
#| fig-cap: "Reliability diagram (mobilization): predicted vs. actual rate by decile"
#| fig-width: 6
#| fig-height: 6
mob_rel <- data.frame(predicted = mob_cal$reliability$predicted,
                      actual    = mob_cal$reliability$actual,
                      n         = mob_cal$reliability$n)
ggplot(mob_rel, aes(x = predicted, y = actual)) +
  geom_abline(slope = 1, intercept = 0, color = "#A89880", linetype = "dashed") +
  geom_line(color = srs_palette$accent, linewidth = 0.8) +
  geom_point(aes(size = n), color = srs_palette$accent) +
  scale_size_continuous(range = c(2, 8), name = "n in bin") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted probability", y = "Empirical mobilization rate",
       caption = "Diagonal = perfect calibration") +
  srs_theme()
```

## Reliability — violence tier

```{r}
#| label: reliability-vio
#| fig-cap: "Reliability diagram (violence): predicted vs. actual rate by decile"
#| fig-width: 6
#| fig-height: 6
vio_rel <- data.frame(predicted = vio_cal$reliability$predicted,
                      actual    = vio_cal$reliability$actual,
                      n         = vio_cal$reliability$n)
ggplot(vio_rel, aes(x = predicted, y = actual)) +
  geom_abline(slope = 1, intercept = 0, color = "#A89880", linetype = "dashed") +
  geom_line(color = srs_palette$crisis, linewidth = 0.8) +
  geom_point(aes(size = n), color = srs_palette$crisis) +
  scale_size_continuous(range = c(2, 8), name = "n in bin") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted probability", y = "Empirical violence rate",
       caption = "Diagonal = perfect calibration") +
  srs_theme()
```

## Persist calibration results

## Time-blocked cross-validation across multiple holdout years

A single-year holdout (n=144) has substantial sampling variance — the BSS
point estimate above could be ±0.04 to ±0.06 by chance. To get an honest
picture, we re-fit the model 6 times: for each holdout outcome year in
2019-2024, train on every panel year before it and score that one year.
Report the distribution of BSS across years, not just a point estimate.

The final shipped model (in `fit.rds` / `weights.json`) is still the one
fit on panel year ≤ 2022 — this CV is purely for honest calibration
reporting.

```{r}
#| label: time-blocked-cv-setup
library(glmnet)

# Re-pull panel with the features list (same as 10_fit_model.qmd)
features <- fit$features

# Pull a clean panel for the CV runs. cpi_imputed = 0 listwise drop matches 10.
con2 <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
needed <- paste(c("iso3c", "year", "regime_type",
                  "outcome_mobilization", "outcome_violence",
                  "cpi_imputed", "manual_label", features), collapse = ", ")
panel_cv <- dbGetQuery(con2, sprintf("SELECT %s FROM panel_labeled", needed))
dbDisconnect(con2, shutdown = TRUE)

# Apply the same missing-indicator zero-fill as 10
panel_cv$acled_covered_lag1[is.na(panel_cv$acled_covered_lag1)] <- 0L
panel_cv$log_acled_violent_demo_fat_per_cap_lag1[
  is.na(panel_cv$log_acled_violent_demo_fat_per_cap_lag1)] <- 0
panel_cv$log_acled_pv_fat_per_cap_lag1[
  is.na(panel_cv$log_acled_pv_fat_per_cap_lag1)] <- 0

# Same monotonicity constraints as 10_fit_model.qmd
feature_signs <- c(-1, -1, -1, +1, -1, +1, +1, +1, +1, +1, +1, +1, +1, 0)
names(feature_signs) <- features

build_limits_cv <- function(design_cols) {
  signs <- as.numeric(feature_signs[design_cols])
  lower <- rep(-Inf, length(signs)); lower[signs == +1] <- 0
  upper <- rep( Inf, length(signs)); upper[signs == -1] <- 0
  list(lower = lower, upper = upper)
}

# Refit + score one fold
score_fold <- function(outcome_col, holdout_outcome_year) {
  # Panel year T predicts outcome year T+1.
  # Holdout outcome year H => holdout panel year = H - 1.
  # Train: panel year < H - 1, with manual_label ride-alongs.
  hp <- holdout_outcome_year - 1L

  fc <- complete.cases(panel_cv[, c("regime_type", outcome_col, features)])
  rc <- panel_cv$cpi_imputed == 0
  comp <- panel_cv[fc & rc, ]

  train <- comp[(comp$year < hp) | (comp$manual_label == 1 & comp$year != hp), ]
  test  <- comp[comp$year == hp, ]

  if (nrow(test) < 30 || sum(test[[outcome_col]] == 1) < 3) {
    return(data.frame(holdout_outcome_year = holdout_outcome_year,
                      n_holdout = nrow(test), n_events = sum(test[[outcome_col]] == 1),
                      brier = NA, brier_baseline = NA, brier_skill_score = NA,
                      base_rate = NA, mean_pred = NA))
  }

  form <- as.formula(paste(outcome_col, "~", paste(features, collapse = " + ")))
  mm_tr <- model.matrix(form, train); x_tr <- mm_tr[, -1, drop = FALSE]
  mm_te <- model.matrix(form, test);  x_te <- mm_te[, -1, drop = FALSE]
  y_tr  <- train[[outcome_col]]
  y_te  <- test[[outcome_col]]

  lims <- build_limits_cv(colnames(x_tr))
  n_events_tr <- sum(y_tr == 1)
  nfolds <- if (n_events_tr < 30) 5 else 10
  fit_cv <- cv.glmnet(x_tr, y_tr, family = "binomial",
                      alpha = 1, nfolds = nfolds,
                      lower.limits = lims$lower, upper.limits = lims$upper)
  preds <- as.numeric(predict(fit_cv, newx = x_te, s = "lambda.min", type = "response"))

  base <- mean(y_tr)
  brier <- mean((preds - y_te)^2)
  brier_baseline <- mean((base - y_te)^2)
  data.frame(holdout_outcome_year = holdout_outcome_year,
             n_holdout = nrow(test), n_events = sum(y_te == 1),
             brier = round(brier, 4),
             brier_baseline = round(brier_baseline, 4),
             brier_skill_score = round(1 - brier / brier_baseline, 4),
             base_rate = round(base, 4),
             mean_pred = round(mean(preds), 4))
}
```

```{r}
#| label: time-blocked-cv-run
holdout_years <- 2019:2024
cv_mob <- do.call(rbind, lapply(holdout_years, function(y) score_fold("outcome_mobilization", y)))
cv_vio <- do.call(rbind, lapply(holdout_years, function(y) score_fold("outcome_violence",     y)))

cat("\n=== Time-blocked CV — mobilization tier ===\n")
print(cv_mob)
cat("\nMobilization BSS distribution: median =",
    round(median(cv_mob$brier_skill_score, na.rm = TRUE), 4),
    "  IQR =",
    paste(round(quantile(cv_mob$brier_skill_score,
                         c(0.25, 0.75), na.rm = TRUE), 4), collapse = "  "), "\n")

cat("\n=== Time-blocked CV — violence tier ===\n")
print(cv_vio)
cat("\nViolence BSS distribution: median =",
    round(median(cv_vio$brier_skill_score, na.rm = TRUE), 4),
    "  IQR =",
    paste(round(quantile(cv_vio$brier_skill_score,
                         c(0.25, 0.75), na.rm = TRUE), 4), collapse = "  "), "\n")
```

## Persist calibration results

```{r}
#| label: persist
saveRDS(list(
  mobilization = mob_cal,
  violence     = vio_cal,
  cv_mobilization = cv_mob,
  cv_violence     = cv_vio
), "../data/model/calibration.rds")
cat("Saved data/model/calibration.rds with both tiers + CV distributions\n")
```
