---
title: "11 — Holdout calibration (both tiers)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Score the 2023 holdout for each tier (mobilization, violence) separately.
Each tier gets its own Brier / log loss / reliability bins. The dashboard
displays both calibration metrics so users can judge each prediction.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(ggplot2)
library(glmnet)
source("_helpers.R")
```

## Load fitted model + holdout panel

```{r}
#| label: load
fit      <- readRDS("../data/model/fit.rds")
features <- fit$features
tiers    <- fit$tiers

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
needed_cols <- paste(c("iso3c", "year", "regime_type",
                       "outcome_mobilization", "outcome_violence", features),
                     collapse = ", ")
panel <- dbGetQuery(con, sprintf("SELECT %s FROM panel_labeled", needed_cols))
dbDisconnect(con, shutdown = TRUE)
```

## Per-tier scoring helper

```{r}
#| label: scorer
score_tier <- function(tier, outcome_col) {
  pooled_model <- tier$pooled_model
  pool_form    <- tier$pool_formula

  holdout <- panel[
    complete.cases(panel[, c("regime_type", outcome_col, features)]) &
    panel$year == 2023, ]

  mm_pool_h <- model.matrix(pool_form, holdout)
  x_pool_h  <- mm_pool_h[, -1, drop = FALSE]
  holdout$pred <- as.numeric(predict(pooled_model$fit_cv, newx = x_pool_h,
                                     s = "lambda.min", type = "response"))
  holdout <- holdout[!is.na(holdout$pred), ]

  y <- holdout[[outcome_col]]
  brier          <- mean((holdout$pred - y)^2)
  log_loss       <- -mean(y * log(pmax(holdout$pred, 1e-9)) +
                          (1 - y) * log(pmax(1 - holdout$pred, 1e-9)))
  base_rate      <- mean(y)
  brier_baseline <- mean((base_rate - y)^2)

  breaks <- seq(0, 1, by = 0.1)
  holdout$bin <- cut(holdout$pred, breaks = breaks, include.lowest = TRUE, right = FALSE)
  reliability <- do.call(rbind, lapply(levels(holdout$bin), function(b) {
    rows <- holdout[holdout$bin == b, ]
    if (nrow(rows) == 0) return(NULL)
    data.frame(bin = b,
               predicted = mean(rows$pred),
               actual    = mean(rows[[outcome_col]]),
               n         = nrow(rows))
  }))

  list(
    outcome_col       = outcome_col,
    holdout_year      = 2023,
    n_holdout         = nrow(holdout),
    brier             = brier,
    brier_baseline    = brier_baseline,
    brier_skill_score = 1 - brier / brier_baseline,
    log_loss          = log_loss,
    base_rate         = base_rate,
    mean_pred         = mean(holdout$pred),
    reliability       = list(
      bin_centers = seq(0.05, 0.95, by = 0.1),
      predicted   = reliability$predicted,
      actual      = reliability$actual,
      n           = reliability$n
    )
  )
}

mob_cal <- score_tier(tiers$mobilization, "outcome_mobilization")
vio_cal <- score_tier(tiers$violence,     "outcome_violence")
```

## Calibration summary

```{r}
#| label: summary
summary_tbl <- data.frame(
  tier              = c("mobilization", "violence"),
  n_holdout         = c(mob_cal$n_holdout, vio_cal$n_holdout),
  base_rate         = round(c(mob_cal$base_rate, vio_cal$base_rate), 3),
  mean_pred         = round(c(mob_cal$mean_pred, vio_cal$mean_pred), 3),
  brier             = round(c(mob_cal$brier, vio_cal$brier), 4),
  brier_baseline    = round(c(mob_cal$brier_baseline, vio_cal$brier_baseline), 4),
  brier_skill_score = round(c(mob_cal$brier_skill_score, vio_cal$brier_skill_score), 3),
  log_loss          = round(c(mob_cal$log_loss, vio_cal$log_loss), 4)
)
print(summary_tbl)
```

## Reliability — mobilization tier

```{r}
#| label: reliability-mob
#| fig-cap: "Reliability diagram (mobilization): predicted vs. actual rate by decile"
#| fig-width: 6
#| fig-height: 6
mob_rel <- data.frame(predicted = mob_cal$reliability$predicted,
                      actual    = mob_cal$reliability$actual,
                      n         = mob_cal$reliability$n)
ggplot(mob_rel, aes(x = predicted, y = actual)) +
  geom_abline(slope = 1, intercept = 0, color = "#A89880", linetype = "dashed") +
  geom_line(color = srs_palette$accent, linewidth = 0.8) +
  geom_point(aes(size = n), color = srs_palette$accent) +
  scale_size_continuous(range = c(2, 8), name = "n in bin") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted probability", y = "Empirical mobilization rate",
       caption = "Diagonal = perfect calibration") +
  srs_theme()
```

## Reliability — violence tier

```{r}
#| label: reliability-vio
#| fig-cap: "Reliability diagram (violence): predicted vs. actual rate by decile"
#| fig-width: 6
#| fig-height: 6
vio_rel <- data.frame(predicted = vio_cal$reliability$predicted,
                      actual    = vio_cal$reliability$actual,
                      n         = vio_cal$reliability$n)
ggplot(vio_rel, aes(x = predicted, y = actual)) +
  geom_abline(slope = 1, intercept = 0, color = "#A89880", linetype = "dashed") +
  geom_line(color = srs_palette$crisis, linewidth = 0.8) +
  geom_point(aes(size = n), color = srs_palette$crisis) +
  scale_size_continuous(range = c(2, 8), name = "n in bin") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted probability", y = "Empirical violence rate",
       caption = "Diagonal = perfect calibration") +
  srs_theme()
```

## Persist calibration results

```{r}
#| label: persist
saveRDS(list(mobilization = mob_cal, violence = vio_cal),
        "../data/model/calibration.rds")
cat("Saved data/model/calibration.rds with both tiers\n")
```
