---
title: "11 — Holdout calibration"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Score the 2023 holdout with the fitted model. Compute:

- **Brier score**: mean squared error of probability predictions (lower = better; 0.25 = random for balanced classes)
- **Log loss**: cross-entropy (lower = better; log(2) ≈ 0.69 = random)
- **Reliability**: bin predictions into 10 deciles, compare predicted vs. actual rate per bin

The reliability diagram is what tells us whether "predicted 20%" actually
corresponds to a ~20% empirical rate.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(ggplot2)
library(glmnet)
source("_helpers.R")
```

## Load fitted model + holdout panel

```{r}
#| label: load
fit <- readRDS("../data/model/fit.rds")
models   <- fit$models
features <- fit$features
form     <- fit$formula

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
needed_cols <- paste(c("iso3c", "year", "regime_type", "unrest_next_year", features), collapse = ", ")
panel <- dbGetQuery(con, sprintf("SELECT %s FROM panel_labeled", needed_cols))
dbDisconnect(con, shutdown = TRUE)

holdout <- panel[
  complete.cases(panel[, c("regime_type", "unrest_next_year", features)]) &
  panel$year == 2023, ]
cat("Holdout rows:", nrow(holdout),
    "  | unrest rate:", round(mean(holdout$unrest_next_year), 3), "\n")
```

## Score the holdout

Each row uses the ridge model for its regime stratum. `glmnet`'s `predict()`
wants the design matrix as `newx` (not a `newdata` data.frame), so build it
with the same formula the fit used.

```{r}
#| label: score
holdout$pred <- NA_real_
for (k in names(models)) {
  idx <- holdout$regime_type == as.integer(k)
  if (any(idx)) {
    mm_h <- model.matrix(form, holdout[idx, ])
    x_h  <- mm_h[, -1, drop = FALSE]
    holdout$pred[idx] <- as.numeric(predict(models[[k]]$fit_cv,
                                            newx = x_h,
                                            s    = "lambda.min",
                                            type = "response"))
  }
}
holdout <- holdout[!is.na(holdout$pred), ]
cat("Scored rows:", nrow(holdout), "\n")
```

## Brier and log loss

```{r}
#| label: metrics
brier   <- mean((holdout$pred - holdout$unrest_next_year)^2)
log_loss <- -mean(holdout$unrest_next_year * log(pmax(holdout$pred, 1e-9)) +
                  (1 - holdout$unrest_next_year) * log(pmax(1 - holdout$pred, 1e-9)))
base_rate <- mean(holdout$unrest_next_year)
brier_baseline <- mean((base_rate - holdout$unrest_next_year)^2)

cat("Brier score:        ", round(brier, 4),
    "  (baseline = always-predict-base-rate:", round(brier_baseline, 4), ")\n")
cat("Log loss:           ", round(log_loss, 4), "\n")
cat("Brier skill score:  ", round(1 - brier / brier_baseline, 3),
    "  (positive = better than baseline)\n")
```

## Reliability bins

Bucket predictions into deciles and compare predicted vs actual unrest rate
within each bin. A perfectly calibrated model: predicted ≈ actual.

```{r}
#| label: reliability
breaks <- seq(0, 1, by = 0.1)
holdout$bin <- cut(holdout$pred, breaks = breaks, include.lowest = TRUE, right = FALSE)

reliability <- do.call(rbind, lapply(levels(holdout$bin), function(b) {
  rows <- holdout[holdout$bin == b, ]
  if (nrow(rows) == 0) return(NULL)
  data.frame(
    bin       = b,
    predicted = mean(rows$pred),
    actual    = mean(rows$unrest_next_year),
    n         = nrow(rows)
  )
}))
reliability
```

```{r}
#| label: reliability-plot
#| fig-cap: "Reliability diagram on 2023 holdout (predicted vs. actual unrest rate by decile)"
#| fig-width: 6
#| fig-height: 6
ggplot(reliability, aes(x = predicted, y = actual)) +
  geom_abline(slope = 1, intercept = 0, color = "#A89880", linetype = "dashed") +
  geom_line(color = srs_palette$accent, linewidth = 0.8) +
  geom_point(aes(size = n), color = srs_palette$accent) +
  scale_size_continuous(range = c(2, 8), name = "n in bin") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Mean predicted probability", y = "Empirical unrest rate",
       caption = "Diagonal = perfect calibration") +
  srs_theme()
```

## Persist calibration results

```{r}
#| label: persist
calibration <- list(
  holdout_year = 2023,
  n_holdout    = nrow(holdout),
  brier        = brier,
  brier_baseline = brier_baseline,
  brier_skill_score = 1 - brier / brier_baseline,
  log_loss     = log_loss,
  base_rate    = base_rate,
  reliability  = list(
    bin_centers = seq(0.05, 0.95, by = 0.1),
    predicted   = reliability$predicted,
    actual      = reliability$actual,
    n           = reliability$n
  )
)
saveRDS(calibration, "../data/model/calibration.rds")
cat("Saved data/model/calibration.rds\n")
```
