---
title: "12 — Export frozen artifacts for the dashboard"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Writes four JSON files to `../artifacts/` that the SvelteKit dashboard
consumes. Math runs in the browser — no backend at runtime.

- **`weights.json`** — coefficients + vcov per regime stratum; feature list
- **`base_rates.json`** — per-regime base rates from training data
- **`country_snapshot.json`** — per-country: regime, latest raw values,
  within-country mean+sd (for slider math), predicted probability + 90% CI,
  and per-feature contribution
- **`calibration.json`** — Brier, log loss, reliability bins

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(jsonlite)
```

## Load fitted model + panel

```{r}
#| label: load
fit <- readRDS("../data/model/fit.rds")
models   <- fit$models
features <- fit$features
regime_labels <- fit$regime_labels
calibration_obj <- readRDS("../data/model/calibration.rds")

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
panel <- dbGetQuery(con, "SELECT * FROM panel_labeled")
dbDisconnect(con, shutdown = TRUE)
```

## Build weights.json

For each fitted stratum, store the coefficients and the variance-covariance
matrix. The dashboard samples from `MVN(coef, vcov)` to compute 90% CIs on
predictions.

```{r}
#| label: weights
strata <- list()
for (k in names(models)) {
  fit_k <- models[[k]]
  co <- coef(fit_k)
  vc <- vcov(fit_k)
  strata[[regime_labels[[k]]]] <- list(
    regime_code   = as.integer(k),
    coefficients  = as.list(co),
    vcov          = unname(as.matrix(vc)),
    feature_order = names(co),
    n_train       = length(fit_k$y),
    deviance      = unname(fit_k$deviance),
    null_deviance = unname(fit_k$null.deviance)
  )
}

# IMPORTANT: pass named vectors as `as.list()` so jsonlite emits JSON objects
# with the names as keys. Plain named vectors get unboxed to arrays and lose
# their names, which breaks the dashboard's `weights.feature_labels[z_name]`
# lookups (they'd return undefined).
weights <- list(
  model_version = "unrest-risk-v0.1",
  fitted_at     = format(fit$fitted_at, "%Y-%m-%d"),
  train_years   = fit$train_years,
  features      = features,
  feature_labels = as.list(c(
    civil_liberties            = "Civil liberties",
    z_gdp_growth               = "GDP growth",
    z_cpi_inflation            = "Inflation",
    z_unemployment             = "Unemployment",
    z_political_violence_ord   = "Political violence",
    z_refugee_outflow          = "Refugee outflow",
    z_ffpi_avg                 = "Global food prices",
    z_coup_proof_fragmentation = "Coup-proofing fragmentation"
  )),
  feature_direction = as.list(c(   # +1 = higher value raises risk, -1 = lowers
    civil_liberties            = -1,
    z_gdp_growth               = -1,
    z_cpi_inflation            = +1,
    z_unemployment             = +1,
    z_political_violence_ord   = +1,
    z_refugee_outflow          = +1,
    z_ffpi_avg                 = +1,
    z_coup_proof_fragmentation = -1
  )),
  strata = strata
)

dir.create("../artifacts", showWarnings = FALSE, recursive = TRUE)
write_json(weights, "../artifacts/weights.json", auto_unbox = TRUE, pretty = TRUE, digits = 6)
cat("Wrote artifacts/weights.json\n")
```

## Build base_rates.json

```{r}
#| label: base-rates
features_check <- c("regime_type", "unrest_next_year", features)
train_idx <- complete.cases(panel[, features_check]) & panel$year <= 2022
train <- panel[train_idx, ]

base_rates <- list()
for (k in 0:3) {
  stratum <- train[train$regime_type == k, ]
  if (nrow(stratum) == 0) next
  base_rates[[regime_labels[[as.character(k)]]]] <- list(
    regime_code = k,
    n_obs       = nrow(stratum),
    n_unrest    = sum(stratum$unrest_next_year),
    base_rate   = round(mean(stratum$unrest_next_year), 4)
  )
}

write_json(list(
  model_version = "unrest-risk-v0.1",
  train_years   = fit$train_years,
  regimes       = base_rates
), "../artifacts/base_rates.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/base_rates.json\n")
```

## Build country_snapshot.json

For each country, take its **latest** year (2023 in v0). Pull raw values
and within-country mean/sd so the dashboard can recompute z-scores when
the user moves a slider. Then score with the regime's model and bootstrap
a 90% CI by sampling 1,000 times from `MVN(coef, vcov)`.

```{r}
#| label: snapshot
# Map z-scored features back to their raw counterparts so the dashboard can
# work with intelligible values (e.g. show "GDP growth = -10.95%", not "z = -2.3").
raw_features <- c(
  civil_liberties            = "civil_liberties",
  z_gdp_growth               = "gdp_growth",
  z_cpi_inflation            = "cpi_inflation",
  z_unemployment             = "unemployment",
  z_political_violence_ord   = "political_violence_ord",
  z_refugee_outflow          = "refugee_outflow",
  z_ffpi_avg                 = "ffpi_avg",
  z_coup_proof_fragmentation = "coup_proof_fragmentation"
)

# Compute within-country mean and sd for each raw feature, across all history.
country_stats <- list()
for (iso in unique(panel$iso3c)) {
  rows <- panel[panel$iso3c == iso, ]
  stats <- list()
  for (z_name in names(raw_features)) {
    raw_name <- raw_features[[z_name]]
    x <- rows[[raw_name]]
    if (sum(!is.na(x)) < 5) next
    stats[[raw_name]] <- list(
      country_mean = mean(x, na.rm = TRUE),
      country_sd   = sd(x, na.rm = TRUE)
    )
  }
  country_stats[[iso]] <- stats
}

# Sampling helper: draw from MVN given coef + vcov, score linearly + sigmoid.
mvn_sample_ci <- function(model, x_named, n_draw = 1000, alpha = 0.10) {
  co <- coef(model); vc <- vcov(model)
  L  <- chol(vc + diag(1e-8, nrow(vc)))  # tiny ridge for numerical safety
  draws <- t(L) %*% matrix(rnorm(length(co) * n_draw), nrow = length(co)) + co
  # design vector with intercept first
  x_full <- c(1, x_named[names(co)[-1]])
  lp <- as.numeric(t(draws) %*% x_full)
  p  <- 1 / (1 + exp(-lp))
  c(point = 1 / (1 + exp(-sum(co * x_full))),
    lo    = unname(quantile(p, alpha / 2, na.rm = TRUE)),
    hi    = unname(quantile(p, 1 - alpha / 2, na.rm = TRUE)))
}

# Build per-country snapshot for the latest year that has complete features.
latest_year <- max(panel$year[!is.na(panel$mass_mobilization_ord)])

snapshot_rows <- list()
for (iso in unique(panel$iso3c)) {
  rows <- panel[panel$iso3c == iso & panel$year <= latest_year, ]
  if (nrow(rows) == 0) next
  # Latest non-NA-features row for this country.
  rows <- rows[order(-rows$year), ]
  good <- rows[complete.cases(rows[, c("regime_type", features)]), ]
  if (nrow(good) == 0) next
  cur <- good[1, ]
  k <- as.character(cur$regime_type)
  if (!k %in% names(models)) next

  x_named <- setNames(as.numeric(cur[, features]), features)
  pred <- mvn_sample_ci(models[[k]], x_named)

  # Contribution = coef * z for each feature (linear-predictor units)
  co <- coef(models[[k]])
  contribs <- co[features] * x_named[features]

  # Raw values for sliders
  raw_vals <- list()
  stats <- country_stats[[iso]]
  for (z_name in features) {
    raw_name <- raw_features[[z_name]]
    raw_vals[[raw_name]] <- list(
      value = unname(cur[[raw_name]]),
      z     = unname(x_named[[z_name]]),
      mean  = if (!is.null(stats[[raw_name]])) unname(stats[[raw_name]]$country_mean) else NA,
      sd    = if (!is.null(stats[[raw_name]])) unname(stats[[raw_name]]$country_sd) else NA,
      contribution = unname(contribs[[z_name]])
    )
  }

  snapshot_rows[[iso]] <- list(
    iso3c        = iso,
    name         = cur$name,
    region       = cur$region,
    subregion    = cur$subregion,
    regime_type  = as.integer(cur$regime_type),
    regime_label = regime_labels[[k]],
    snapshot_year = as.integer(cur$year),
    predicted_probability = unname(pred["point"]),
    ci_lo_90  = unname(pred["lo"]),
    ci_hi_90  = unname(pred["hi"]),
    features  = raw_vals
  )
}

cat("Scored", length(snapshot_rows), "countries for snapshot\n")

write_json(list(
  model_version = "unrest-risk-v0.1",
  snapshot_year = latest_year,
  horizon_months = 12,
  countries     = snapshot_rows
), "../artifacts/country_snapshot.json", auto_unbox = TRUE, pretty = TRUE, digits = 6)
cat("Wrote artifacts/country_snapshot.json\n")
```

## Build calibration.json

```{r}
#| label: calibration
write_json(list(
  model_version = "unrest-risk-v0.1",
  holdout_year  = calibration_obj$holdout_year,
  n_holdout     = calibration_obj$n_holdout,
  brier         = calibration_obj$brier,
  brier_baseline = calibration_obj$brier_baseline,
  brier_skill_score = calibration_obj$brier_skill_score,
  log_loss      = calibration_obj$log_loss,
  base_rate     = calibration_obj$base_rate,
  reliability   = calibration_obj$reliability
), "../artifacts/calibration.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/calibration.json\n")
```

## Summary

```{r}
#| label: final
cat("=== Artifacts in ../artifacts/ ===\n")
for (f in list.files("../artifacts", pattern = "\\.json$")) {
  size <- file.info(file.path("../artifacts", f))$size
  cat(sprintf("  %-30s %8s bytes\n", f, format(size, big.mark = ",")))
}

# Spot-check Cuba (defensive: skip if not in snapshot for any reason)
cat("\n=== Cuba snapshot ===\n")
cuba <- snapshot_rows[["CUB"]]
if (is.null(cuba)) {
  cat("  (Cuba not in snapshot - check data coverage)\n")
} else {
  cat("  Regime:                ", cuba$regime_label, "\n")
  cat("  Snapshot year:         ", cuba$snapshot_year, "\n")
  cat("  Predicted probability: ", round(cuba$predicted_probability, 3), "\n")
  cat("  90% CI:                ", round(cuba$ci_lo_90, 3), "to", round(cuba$ci_hi_90, 3), "\n")
}
```
