---
title: "12 — Export frozen artifacts for the dashboard"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Writes JSON files to `../artifacts/` that the static dashboard consumes.
Math runs in the browser — no backend at runtime.

- **`weights.json`** — coefficients + vcov per tier (mobilization, violence),
  with both pooled and per-regime models
- **`base_rates.json`** — per-regime base rates from training data, per tier
- **`country_snapshot.json`** — per-country: regime, latest raw feature
  values, predicted probability + 90% CI **for each tier**, per-feature
  contribution per tier
- **`calibration.json`** — Brier, log loss, reliability bins per tier

v0.6: two-tier outcome stratification. Each country has two probabilities —
P(mobilization) for protest/demonstration scale and P(violence) for armed
conflict/coup risk.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(jsonlite)
```

## Load fitted model + panel

```{r}
#| label: load
fit             <- readRDS("../data/model/fit.rds")
tiers           <- fit$tiers
features        <- fit$features
regime_labels   <- fit$regime_labels
calibration_obj <- readRDS("../data/model/calibration.rds")

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
panel <- dbGetQuery(con, "SELECT * FROM panel_labeled")
dbDisconnect(con, shutdown = TRUE)
```

## Build weights.json

Two tiers; each tier has pooled + per-regime strata.

```{r}
#| label: weights
feature_labels <- as.list(c(
  civil_liberties                         = "Civil liberties",
  log_gdp_per_cap                         = "GDP per capita",
  loser_consent                           = "Loser consent",
  conflict_neighbors_count                = "Regional conflict",
  gdp_growth                              = "GDP growth",
  cpi_inflation                           = "Inflation",
  unemployment                            = "Unemployment",
  political_violence_ord_lag1             = "Political violence (prior year)",
  mass_mobilization_ord_lag1              = "Mass mobilization (prior year)",
  log_refugee_outflow                     = "Refugee outflow (log)",
  coup_proof_fragmentation                = "Coup-proofing fragmentation",
  log_acled_violent_demo_fat_per_cap_lag1 = "ACLED deadly-protest fatalities (prior year)",
  log_acled_pv_fat_per_cap_lag1           = "ACLED political-violence fatalities (prior year)",
  acled_covered_lag1                      = "ACLED coverage active (prior year)"
))
feature_direction <- as.list(c(   # +1 = higher value raises risk, -1 = lowers, 0 = no constraint
  civil_liberties                         = -1,
  log_gdp_per_cap                         = -1,
  loser_consent                           = -1,
  conflict_neighbors_count                = +1,
  gdp_growth                              = -1,
  cpi_inflation                           = +1,
  unemployment                            = +1,
  political_violence_ord_lag1             = +1,
  mass_mobilization_ord_lag1              = +1,
  log_refugee_outflow                     = +1,
  coup_proof_fragmentation                = +1,
  log_acled_violent_demo_fat_per_cap_lag1 = +1,
  log_acled_pv_fat_per_cap_lag1           = +1,
  acled_covered_lag1                      = 0
))

build_tier_weights <- function(tier) {
  strata <- list()
  for (k in names(tier$models)) {
    fit_k <- tier$models[[k]]
    strata[[regime_labels[[k]]]] <- list(
      regime_code   = as.integer(k),
      coefficients  = as.list(fit_k$coefficients),
      vcov          = unname(as.matrix(fit_k$vcov)),
      feature_order = fit_k$feature_order,
      n_train       = fit_k$n_train,
      n_events      = fit_k$n_events,
      lambda        = fit_k$lambda,
      n_bootstrap   = fit_k$n_bootstrap
    )
  }
  list(
    strata = strata,
    pooled = list(
      coefficients  = as.list(tier$pooled_model$coefficients),
      vcov          = unname(as.matrix(tier$pooled_model$vcov)),
      feature_order = tier$pooled_model$feature_order,
      lambda        = tier$pooled_model$lambda,
      n_train       = tier$pooled_model$n_train,
      n_events      = tier$pooled_model$n_events,
      n_bootstrap   = tier$pooled_model$n_bootstrap
    ),
    train_years = tier$train_years
  )
}

weights <- list(
  model_version    = fit$model_version,
  fitted_at        = format(fit$fitted_at, "%Y-%m-%d"),
  features         = features,
  feature_labels   = feature_labels,
  feature_direction = feature_direction,
  tiers = list(
    mobilization = build_tier_weights(tiers$mobilization),
    violence     = build_tier_weights(tiers$violence)
  )
)

dir.create("../artifacts", showWarnings = FALSE, recursive = TRUE)
write_json(weights, "../artifacts/weights.json",
           auto_unbox = TRUE, pretty = TRUE, digits = 6)
cat("Wrote artifacts/weights.json\n")
```

## Build base_rates.json

```{r}
#| label: base-rates
build_tier_base_rates <- function(outcome_col) {
  features_check <- c("regime_type", outcome_col, features)
  train_idx <- complete.cases(panel[, features_check]) & panel$year <= 2022
  train <- panel[train_idx, ]
  out <- list()
  for (k in 0:3) {
    s <- train[train$regime_type == k, ]
    if (nrow(s) == 0) next
    out[[regime_labels[[as.character(k)]]]] <- list(
      regime_code = k,
      n_obs       = nrow(s),
      n_events    = sum(s[[outcome_col]]),
      base_rate   = round(mean(s[[outcome_col]]), 4)
    )
  }
  out
}

write_json(list(
  model_version = fit$model_version,
  tiers = list(
    mobilization = build_tier_base_rates("outcome_mobilization"),
    violence     = build_tier_base_rates("outcome_violence")
  )
), "../artifacts/base_rates.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/base_rates.json\n")
```

## Build country_snapshot.json

Each country snapshot ships **two predictions** — one per tier — alongside
the raw feature values (same features feed both tiers).

```{r}
#| label: snapshot
# MVN sampling helper.
mvn_sample_ci <- function(model, x_named, n_draw = 1000, alpha = 0.10) {
  co <- model$coefficients
  vc <- model$vcov
  L  <- chol(vc + diag(1e-8, nrow(vc)))
  draws <- t(L) %*% matrix(rnorm(length(co) * n_draw), nrow = length(co)) + co
  x_full <- c(1, x_named[names(co)[-1]])
  lp <- as.numeric(t(draws) %*% x_full)
  p  <- 1 / (1 + exp(-lp))
  c(point = 1 / (1 + exp(-sum(co * x_full))),
    lo    = unname(quantile(p, alpha / 2, na.rm = TRUE)),
    hi    = unname(quantile(p, 1 - alpha / 2, na.rm = TRUE)))
}

# Score one country with one tier — returns point + CI + per-feature contributions.
score_one <- function(tier, x_named) {
  pool_coef <- tier$pooled_model$coefficients
  pool_vec  <- setNames(rep(0, length(pool_coef)), names(pool_coef))
  pool_vec["(Intercept)"] <- 1
  for (f in features) if (f %in% names(pool_vec)) pool_vec[f] <- x_named[[f]]
  pool_lp <- sum(pool_coef * pool_vec)
  pool_p  <- 1 / (1 + exp(-pool_lp))

  L <- chol(tier$pooled_model$vcov + diag(1e-8, nrow(tier$pooled_model$vcov)))
  draws <- t(L) %*% matrix(rnorm(length(pool_coef) * 1000),
                           nrow = length(pool_coef)) + pool_coef
  p_draws <- 1 / (1 + exp(-as.numeric(t(draws) %*% pool_vec)))

  contribs <- sapply(features, function(f) {
    if (f %in% names(pool_coef)) unname(pool_coef[f] * x_named[[f]]) else 0
  })
  names(contribs) <- features

  list(
    point    = unname(pool_p),
    ci_lo_90 = unname(quantile(p_draws, 0.05)),
    ci_hi_90 = unname(quantile(p_draws, 0.95)),
    contribs = contribs
  )
}

latest_year <- max(panel$year[!is.na(panel$mass_mobilization_ord)])

snapshot_rows <- list()
for (iso in unique(panel$iso3c)) {
  rows <- panel[panel$iso3c == iso & panel$year <= latest_year, ]
  if (nrow(rows) == 0) next
  rows <- rows[order(-rows$year), ]
  good <- rows[complete.cases(rows[, c("regime_type", features)]), ]
  if (nrow(good) == 0) next
  cur <- good[1, ]
  k <- as.character(cur$regime_type)

  x_named <- setNames(as.numeric(cur[, features]), features)

  s_mob <- score_one(tiers$mobilization, x_named)
  s_vio <- score_one(tiers$violence,     x_named)

  raw_vals <- list()
  for (f in features) {
    raw_vals[[f]] <- list(
      value                     = unname(cur[[f]]),
      mobilization_contribution = unname(s_mob$contribs[[f]]),
      violence_contribution     = unname(s_vio$contribs[[f]])
    )
  }

  is_imputed <- isTRUE(as.logical(cur$cpi_imputed))

  snapshot_rows[[iso]] <- list(
    iso3c         = iso,
    name          = cur$name,
    region        = cur$region,
    subregion     = cur$subregion,
    regime_type   = as.integer(cur$regime_type),
    regime_label  = regime_labels[[k]],
    snapshot_year = as.integer(cur$year),
    mobilization  = list(
      probability = s_mob$point,
      ci_lo_90    = s_mob$ci_lo_90,
      ci_hi_90    = s_mob$ci_hi_90
    ),
    violence      = list(
      probability = s_vio$point,
      ci_lo_90    = s_vio$ci_lo_90,
      ci_hi_90    = s_vio$ci_hi_90
    ),
    cpi_imputed   = is_imputed,
    features      = raw_vals
  )
}

cat("Scored", length(snapshot_rows), "countries (both tiers)\n")

write_json(list(
  model_version  = fit$model_version,
  snapshot_year  = latest_year,
  horizon_months = 12,
  countries      = snapshot_rows
), "../artifacts/country_snapshot.json",
   auto_unbox = TRUE, pretty = TRUE, digits = 6)
cat("Wrote artifacts/country_snapshot.json\n")
```

## Build hanke_inflation.json (dashboard toggle data)

```{r}
#| label: hanke-export
hanke_raw <- dbGetQuery(
  dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb", read_only = TRUE),
  "SELECT iso3c, year, cpi_inflation_hanke FROM hanke_inflation"
)
hanke_latest <- list()
for (iso in unique(hanke_raw$iso3c)) {
  rows <- hanke_raw[hanke_raw$iso3c == iso, ]
  rows <- rows[order(-rows$year), ]
  hanke_latest[[iso]] <- list(
    iso3c = iso,
    year  = as.integer(rows$year[1]),
    cpi_inflation = unname(rows$cpi_inflation_hanke[1])
  )
}

write_json(list(
  source    = "Steve Hanke Troubled Currencies Project (Cato/Hopkins)",
  countries = hanke_latest
), "../artifacts/hanke_inflation.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/hanke_inflation.json — countries:", length(hanke_latest), "\n")
```

## Build calibration.json

```{r}
#| label: calibration
build_cal_json <- function(c) {
  list(
    outcome_col       = c$outcome_col,
    holdout_year      = c$holdout_year,
    n_holdout         = c$n_holdout,
    base_rate         = c$base_rate,
    brier             = c$brier,
    brier_baseline    = c$brier_baseline,
    brier_skill_score = c$brier_skill_score,
    log_loss          = c$log_loss,
    reliability       = c$reliability
  )
}
write_json(list(
  model_version = fit$model_version,
  tiers = list(
    mobilization = build_cal_json(calibration_obj$mobilization),
    violence     = build_cal_json(calibration_obj$violence)
  )
), "../artifacts/calibration.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/calibration.json\n")
```

## Sync dashboard artifact copy

```{r}
#| label: sync-web-artifacts
dir.create("../web/artifacts", showWarnings = FALSE, recursive = TRUE)
json_files <- list.files("../artifacts", pattern = "\\.json$", full.names = TRUE)
file.copy(json_files, "../web/artifacts", overwrite = TRUE)
cat("Synced", length(json_files), "JSON artifacts to web/artifacts\n")

# Also keep mocks/artifacts in sync IF it exists. Mocks are local-only
# (untracked) prototypes; we don't auto-create the folder so other devs
# don't get a useless empty directory.
if (dir.exists("../mocks/artifacts")) {
  file.copy(json_files, "../mocks/artifacts", overwrite = TRUE)
  cat("Synced", length(json_files), "JSON artifacts to mocks/artifacts\n")
}
```

## Summary

```{r}
#| label: final
cat("=== Artifacts in ../artifacts/ ===\n")
for (f in list.files("../artifacts", pattern = "\\.json$")) {
  size <- file.info(file.path("../artifacts", f))$size
  cat(sprintf("  %-30s %8s bytes\n", f, format(size, big.mark = ",")))
}

cat("\n=== Cuba snapshot (both tiers) ===\n")
cuba <- snapshot_rows[["CUB"]]
if (is.null(cuba)) {
  cat("  (Cuba not in snapshot)\n")
} else {
  cat("  Regime:        ", cuba$regime_label, "\n")
  cat("  Snapshot year: ", cuba$snapshot_year, "\n")
  cat("  P(mobilization):", round(cuba$mobilization$probability, 3),
      "  CI [", round(cuba$mobilization$ci_lo_90, 3), ",",
                round(cuba$mobilization$ci_hi_90, 3), "]\n")
  cat("  P(violence):    ", round(cuba$violence$probability, 3),
      "  CI [", round(cuba$violence$ci_lo_90, 3), ",",
                round(cuba$violence$ci_hi_90, 3), "]\n")
}
```
