---
title: "12 — Export frozen artifacts for the dashboard"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Writes JSON files to `../artifacts/` that the static dashboard consumes.
Math runs in the browser — no backend at runtime.

- **`weights.json`** — coefficients + vcov per regime stratum + pooled; feature list
- **`base_rates.json`** — per-regime base rates from training data
- **`country_snapshot.json`** — per-country: regime, latest raw feature
  values, predicted probability + 90% CI, per-feature contribution
- **`calibration.json`** — Brier, log loss, reliability bins

v0.4: features are raw cross-country levels (no z-scoring). Sliders in the
dashboard bind directly to raw values. No `winsorize_bound`, no z↔raw
conversion, no within-country mean/sd shipped to the client.

## Libraries

```{r}
#| label: setup
library(DBI)
library(duckdb)
library(jsonlite)
```

## Load fitted model + panel

```{r}
#| label: load
fit <- readRDS("../data/model/fit.rds")
models       <- fit$models
pooled_model <- fit$pooled_model
features     <- fit$features
regime_labels <- fit$regime_labels
calibration_obj <- readRDS("../data/model/calibration.rds")

con <- dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb")
panel <- dbGetQuery(con, "SELECT * FROM panel_labeled")
dbDisconnect(con, shutdown = TRUE)
```

## Build weights.json

For each fitted stratum, store the coefficients and the variance-covariance
matrix. The dashboard samples from `MVN(coef, vcov)` to compute 90% CIs on
predictions.

```{r}
#| label: weights
strata <- list()
for (k in names(models)) {
  fit_k <- models[[k]]
  # Ridge fits store the cv.glmnet coefficients +
  # bootstrap-derived empirical covariance directly on the model list,
  # since glmnet has no closed-form vcov(). Dashboard's MVN sampling for
  # 90% CIs works the same way - sample from MVN(coef, vcov) and apply.
  co <- fit_k$coefficients
  vc <- fit_k$vcov
  strata[[regime_labels[[k]]]] <- list(
    regime_code   = as.integer(k),
    coefficients  = as.list(co),
    vcov          = unname(as.matrix(vc)),
    feature_order = fit_k$feature_order,
    n_train       = fit_k$n_train,
    n_events      = fit_k$n_events,
    lambda        = fit_k$lambda,
    n_bootstrap   = fit_k$n_bootstrap
  )
}

# IMPORTANT: pass named vectors as `as.list()` so jsonlite emits JSON objects
# with the names as keys. Plain named vectors get unboxed to arrays and lose
# their names, which breaks the dashboard's `weights.feature_labels[z_name]`
# lookups (they'd return undefined).
weights <- list(
  model_version = fit$model_version,
  fitted_at     = format(fit$fitted_at, "%Y-%m-%d"),
  train_years   = fit$train_years,
  features      = features,
  feature_labels = as.list(c(
    civil_liberties          = "Civil liberties",
    log_gdp_per_cap          = "GDP per capita",
    loser_consent            = "Loser consent",
    conflict_neighbors_count = "Regional conflict",
    gdp_growth               = "GDP growth",
    cpi_inflation            = "Inflation",
    unemployment             = "Unemployment",
    political_violence_ord   = "Political violence",
    log_refugee_outflow      = "Refugee outflow (log)",
    coup_proof_fragmentation = "Coup-proofing fragmentation"
  )),
  feature_direction = as.list(c(   # +1 = higher value raises risk, -1 = lowers
    civil_liberties          = -1,
    log_gdp_per_cap          = -1,
    loser_consent            = -1,
    conflict_neighbors_count = +1,
    gdp_growth               = -1,
    cpi_inflation            = +1,
    unemployment             = +1,
    political_violence_ord   = +1,
    log_refugee_outflow      = +1,
    coup_proof_fragmentation = -1
  )),
  strata = strata,
  pooled = list(
    coefficients  = as.list(pooled_model$coefficients),
    vcov          = unname(as.matrix(pooled_model$vcov)),
    feature_order = pooled_model$feature_order,
    lambda        = pooled_model$lambda,
    n_train       = pooled_model$n_train,
    n_events      = pooled_model$n_events,
    n_bootstrap   = pooled_model$n_bootstrap
  )
)

dir.create("../artifacts", showWarnings = FALSE, recursive = TRUE)
write_json(weights, "../artifacts/weights.json", auto_unbox = TRUE, pretty = TRUE, digits = 6)
cat("Wrote artifacts/weights.json\n")
```

## Build base_rates.json

```{r}
#| label: base-rates
features_check <- c("regime_type", "unrest_next_year", features)
train_idx <- complete.cases(panel[, features_check]) & panel$year <= 2022
train <- panel[train_idx, ]

base_rates <- list()
for (k in 0:3) {
  stratum <- train[train$regime_type == k, ]
  if (nrow(stratum) == 0) next
  base_rates[[regime_labels[[as.character(k)]]]] <- list(
    regime_code = k,
    n_obs       = nrow(stratum),
    n_unrest    = sum(stratum$unrest_next_year),
    base_rate   = round(mean(stratum$unrest_next_year), 4)
  )
}

write_json(list(
  model_version = fit$model_version,
  train_years   = fit$train_years,
  regimes       = base_rates
), "../artifacts/base_rates.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/base_rates.json\n")
```

## Build country_snapshot.json

For each country, take its **latest** year (typically 2024). Pull raw
feature values, score with the pooled cross-sectional model, and bootstrap a
90% CI by sampling 1,000 times from `MVN(coef, vcov)`.

v0.4 simplification: features are raw cross-country levels, so the snapshot
ships just `{value, contribution}` per feature — no `z`, `mean`, or `sd`
fields. Sliders in the dashboard bind directly to raw values.

```{r}
#| label: snapshot
# Sampling helper: draw from MVN given coef + (bootstrap-empirical) cov,
# score linearly + sigmoid. cv.glmnet has no closed-form vcov, so we use the
# persisted bootstrap covariance directly.
mvn_sample_ci <- function(stratum_model, x_named, n_draw = 1000, alpha = 0.10) {
  co <- stratum_model$coefficients
  vc <- stratum_model$vcov
  L  <- chol(vc + diag(1e-8, nrow(vc)))
  draws <- t(L) %*% matrix(rnorm(length(co) * n_draw), nrow = length(co)) + co
  x_full <- c(1, x_named[names(co)[-1]])
  lp <- as.numeric(t(draws) %*% x_full)
  p  <- 1 / (1 + exp(-lp))
  c(point = 1 / (1 + exp(-sum(co * x_full))),
    lo    = unname(quantile(p, alpha / 2, na.rm = TRUE)),
    hi    = unname(quantile(p, 1 - alpha / 2, na.rm = TRUE)))
}

# Build per-country snapshot for the latest year that has complete features.
latest_year <- max(panel$year[!is.na(panel$mass_mobilization_ord)])

snapshot_rows <- list()
for (iso in unique(panel$iso3c)) {
  rows <- panel[panel$iso3c == iso & panel$year <= latest_year, ]
  if (nrow(rows) == 0) next
  rows <- rows[order(-rows$year), ]
  good <- rows[complete.cases(rows[, c("regime_type", features)]), ]
  if (nrow(good) == 0) next
  cur <- good[1, ]
  k <- as.character(cur$regime_type)
  if (!k %in% names(models)) next

  x_named <- setNames(as.numeric(cur[, features]), features)

  # PRIMARY: pooled cross-sectional model.
  pool_coef <- pooled_model$coefficients
  pool_vec  <- setNames(rep(0, length(pool_coef)), names(pool_coef))
  pool_vec["(Intercept)"] <- 1
  for (f in features) if (f %in% names(pool_vec)) pool_vec[f] <- x_named[[f]]
  pool_lp  <- sum(pool_coef * pool_vec)
  pool_pred <- 1 / (1 + exp(-pool_lp))
  L_pool <- chol(pooled_model$vcov + diag(1e-8, nrow(pooled_model$vcov)))
  draws_pool <- t(L_pool) %*% matrix(rnorm(length(pool_coef) * 1000),
                                     nrow = length(pool_coef)) + pool_coef
  pool_lp_draws <- as.numeric(t(draws_pool) %*% pool_vec)
  pool_p_draws  <- 1 / (1 + exp(-pool_lp_draws))
  pool_ci_lo <- unname(quantile(pool_p_draws, 0.05))
  pool_ci_hi <- unname(quantile(pool_p_draws, 0.95))

  # SECONDARY: stratified per-regime model (analyst context only).
  pred <- mvn_sample_ci(models[[k]], x_named)

  # Per-feature contribution = coef * value (logit units, pooled coefs).
  # sapply over a named character vector loses outer names when the inner
  # value is named (pool_coef[f] is a named scalar) — set names explicitly.
  contribs <- sapply(features, function(f) {
    if (f %in% names(pool_coef)) unname(pool_coef[f] * x_named[[f]]) else 0
  })
  names(contribs) <- features

  # v0.4: ship just {value, contribution} per feature. No z, mean, sd.
  raw_vals <- list()
  for (f in features) {
    raw_vals[[f]] <- list(
      value        = unname(cur[[f]]),
      contribution = unname(contribs[[f]])
    )
  }

  # Data-quality flag: 1 when the snapshot-year CPI was region-median
  # imputed because no real source (WDI / IMF curated / IMF WEO) had a
  # value for this country-year. The dashboard surfaces this as a "low
  # data quality" badge — the prediction is still produced, but the user
  # should down-weight it for affected countries (PRK / pre-2018 CUB / etc.).
  is_imputed <- isTRUE(as.logical(cur$cpi_imputed))

  snapshot_rows[[iso]] <- list(
    iso3c        = iso,
    name         = cur$name,
    region       = cur$region,
    subregion    = cur$subregion,
    regime_type  = as.integer(cur$regime_type),
    regime_label = regime_labels[[k]],
    snapshot_year = as.integer(cur$year),
    # PRIMARY: pooled-model probability + 90% CI. Cross-country comparable.
    predicted_probability = unname(pool_pred),
    ci_lo_90              = unname(pool_ci_lo),
    ci_hi_90              = unname(pool_ci_hi),
    # SECONDARY: stratified prediction = "within-regime stress" signal.
    # Dashboard surfaces this as analyst context only.
    stratified_probability     = unname(pred["point"]),
    stratified_ci_lo_90        = unname(pred["lo"]),
    stratified_ci_hi_90        = unname(pred["hi"]),
    cpi_imputed = is_imputed,
    features  = raw_vals
  )
}

cat("Scored", length(snapshot_rows), "countries for snapshot\n")

write_json(list(
  model_version = fit$model_version,
  snapshot_year = latest_year,
  horizon_months = 12,
  countries     = snapshot_rows
), "../artifacts/country_snapshot.json", auto_unbox = TRUE, pretty = TRUE, digits = 6)
cat("Wrote artifacts/country_snapshot.json\n")
```

## Build hanke_inflation.json (dashboard toggle data)

Steve Hanke's Troubled Currencies Project estimates for hyperinflation
countries — exposed as a separate artifact so the dashboard can offer a
"see what the prediction looks like with Hanke's higher estimates" toggle
without re-fitting the model. Keys are iso3c → latest-year inflation %.

```{r}
#| label: hanke-export
hanke_raw <- dbGetQuery(
  dbConnect(duckdb::duckdb(), dbdir = "../data/panel.duckdb", read_only = TRUE),
  "SELECT iso3c, year, cpi_inflation_hanke FROM hanke_inflation"
)
# For each country: pick the latest year's Hanke estimate.
hanke_latest <- list()
for (iso in unique(hanke_raw$iso3c)) {
  rows <- hanke_raw[hanke_raw$iso3c == iso, ]
  rows <- rows[order(-rows$year), ]
  hanke_latest[[iso]] <- list(
    iso3c = iso,
    year  = as.integer(rows$year[1]),
    cpi_inflation = unname(rows$cpi_inflation_hanke[1])
  )
}

write_json(list(
  source = "Steve Hanke Troubled Currencies Project (Cato/Hopkins)",
  countries = hanke_latest
), "../artifacts/hanke_inflation.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/hanke_inflation.json — countries:", length(hanke_latest), "\n")
```

## Build calibration.json

```{r}
#| label: calibration
write_json(list(
  model_version = fit$model_version,
  holdout_year  = calibration_obj$holdout_year,
  n_holdout     = calibration_obj$n_holdout,
  brier         = calibration_obj$brier,
  brier_baseline = calibration_obj$brier_baseline,
  brier_skill_score = calibration_obj$brier_skill_score,
  log_loss      = calibration_obj$log_loss,
  base_rate     = calibration_obj$base_rate,
  reliability   = calibration_obj$reliability
), "../artifacts/calibration.json", auto_unbox = TRUE, pretty = TRUE)
cat("Wrote artifacts/calibration.json\n")
```

## Sync dashboard artifact copy

The dashboard fetches from `web/artifacts/` so it can run as a static site
with no parent-directory access. Keep that served copy in lockstep with the
canonical `artifacts/` directory every time artifacts are exported.

```{r}
#| label: sync-web-artifacts
dir.create("../web/artifacts", showWarnings = FALSE, recursive = TRUE)
json_files <- list.files("../artifacts", pattern = "\\.json$", full.names = TRUE)
file.copy(json_files, "../web/artifacts", overwrite = TRUE)
cat("Synced", length(json_files), "JSON artifacts to web/artifacts\n")
```

## Summary

```{r}
#| label: final
cat("=== Artifacts in ../artifacts/ ===\n")
for (f in list.files("../artifacts", pattern = "\\.json$")) {
  size <- file.info(file.path("../artifacts", f))$size
  cat(sprintf("  %-30s %8s bytes\n", f, format(size, big.mark = ",")))
}

# Spot-check Cuba (defensive: skip if not in snapshot for any reason)
cat("\n=== Cuba snapshot ===\n")
cuba <- snapshot_rows[["CUB"]]
if (is.null(cuba)) {
  cat("  (Cuba not in snapshot - check data coverage)\n")
} else {
  cat("  Regime:                ", cuba$regime_label, "\n")
  cat("  Snapshot year:         ", cuba$snapshot_year, "\n")
  cat("  Predicted probability: ", round(cuba$predicted_probability, 3), "\n")
  cat("  90% CI:                ", round(cuba$ci_lo_90, 3), "to", round(cuba$ci_hi_90, 3), "\n")
}
```
