---
title: "13 — Export ViEWS dashboard overlay"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

Fetches the latest ViEWS country-month conflict forecast and writes
`../artifacts/views_overlay.json` for the dashboard's "ViEWS" benchmark
row. **Not** part of the model pipeline — runs independently so the
overlay can refresh on a different cadence (ViEWS publishes ~monthly)
than the model fit.

ViEWS forecasts state-based armed conflict (≥25 fatalities/month) on the
territory of each country. We aggregate the 12 months following the
run date into a single annual probability:

    P(year) = 1 - prod(1 - main_dich[m])  over the 12-month window

Saturation at 1.0 for active-conflict countries (Ukraine, Sudan, Myanmar,
Iran, etc.) is meaningful — ViEWS is essentially certain.

## Libraries

```{r}
#| label: setup
library(httr2)
library(jsonlite)
library(readr)
library(dplyr)
library(purrr)
```

## Discover the latest published run

ViEWS publishes CSVs under
`https://viewsforecasting.org/wp-content/uploads/{upload_year}/{upload_month}/fatalities003_{data_year}_{data_month}_t01_cm.csv`.
The upload month typically lags the data month by 1 (e.g. data-month
2026-03 is uploaded under `2026/04/`). We walk backwards from the
current month until we find a 200 response.

```{r}
#| label: discover-run
ua <- "Mozilla/5.0 (compatible; scenario-risk-calculator)"

probe_run <- function(data_year, data_month) {
  upload <- as.Date(sprintf("%04d-%02d-01", data_year, data_month)) + 35
  upload_year <- as.integer(format(upload, "%Y"))
  upload_month <- as.integer(format(upload, "%m"))
  url <- sprintf(
    "https://viewsforecasting.org/wp-content/uploads/%04d/%02d/fatalities003_%04d_%02d_t01_cm.csv",
    upload_year, upload_month, data_year, data_month
  )
  resp <- tryCatch(
    request(url) |> req_user_agent(ua) |>
      req_headers(Range = "bytes=0-200") |>
      req_error(is_error = \(r) FALSE) |>
      req_perform(),
    error = function(e) NULL
  )
  if (!is.null(resp) && resp_status(resp) %in% c(200, 206)) url else NA_character_
}

today <- Sys.Date()
candidates <- expand.grid(
  data_year  = c(as.integer(format(today, "%Y")) - 1, as.integer(format(today, "%Y"))),
  data_month = 1:12
)
# walk most-recent first
candidates <- candidates[order(-candidates$data_year, -candidates$data_month), ]
candidates <- candidates[candidates$data_year * 12 + candidates$data_month <=
                         as.integer(format(today, "%Y")) * 12 + as.integer(format(today, "%m")), ]

run_url <- NA_character_
for (i in seq_len(nrow(candidates))) {
  u <- probe_run(candidates$data_year[i], candidates$data_month[i])
  if (!is.na(u)) { run_url <- u; data_year <- candidates$data_year[i]; data_month <- candidates$data_month[i]; break }
}
stopifnot(!is.na(run_url))
cat("Using run:", run_url, "\n")
```

## Download and parse

```{r}
#| label: download
tmp <- tempfile(fileext = ".csv")
request(run_url) |> req_user_agent(ua) |> req_perform(path = tmp)
views <- read_csv(tmp, show_col_types = FALSE)
unlink(tmp)
cat("Rows:", nrow(views), "  Countries:", length(unique(views$isoab)), "\n")
```

## Aggregate to 12-month annual probability

Window = the 12 months following the data month of the run.

```{r}
#| label: aggregate
window_start <- as.Date(sprintf("%04d-%02d-01", data_year, data_month)) + 31
window_start <- as.Date(sprintf("%04d-%02d-01",
                                as.integer(format(window_start, "%Y")),
                                as.integer(format(window_start, "%m"))))
window_months <- format(seq(window_start, by = "month", length.out = 12), "%Y-%m")

views <- views |>
  mutate(ym = sprintf("%04d-%02d", year, month),
         main_dich = pmin(pmax(main_dich, 0), 1))  # clamp [0,1]

agg <- views |>
  filter(ym %in% window_months) |>
  group_by(isoab, country) |>
  summarise(
    probability       = 1 - prod(1 - main_dich),
    monthly_max       = max(main_dich),
    n_months_covered  = n(),
    .groups = "drop"
  ) |>
  mutate(probability = round(probability, 4),
         monthly_max = round(monthly_max, 4))
```

## Write artifact

```{r}
#| label: write
countries_obj <- setNames(
  lapply(seq_len(nrow(agg)), function(i) list(
    country          = agg$country[i],
    probability      = agg$probability[i],
    monthly_max      = agg$monthly_max[i],
    n_months_covered = agg$n_months_covered[i]
  )),
  agg$isoab
)

out <- list(
  as_of         = format(Sys.Date()),
  source        = "ViEWS (Uppsala / PRIO)",
  source_url    = "https://viewsforecasting.org",
  run_id        = sprintf("fatalities003_%04d_%02d_t01", data_year, data_month),
  window_start  = window_months[1],
  window_end    = window_months[12],
  outcome_note  = "Probability of state-based armed conflict (>=25 fatalities/month) in at least one of the next 12 months on the territory of the country. Covers only the armed-conflict component of the model outcome — not the coup or protest components.",
  aggregation   = "one_minus_product_of_complements",
  license       = "See https://viewsforecasting.org for current license terms (confirm before re-publishing).",
  countries     = countries_obj
)

out_path <- "../artifacts/views_overlay.json"
write_json(out, out_path, auto_unbox = TRUE, pretty = TRUE)
cat("Wrote", out_path, "  (", nrow(agg), "countries )\n")
```

Also mirror into `../web/artifacts/` so GitHub Pages serves it:

```{r}
#| label: mirror
file.copy(out_path, "../web/artifacts/views_overlay.json", overwrite = TRUE)
```

## Sanity check

```{r}
#| label: sanity
top   <- agg |> arrange(desc(probability)) |> head(15)
bottom <- agg |> arrange(probability) |> head(5)
print(top, n = 15)
print(bottom, n = 5)
```
